---
title: "Drought Manuscript Analysis - SPRUCE 2020–2022 surface peat"
author: "Katherine Duchesneau"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  pdf_document: default
---

## 1. HOUSEKEEPING

```{r setup, warning=FALSE, message=FALSE}

set.seed(1234)

PROJ <- "/Users/katherineduchesneau/Library/CloudStorage/Dropbox-GaTech/Katherine Duchesneau/Drought paper/Drought_R_analyses"
knitr::opts_knit$set(root.dir = PROJ)

out_dir <- "2Revisions_out"

# ------------------------------------------------------------------------------
# 1.1  Run switches
# ------------------------------------------------------------------------------

# Refit every Maaslin2 model from the phyloseq objects. FALSE reads 03_maaslin/.
REFIT_MAASLIN <- FALSE

# The two large gene fits (RE_networked_ALL, RE_ALL_CAMPERgenes). Hours to
# overnight. Only consulted when REFIT_MAASLIN is TRUE.
REFIT_LARGE <- FALSE

# Skip a refit when the stored fit came from the same inputs and the same model.
MAASLIN_CACHE <- TRUE

# TRUE runs SpiecEasi and writes se_FULL.RDS / assoMat_FULL.Rds into 08_network/.
# FALSE reloads those two files and rebuilds the graph, tables and figure from them.
# Set TRUE whenever the metabolite filter changes, or Figure 3 can carry an edge
# the current filtering does not produce.
RUN_NETWORK <- FALSE

# Number of resampling replicates behind Check_Fig3_edge_stability.csv.
RESAMPLE_N <- 50

# Edge recovery is read from 00_inputs/Check_Fig3_edge_stability.csv when that
# file matches the edge set actually plotted. When it does not, TRUE recomputes
# the frequencies in place, at a cost of RESAMPLE_N SpiecEasi fits, and FALSE
# stops the knit.
RUN_RESAMPLE <- FALSE

# Delete 01_metadata through 08_network before anything runs. 00_inputs is never
# touched. 08_network is only cleared when RUN_NETWORK is TRUE.
WIPE_OUTPUTS <- FALSE

# ------------------------------------------------------------------------------
# 1.2  Folders and path builders
# ------------------------------------------------------------------------------

DIRS <- c(raw     = "00_inputs",
          meta    = "01_metadata",
          ps      = "02_phyloseq",
          maaslin = "03_maaslin",
          sd      = "04_suppl_data",
          fig     = "06_figures",
          net     = "08_network")

if (isTRUE(WIPE_OUTPUTS)) {
  to_wipe <- setdiff(names(DIRS), "raw")
  if (!isTRUE(RUN_NETWORK)) to_wipe <- setdiff(to_wipe, "net")
  for (k in to_wipe) unlink(file.path(out_dir, DIRS[k]), recursive = TRUE)
  message("WIPE_OUTPUTS: cleared ", paste(DIRS[to_wipe], collapse = ", "))
}
for (d in DIRS) dir.create(file.path(out_dir, d), recursive = TRUE, showWarnings = FALSE)

p_raw  <- function(...) file.path(out_dir, DIRS["raw"],     ...)
p_meta <- function(...) file.path(out_dir, DIRS["meta"],    ...)
p_ps   <- function(...) file.path(out_dir, DIRS["ps"],      ...)
p_ml   <- function(...) file.path(out_dir, DIRS["maaslin"], ...)
p_sd   <- function(...) file.path(out_dir, DIRS["sd"],      ...)
p_fig  <- function(...) file.path(out_dir, DIRS["fig"],     ...)
p_net  <- function(...) file.path(out_dir, DIRS["net"],     ...)

# Every input is read from 00_inputs. Nothing is read from anywhere else.
p_raw_req <- function(name) {
  p <- p_raw(name)
  if (!file.exists(p)) stop(name, " is not in ", p_raw(""))
  p
}

maaslin_path <- function(name) {
  p <- p_ml(name)
  if (!file.exists(p))
    stop(name, " is not in ", p_ml(""), ". Set REFIT_MAASLIN <- TRUE.")
  p
}

net_path <- function(name) {
  p <- p_net(name)
  if (!file.exists(p))
    stop(name, " is not in ", p_net(""), ". Set RUN_NETWORK <- TRUE.")
  p
}

# ------------------------------------------------------------------------------
# 1.3  Packages
# ------------------------------------------------------------------------------

library(data.table)
library(readr)
library(tidyverse)
library(stringr)
library(reshape2)

library(phyloseq)
library(vegan)
library(MASS)
library(glmmTMB)
library(MuMIn)
library(Maaslin2)
library(Hmisc)
library(car)
library(interactions)
library(lmerTest)
library(glmtoolbox)

library(clusterProfiler)
library(KEGGREST)

library(ggplot2)
library(ggpubr)
library(ggrepel)
library(scico)
library(wesanderson)
library(ggsci)
library(RColorBrewer)

library(ggtree)
library(ggtreeExtra)
library(ggstar)
library(ggnewscale)
library(pals)

library(igraph)
library(ggraph)
library(SpiecEasi)

library(conflicted)
conflict_prefer("summarise", "dplyr"); conflict_prefer("summarize", "dplyr")
conflict_prefer("step",      "stats"); conflict_prefer("filter",    "dplyr")
conflict_prefer("select",    "dplyr"); conflict_prefer("rename",    "dplyr")
conflict_prefer("mutate",    "dplyr"); conflict_prefer("arrange",   "dplyr")
conflict_prefer("count",     "dplyr"); conflict_prefer("slice",     "dplyr")

# ------------------------------------------------------------------------------
# 1.4  Colour palettes
# ------------------------------------------------------------------------------

allGroupsColors <- c("#FFDA77","#FFB24C","#FE8C3B","#F03B1F","#BD0026")

cluster_colors <- c("Oxygen sensitive"           = "#D740F2",
                    "Warming-inhibited"          = "#0066CC",
                    "Warming-stimulated oxic"    = "#538F6B",
                    "Warming-stimulated anoxic"  = "#CA9F01",
                    "None"                       = "white")

specialist_colors <- c("Oxic_specialist"   = "#A3C9F1",
                       "Anoxic_specialist" = "#E24D7F",
                       "Ambivalent"        = "#9900CC",
                       "none"              = "white")

# ------------------------------------------------------------------------------
# 1.5  Constants
# ------------------------------------------------------------------------------

# Methanogens, a subset of the oxygen-sensitive cluster.
METHANOGEN_IDS <- c("SPRUCE 510", "SPRUCE 2026", "SPRUCE 2021", "SPRUCE 556")

# Warming-stimulated, oxic-preferring (WS-OX), n = 7
WSOX_IDS <- c(
  "SPRUCE 87",    # Proteobacteria REEB95
  "SPRUCE 397",   # Proteobacteria Roseiarcus
  "SPRUCE 675",   # Actinobacteriota Palsa-601
  "SPRUCE 2042",  # Actinobacteriota PALSA-457
  "SPRUCE 2183",  # Proteobacteria BOG-931
  "SPRUCE 2185",  # Verrucomicrobiota
  "SPRUCE 2313")  # Verrucomicrobiota Palsa-1447

# Warming-stimulated, anoxic-preferring (WS-AN), n = 15
WSAN_IDS <- c(
  "SPRUCE 124",   # Acidobacteriota Sulfotelmatobacter
  "SPRUCE 275",   # Proteobacteria Bog-1198
  "SPRUCE 356",   # Actinobacteriota RAAP-2
  "SPRUCE 607",   # Proteobacteria Bog-1198
  "SPRUCE 2046",  # Proteobacteria 13-2-20CM-66-19
  "SPRUCE 2049",  # Actinobacteriota Palsa-744
  "SPRUCE 2053",  # Proteobacteria Bog-1198
  "SPRUCE 2057",  # Desulfobacterota_B
  "SPRUCE 2060",  # Acidobacteriota Palsa-343
  "SPRUCE 2148",  # Actinobacteriota RAAP-2
  "SPRUCE 2152",  # Acidobacteriota Palsa-343
  "SPRUCE 2155",  # Actinobacteriota Palsa-744
  "SPRUCE 2159",  # Actinobacteriota RAAP-2
  "SPRUCE 2184",  # Acidobacteriota Palsa-288
  "SPRUCE 2191")  # Proteobacteria BOG-931

# Warming-inhibited (WI), n = 9
WI_IDS <- c(
  "SPRUCE 190",   # Myxococcota Fen-1118
  "SPRUCE 600",   # Proteobacteria BOG-994
  "SPRUCE 689",   # Acidobacteriota Acidoferrum
  "SPRUCE 2226",  # Proteobacteria
  "SPRUCE 2282",  # Verrucomicrobiota Palsa-1447
  "SPRUCE 2287",  # Planctomycetota BOG-1363
  "SPRUCE 2292",  # Acidobacteriota Bog-113
  "SPRUCE 2293",  # Eremiobacterota
  "SPRUCE 2294")  # Proteobacteria FEN-1191

WARMING_RESPONSIVE_IDS <- c(WSOX_IDS, WSAN_IDS, WI_IDS)

POLY_SUPERCLASSES <- c("Benzenoids",
                       "Phenylpropanoids and polyketides",
                       "Lignans, neolignans and related compounds")
NETWORK_PHENOLIC_SUPERCLASSES <- c(POLY_SUPERCLASSES, "Organosulfur compounds")

# Maaslin2 feature filtering, applied inside every model.
MIN_PREVALENCE <- 0.10
MIN_ABUNDANCE  <- 0

# PhyloseqObj_metabolites.RDS holds already-normalised, already-log2 intensities,
# identical to norm_auc_mean.csv. The relationship to the raw peak areas is
#
#   norm_ij = log2( 2 * M * x_ij / m_j )
#
# with x_ij the raw area under the curve, m_j the mean raw AUC of sample j across
# all features, and M the grand mean of the raw matrix (M = mean of the m_j).
# This is a sample-wise mean normalisation rescaled to the grand mean, then log2.
# Section 2.9.1b checks the stored intensities against this formula.
RAW_AUC_FILE <- "raw_auc_table.csv"   # optional; place in 00_inputs/

# Compound Discoverer per-feature values: Area CV% (CV of peak area per
# compound across the sample set) and the assigned MSI level. Keyed on `feature`.
METAB_ANNOT_UPDATE_FILE <- "Metabolite_annotation_update.csv"

# Input file names.
MAG_ABUNDANCE_RDS <- "PhyloseqObj_droughtMAG_24_03_2025.RDS"
MAG_ACTIVITY_RDS  <- "PhyloseqObj_droughtMAGactivity_24_03_2025.RDS"
GENES_PS_FILE     <- "PhyloseqObj_GeneswithMAGs_normalizedRibosomal_NORMALIZEDmedian.RDS"
HYDDB_PS_FILE     <- "PhyloseqObj_GeneswithMAGs_normalizedRibosomal_hydDB_noTREE.RDS"
METAB_PS_FILE     <- "PhyloseqObj_metabolites.RDS"
CLUSTER_CSV       <- "Clustered_MAG_edited.csv"
MODULE_CSV        <- "Suppl_Data_15_methanogen_pathways.csv"

# Annual modelled CH4 emission per enclosure, from Cory et al. (2025),
# doi:10.1029/2024JG008371. Not produced by this pipeline; supplied as a keyed
# CSV. Columns: Year, Plot, "CH4 flux" in umol m-2 y-1. The second row carries
# the units and is dropped on read.
CH4_FLUX_CSV <- "annualFluxesSPRUCE.csv"

# The ten enclosures sampled in this study.
STUDY_PLOTS <- c(4, 6, 8, 10, 11, 13, 16, 17, 19, 20)

# Network settings.
DEGREE_THRESHOLD <- 6
ABUND_QUANTILE   <- 0.30

# ------------------------------------------------------------------------------
# 1.6  Helpers
# ------------------------------------------------------------------------------

# The two large phyloseq objects are read once each and cached.
load_genes_ps <- local({
  cached <- NULL
  function() {
    if (is.null(cached)) cached <<- readRDS(p_raw_req(GENES_PS_FILE))
    cached
  }
})
load_hyddb_ps <- local({
  cached <- NULL
  function() {
    if (is.null(cached)) cached <<- readRDS(p_raw_req(HYDDB_PS_FILE))
    cached
  }
})

# Orientation-agnostic sample x feature matrix.
sample_matrix <- function(ps) {
  m <- as(otu_table(ps), "matrix")
  if (taxa_are_rows(ps)) m <- t(m)
  as.data.frame(m)
}

add_effect_size <- function(df) {
  dplyr::mutate(df,
    CI_lower = coef - 1.96 * stderr,
    CI_upper = coef + 1.96 * stderr,
    evidence = dplyr::case_when(
      qval < 0.05 ~ "Confirmatory (FDR < 0.05)",
      qval < 0.10 ~ "Suggestive (FDR < 0.10)",
      pval < 0.05 ~ "Exploratory (nominal P < 0.05 only)",
      TRUE        ~ "Not selected"))
}

fdr_summary <- function(res, model_label) {
  res$results |>
    dplyr::group_by(predictor = metadata) |>
    dplyr::summarise(
      n_tested  = dplyr::n(),
      p_lt_0.05 = sum(pval < 0.05, na.rm = TRUE),
      q_lt_0.05 = sum(qval < 0.05, na.rm = TRUE),
      q_lt_0.10 = sum(qval < 0.10, na.rm = TRUE),
      q_lt_0.25 = sum(qval < 0.25, na.rm = TRUE),
      min_qval  = min(qval, na.rm = TRUE),
      .groups   = "drop") |>
    dplyr::mutate(model = model_label, .before = 1)
}

read_maaslin <- function(path, id_col = "feature", fix_spruce = FALSE) {
  r <- readRDS(path)$results |> dplyr::rename(!!id_col := 1)
  if (fix_spruce)
    r[[id_col]] <- gsub("(SPRUCE)\\.(\\d+)", "\\1 \\2", r[[id_col]])
  add_effect_size(r)
}

tidy_glmmtmb <- function(model, response) {
  cf <- as.data.frame(summary(model)$coefficients$cond)
  colnames(cf) <- c("Estimate", "Std_Error", "z_value", "p_value")
  cf$term <- rownames(cf); rownames(cf) <- NULL
  r2 <- suppressWarnings(try(r.squaredGLMM(model), silent = TRUE))
  ok <- !inherits(r2, "try-error")
  singular <- any(is.nan(cf$Std_Error))
  data.frame(
    response       = response,
    formula        = paste(deparse(formula(model)), collapse = " "),
    n              = nobs(model),
    cf,
    sig            = dplyr::case_when(cf$p_value < 0.001 ~ "***",
                                      cf$p_value < 0.01  ~ "**",
                                      cf$p_value < 0.05  ~ "*",
                                      cf$p_value < 0.1   ~ ".", TRUE ~ ""),
    R2_marginal    = if (ok) round(r2[1, "R2m"], 3) else NA_real_,
    R2_conditional = if (ok) round(r2[1, "R2c"], 3) else NA_real_,
    singular_fit   = singular,
    row.names      = NULL)
}

# Refit without (1|Plot) when its variance is ~0 and the standard errors are NaN.
drop_singular_re <- function(model, data, label) {
  sd_plot <- tryCatch(attr(VarCorr(model)$cond$Plot, "stddev")[1],
                      error = function(e) NA_real_)
  nan_se  <- any(is.nan(as.data.frame(summary(model)$coefficients$cond)$`Std. Error`))
  if (isTRUE(nan_se) || (!is.na(sd_plot) && sd_plot < 0.01)) {
    f <- as.formula(gsub("\\s*\\+\\s*\\(1\\s*\\|\\s*Plot\\)", "",
                         paste(deparse(formula(model)), collapse = " ")))
    message(label, ": (1|Plot) variance is ~0 (sd = ", signif(sd_plot, 3),
            "). Refitting without the random effect.")
    return(list(model = glmmTMB(f, data = data, family = family(model)),
                dropped_re = TRUE))
  }
  list(model = model, dropped_re = FALSE)
}

# Keep the interaction only if the LRT supports it.
lrt_interaction <- function(f_full, f_noint, data, family = gaussian(), label = "") {
  m_full  <- glmmTMB(f_full,  data = data, family = family)
  m_noint <- glmmTMB(f_noint, data = data, family = family)
  p <- anova(m_full, m_noint)$`Pr(>Chisq)`[2]
  keep <- !is.na(p) && p < 0.05
  message(label, ": interaction LRT P = ", signif(p, 3),
          if (keep) " -> retained" else " -> dropped")
  list(model = if (keep) m_full else m_noint, lrt_p = p, retained = keep)
}

# Shared Maaslin2 configuration: prevalence and abundance filtering inside the
# model, no internal normalisation.
# `transform` is passed through to Maaslin2. Its default there is "LOG", which is
# right for count-like inputs but wrong for the metabolites: those are supplied
# already log2 transformed, so leaving the default applies a second log. The
# metabolite call therefore passes transform = "NONE". Every other model keeps
# "LOG" unless and until its input is shown to be pre-transformed; the value
# range printed below is the diagnostic for that.
fit_maaslin <- function(ps, name, fixed, random = "Plot", features = NULL,
                        min_prev = MIN_PREVALENCE, min_abun = MIN_ABUNDANCE,
                        transform = "LOG") {
  dat <- sample_matrix(ps)
  if (!is.null(features)) dat <- dat[, intersect(features, colnames(dat)), drop = FALSE]
  meta <- data.frame(sample_data(ps))
  stopifnot(setequal(rownames(dat), rownames(meta)))
  miss <- setdiff(c(fixed, random), names(meta))
  if (length(miss))
    stop(name, ": predictor(s) not in sample_data: ", paste(miss, collapse = ", "),
         "\n  available: ", paste(sort(names(meta)), collapse = ", "))
  allna <- c(fixed, random)[vapply(c(fixed, random),
                                   \(v) all(is.na(meta[[v]])), logical(1))]
  if (length(allna))
    stop(name, ": predictor(s) present but entirely NA: ",
         paste(allna, collapse = ", "))

  kept <- sum(colMeans(dat > min_abun, na.rm = TRUE) >= min_prev)
  message(sprintf("%s: %d samples, %d features, %d pass prevalence >= %.2f (%d dropped)",
                  name, nrow(dat), ncol(dat), kept, min_prev, ncol(dat) - kept))

  # An input with no zeros and a narrow total range has already been log
  # transformed upstream, in which case transform = "LOG" applies a second one.
  rng <- range(dat, na.rm = TRUE); nz <- sum(dat == 0, na.rm = TRUE)
  message(sprintf("  %s: range %.4g to %.4g, %d zeros, transform = %s",
                  name, rng[1], rng[2], nz, transform))
  if (nz == 0 && rng[1] > 0 && rng[2] / rng[1] < 100 && transform == "LOG")
    warning(name, ": input has no zeros and only a ", signif(rng[2] / rng[1], 3),
            "-fold total range, so it looks already log transformed, but ",
            "transform = \"LOG\" will log it again. Pass transform = \"NONE\" ",
            "if that is the case.")

  chr_sig <- function(x) sum(utf8ToInt(paste0(x, collapse = "")))
  sig <- list(
    fixed = fixed, random = random, min_prev = min_prev, min_abun = min_abun,
    transform = transform,
    dims = dim(dat),
    names = c(chr_sig(rownames(dat)), chr_sig(colnames(dat))),
    values = c(sum(dat, na.rm = TRUE), sum(dat^2, na.rm = TRUE), sum(is.na(dat))),
    predictors = vapply(c(fixed, random), function(v) {
      x <- meta[rownames(dat), v]
      if (is.numeric(x)) sum(x, na.rm = TRUE) else chr_sig(as.character(x))
    }, numeric(1)))

  rds  <- p_ml(paste0(name, ".rds"))
  sigf <- p_ml(paste0(name, ".signature.rds"))
  if (isTRUE(MAASLIN_CACHE) && file.exists(rds) && file.exists(sigf) &&
      isTRUE(all.equal(readRDS(sigf), sig))) {
    message("  ", name, ": cached fit matches these inputs, skipping refit.")
    return(invisible(readRDS(rds)))
  }

  m <- Maaslin2(
    input_data = dat, input_metadata = meta[rownames(dat), , drop = FALSE],
    min_abundance = min_abun, min_prevalence = min_prev, normalization = "NONE",
    transform = transform,
    output = p_ml(name), fixed_effects = fixed, random_effects = random)
  saveRDS(m, rds)
  saveRDS(sig, sigf)
  invisible(m)
}

MODEL_TABLE <- list()
FILE_LOG    <- character(0)
register_model <- function(x) { MODEL_TABLE[[length(MODEL_TABLE) + 1]] <<- x; invisible(x) }
write_out <- function(x, path, ...) {
  readr::write_csv(x, path, ...)
  FILE_LOG <<- c(FILE_LOG, path)
  message("wrote ", path, "  (", nrow(x), " rows)")
  invisible(x)
}
save_fig <- function(plot, path, ...) {
  ggsave(path, plot, ...)
  FILE_LOG <<- c(FILE_LOG, path)
  invisible(path)
}

# ------------------------------------------------------------------------------
# Metabolite feature-ID parsing and annotation-name bookkeeping.
#
# Feature IDs are "<acquisition mode>_<retention time, min>_<precursor m/z>",
# e.g. "RP_6.745_275.15998". Splitting them into their own columns lets a reader
# judge whether two features carrying the same compound name are isomers,
# adducts or redundant assignments.
# ------------------------------------------------------------------------------

parse_feature_id <- function(feature) {
  m   <- stringr::str_match(feature, "^([A-Za-z]+)_([0-9.]+)_([0-9.]+)$")
  bad <- is.na(m[, 1]) & !is.na(feature)
  if (any(bad))
    warning(sum(bad), " feature ID(s) do not match '<mode>_<RT>_<mz>' and were ",
            "left as NA, e.g. ", paste(utils::head(feature[bad], 3), collapse = ", "))
  data.frame(Acquisition_mode = m[, 2],
             RT_min           = suppressWarnings(as.numeric(m[, 3])),
             precursor_mz     = suppressWarnings(as.numeric(m[, 4])),
             stringsAsFactors = FALSE)
}

# ------------------------------------------------------------------------------
# MSI annotation levels
# ------------------------------------------------------------------------------
# Annotation sub-levels are mapped onto the standard four-level MSI scheme
# (Rajakaruna et al. 2026, doi:10.1093/ismeco/ycag223): L1 authentic standard;
# L2 mzCloud MS/MS match; L3 computational structure proposal; L4 formula only.
MSI4_MAP <- c(L1 = "L1", L2 = "L2",
              L2.1 = "L3", L2.2 = "L3", L2.3 = "L3",
              L3.1 = "L4", L3.2 = "L4", L3.3 = "L4",
              L3 = "L3", L4 = "L4")   # idempotent: recoding twice is a no-op

to_msi4 <- function(conf) {
  conf <- trimws(as.character(conf))
  conf[conf %in% c("", "NA")] <- NA_character_
  out  <- unname(MSI4_MAP[conf])
  unmapped <- !is.na(conf) & is.na(out)
  if (any(unmapped))
    stop("Annotation_Confidence values with no entry in MSI4_MAP: ",
         paste(sort(unique(conf[unmapped])), collapse = ", "),
         ". Extend MSI4_MAP rather than letting them through as NA.")
  out
}

# A name is an identification claim; L3 and L4 names are the best-matching
# candidate structure. Anything at or below L3, and anything with no level at
# all, is labelled in Figure 3 by compound class and feature ID instead.
MSI_NAMEABLE <- c("L1", "L2")

# Two names for what is one putative assignment. In-silico fragmentation and
# predicted retention time cannot resolve stereochemistry, so the diastereomer
# pair below is reported once, without a stereodescriptor. Matched on Final_name.
NAME_ALIASES <- c(
  "(5R,6S)-5-Hydroxy-4-methoxy-6-(2-phenylethyl)-5,6-dihydro-2H-pyran-2-one" =
    "5-Hydroxy-4-methoxy-6-(2-phenylethyl)-5,6-dihydro-2H-pyran-2-one (stereochemistry unresolved)",
  "(5S,6S)-5-hydroxy-4-methoxy-6-(2-phenylethyl)-5,6-dihydro-2H-pyran-2-one" =
    "5-Hydroxy-4-methoxy-6-(2-phenylethyl)-5,6-dihydro-2H-pyran-2-one (stereochemistry unresolved)")

apply_name_aliases <- function(x) {
  hit <- !is.na(x) & x %in% names(NAME_ALIASES)
  if (any(hit))
    message("name aliases: collapsed ", sum(hit), " feature name(s) onto ",
            length(unique(unname(NAME_ALIASES[x[hit]]))), " assignment(s)")
  x[hit] <- unname(NAME_ALIASES[x[hit]])
  x
}

# Features are never merged. A compound name carried by more than one feature is
# labelled as such and the count travels with it into the supplementary tables,
# so a shared name can be read as what it is rather than as replication.
add_name_multiplicity <- function(df, name_col = "Final_name") {
  nm <- df[[name_col]]
  nm[nm %in% c("", "M", "NA")] <- NA_character_
  n  <- ave(rep(1L, length(nm)), ifelse(is.na(nm), "unnamed", nm), FUN = length)
  df$Name_n_features <- ifelse(is.na(nm), NA_integer_, as.integer(n))
  df$Name_status <- dplyr::case_when(
    is.na(nm)              ~ "unnamed feature",
    df$Name_n_features > 1 ~ "name shared with other features",
    TRUE                   ~ "name unique to this feature")
  df
}

# ------------------------------------------------------------------------------
# The one place the metabolite phyloseq is read
# ------------------------------------------------------------------------------
# The NOSC block, the Maaslin2 block and the network all read the metabolite
# phyloseq through this function, so the same filtered feature set is used
# throughout.
#
# Two transformations happen here and nowhere else:
#
#   1. "M" is the unnamed sentinel in the annotation table, not a
#      compound. Setting it to NA is what filters it: add_name_multiplicity(),
#      the class-level analysis, the NOSC subset and the network all key off a
#      non-missing Final_name.
#   2. Annotation_Confidence is mapped onto the standard four MSI levels.
#      The original value is kept in
#      Annotation_Confidence_sublevel so nothing is lost.
#
# METAB_M_FILTERED and METAB_MSI_CENSUS are written on the first call and read
# back by the reporting code, so the counts quoted in the Methods come from the
# same object the analysis used.
METAB_M_FILTERED <- NA_integer_
METAB_MSI_CENSUS <- NULL

read_metab_ps <- function(quiet = FALSE) {
  ps  <- readRDS(p_raw_req(METAB_PS_FILE))
  tx  <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE)

  n_M <- sum(tx$Final_name %in% "M", na.rm = TRUE)
  tx$Final_name[tx$Final_name %in% c("M", "")] <- NA_character_
  tx$Final_name <- apply_name_aliases(tx$Final_name)

  tx$Annotation_Confidence_sublevel <- tx$Annotation_Confidence
  tx$Annotation_Confidence <- to_msi4(tx$Annotation_Confidence_sublevel)

  census <- as.data.frame(table(
    MSI_level = factor(tx$Annotation_Confidence,
                       levels = c("L1", "L2", "L3", "L4"), exclude = NULL)),
    responseName = "n_features")

  METAB_M_FILTERED <<- n_M
  METAB_MSI_CENSUS <<- census

  if (!quiet) {
    cat(sprintf("Unnamed sentinel: %d feature(s) had Final_name == \"M\", filtered to NA\n",
                n_M))
    cat("\n=== Annotation confidence census, four-level scheme ===\n")
    print(as.data.frame(table(
      old = factor(tx$Annotation_Confidence_sublevel, exclude = NULL),
      new = factor(tx$Annotation_Confidence,          exclude = NULL))) |>
      subset(Freq > 0))
    print(census)
  }

  tax_table(ps) <- as.matrix(tx)
  ps
}

# Legends are written to <name>_legend.pdf rather than placed on the panels, so
# panel aspect ratios stay identical across a multi-panel figure.
save_legend <- function(plot, path, width = 3, height = 4) {
  lg <- try(ggpubr::get_legend(plot + theme(legend.position = "right")),
            silent = TRUE)
  if (inherits(lg, "try-error") || is.null(lg)) {
    message("no legend to extract for ", basename(path)); return(invisible(NULL))
  }
  ggsave(path, ggpubr::as_ggplot(lg), width = width, height = height)
  FILE_LOG <<- c(FILE_LOG, path)
  invisible(path)
}

legend_path <- function(path) sub("\\.pdf$", "_legend.pdf", path)

save_fig_nolegend <- function(plot, path, ..., legend_width = 3, legend_height = 4) {
  save_legend(plot, legend_path(path), width = legend_width, height = legend_height)
  save_fig(plot + theme(legend.position = "none"), path, ...)
}

# ------------------------------------------------------------------------------
# 1.7  Hand-curated supplementary sheets
# ------------------------------------------------------------------------------
# Supplementary Data 15 and 18 have no code path. They are copied from 00_inputs
# so that 04_suppl_data can be regenerated in full.

for (f in c("Suppl_Data_15_methanogen_pathways.csv",
            "Suppl_Data_18_annotation_citations.csv")) {
  file.copy(p_raw_req(f), p_sd(f), overwrite = TRUE)
  FILE_LOG <- c(FILE_LOG, p_sd(f))
}

message("REFIT_MAASLIN = ", REFIT_MAASLIN, " | REFIT_LARGE = ", REFIT_LARGE,
        " | RUN_NETWORK = ", RUN_NETWORK, " | WIPE_OUTPUTS = ", WIPE_OUTPUTS)
```

## 2. METADATA AND ENVIRONMENTAL MODELS

Primary predictors throughout:

- `Temp_avg_top`   = (AvgTemp10cm + AvgTemp20cm) / 2, May–Aug (°C)
- `WT_drawdown_cm` = positive values are deeper and drier (cm)

```{r load-metadata}

metadata_df <- left_join(
    read.csv(p_raw_req("metadata.csv")),
    read.csv(p_raw_req("Spruce_summer2020-2022_temp_WT_Moisture_MayJuneJulyAugust.csv")) |>
      dplyr::select(-any_of("X")),
    by = c("Plot", "Year")) |>
  mutate(
    Year              = as.factor(Year),
    Plot              = as.factor(Plot),
    Warming           = as.factor(Warming),
    Warming_Treatment = as.factor(Warming_Treatment),
    CO2_Treatment     = as.factor(CO2_Treatment),
    Depth             = as.factor(Depth),
    CO2               = as.factor(CO2),
    Warming_Treatment.ord = forcats::fct_relevel(
      Warming, "0 ºC", "2.25 ºC", "4.5 ºC", "6.75 ºC", "9 ºC"),
    WT_drawdown       = 0 - AvgWTE_normalized,
    WT_drawdown_cm    = WT_drawdown * 100,
    Temp_avg_top      = (AvgTemp10cm + AvgTemp20cm) / 2,
    SoilWater_avg_top = (AvgMoisture0cm + AvgMoisture20cm) / 2)

write_out(metadata_df, p_meta("metadata_df.csv"))
```

## 2.2 WATER TABLE AND MOISTURE MODELS

One observation per enclosure per year, so no random effect. Stepwise AIC.

```{r env-models}

cat("=== Normality checks ===\n")
cat("WT_drawdown_cm: p =", round(shapiro.test(metadata_df$WT_drawdown_cm)$p.value, 3), "\n")
cat("AvgMoisture0cm: p =", round(shapiro.test(metadata_df$AvgMoisture0cm)$p.value, 3), "\n")

step.WTE <- step(glm((WT_drawdown_cm + 100) ~ Temp_avg_top + Temp_avg_top:Year,
                     data = metadata_df, family = Gamma(link = "identity")), trace = 0)
cat("\n===== Water table drawdown (Gamma GLM) =====\n")
cat("Adj R2 =", round(glmtoolbox::adjR2(step.WTE), 3), "\n"); print(summary(step.WTE))

step.moist0 <- step(lm(AvgMoisture0cm ~ Temp_avg_top + Temp_avg_top:Year,
                       data = metadata_df), trace = 0)
cat("\n===== Surface moisture (0 cm) =====\n")
cat("Adj R2 =", round(summary(step.moist0)$adj.r.squared, 3), "\n"); print(summary(step.moist0))

step.moist20 <- step(lm(AvgMoisture20cm ~ Temp_avg_top + Temp_avg_top:Year,
                        data = metadata_df), trace = 0)
cat("\n===== Deep moisture (20 cm) =====\n")
cat("Adj R2 =", round(summary(step.moist20)$adj.r.squared, 3), "\n"); print(summary(step.moist20))
```

## 2.3 SUPPLEMENTARY FIGURE S1

```{r suppl-fig-watertable, fig.width=12, fig.height=6}

env_plot <- metadata_df[!is.na(metadata_df$Warming_Treatment.ord), ]

boxplot_by_year <- function(y, ylab) {
  ggplot(env_plot, aes(x = Year, y = .data[[y]], fill = Warming_Treatment.ord)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.5, size = 2, pch = 21, color = "black") +
    scale_fill_manual(values = allGroupsColors) +
    labs(x = "Year", y = ylab, fill = "Warming\ntreatment") +
    theme_bw() +
    theme(axis.text = element_text(size = 14), axis.title = element_text(size = 16),
          legend.text = element_text(size = 12), legend.title = element_text(size = 13))
}

FIG_S1 <- ggarrange(
  boxplot_by_year("AvgTemp10cm",    "Mean temperature at 10 cm\n(May-Aug, ºC)"),
  boxplot_by_year("WT_drawdown_cm", "Water table drawdown (cm)\n(May-Aug average)"),
  ncol = 2, common.legend = TRUE, legend = "right",
  labels = c("C", "D"), font.label = list(size = 18))
save_fig(FIG_S1, p_fig("Figure_S1_environmental_drivers.pdf"), width = 12, height = 5)
```

## 2.4 MAG TAXONOMY AND CAMPER ANNOTATION

```{r load-mag-annotations}

Genes_withMAG_NORMALIZED <- load_genes_ps()
tax_df <- data.frame(tax_table(readRDS(p_raw_req(MAG_ABUNDANCE_RDS))))

df_CAMPER <- subset_taxa(Genes_withMAG_NORMALIZED, CAMPER_id != "") |>
  subset_taxa(!is.na(CAMPER_module)) |>
  psmelt() |>
  dplyr::select(SPRUCE_ID, CAMPER_id, CAMPER_oxygen) |>
  group_by(SPRUCE_ID, CAMPER_oxygen) |>
  summarise(num_CAMPER_ids_detected = n_distinct(CAMPER_id), .groups = "drop") |>
  pivot_wider(names_from = CAMPER_oxygen, values_from = num_CAMPER_ids_detected) |>
  replace_na(list(oxic = 0, anoxic = 0)) |>
  mutate(Specialist_phenol_decomposition = case_when(
    anoxic / (anoxic + oxic) >= 0.6 ~ "Anoxic_phenol_decomposition",
    oxic   / (anoxic + oxic) >= 0.6 ~ "Oxic_phenol_decomposition",
    TRUE                            ~ "Ambivalent_phenol_decomposition"))
write_out(df_CAMPER, p_meta("df_CAMPER.csv"))

tax_df$SPRUCE_ID_num <- as.numeric(gsub("SPRUCE (\\d+)", "\\1", tax_df$SPRUCE_ID))
cat("New MAGs (ID >= 2000):", sum(tax_df$SPRUCE_ID_num >= 2000, na.rm = TRUE), "\n")

# ------------------------------------------------------------------------------
# MAG set reconciliation
# ------------------------------------------------------------------------------
# The Results state that polyphenol degradation genes are nearly ubiquitous,
# ">99% encode oxygen-dependent or oxygen-independent enzymes". The numerator of
# that percentage is the number of genomes in df_CAMPER, which is derived from
# the gene object. The paper's denominator is the 294 dereplicated MAGs in the
# abundance and activity objects. The two are not the same set, so the block
# below reports every count and both percentages rather than assuming they agree.

mag_ids <- function(ps) {
  tt <- data.frame(tax_table(ps))
  if (!"SPRUCE_ID" %in% names(tt)) return(character(0))
  unique(na.omit(as.character(tt$SPRUCE_ID)))
}
mag_sets <- list(
  gene_object       = mag_ids(Genes_withMAG_NORMALIZED),
  abundance_object  = unique(na.omit(as.character(tax_df$SPRUCE_ID))),
  activity_object   = mag_ids(readRDS(p_raw_req(MAG_ACTIVITY_RDS))),
  df_CAMPER         = unique(as.character(df_CAMPER$SPRUCE_ID)))

mag_set_sizes <- data.frame(object = names(mag_sets),
                            n_MAGs = vapply(mag_sets, length, integer(1)),
                            row.names = NULL)
print(mag_set_sizes, row.names = FALSE)

mag_pairs <- list(c("gene_object", "abundance_object"),
                  c("df_CAMPER",   "abundance_object"),
                  c("df_CAMPER",   "gene_object"),
                  c("activity_object", "abundance_object"))
mag_set_diffs <- purrr::map_dfr(mag_pairs, function(pr) {
  a <- mag_sets[[pr[1]]]; b <- mag_sets[[pr[2]]]
  data.frame(set_A = pr[1], set_B = pr[2],
             n_A = length(a), n_B = length(b),
             n_shared    = length(intersect(a, b)),
             n_only_in_A = length(setdiff(a, b)),
             n_only_in_B = length(setdiff(b, a)),
             only_in_A = paste(sort(setdiff(a, b)), collapse = "; "),
             only_in_B = paste(sort(setdiff(b, a)), collapse = "; "))
})
print(mag_set_diffs[, 1:7], row.names = FALSE)

# The percentage as it can be computed on each candidate denominator.
n_camper <- length(intersect(mag_sets$df_CAMPER, mag_sets$gene_object))
camper_pct <- data.frame(
  denominator = c("genomes in the gene object", "294 MAGs in the abundance object"),
  n_denominator = c(length(mag_sets$gene_object), length(mag_sets$abundance_object)),
  n_with_phenol_active_gene = c(
    n_camper,
    length(intersect(mag_sets$df_CAMPER, mag_sets$abundance_object))))
camper_pct$percent <- round(100 * camper_pct$n_with_phenol_active_gene /
                            camper_pct$n_denominator, 1)
print(camper_pct, row.names = FALSE)

if (length(setdiff(mag_sets$gene_object, mag_sets$abundance_object)))
  message("The gene object carries ", length(setdiff(mag_sets$gene_object,
          mag_sets$abundance_object)), " genome(s) absent from the 294-MAG set.")
```

## 2.5 ENZYME ASSAY DATA

```{r load-enzyme-data}

ABTS <- read.csv(p_raw_req("ABTS_fullDataset.csv"), row.names = 1) |>
  (\(x) x[x$Depth == "10-20 cm", c(1, 6, 18, 28, 29)])() |>
  mutate(Year = as.factor(year), Plot = as.factor(Plot), Depth = as.factor(Depth))

enz2 <- read.csv(p_raw_req("ExoEnz_fullDataset.csv"), row.names = 1) |>
  (\(x) x[x$Duplicate != "Yes" & x$Depth == "10-20 cm", c(1, 3, 8:11, 21, 22)])() |>
  dplyr::select(-Duplicate) |>
  dplyr::rename(Year = 7) |>
  mutate(Year = as.factor(Year), Plot = as.factor(Plot)) |>
  left_join(metadata_df, by = c("Plot", "Year", "Depth")) |>
  filter(!is.na(Warming_Treatment)) |>
  mutate(BG.h_log  = log(BG.h),  CB.h_log   = log(CB.h),
         NAG.h_log = log(NAG.h), PHOS.h_log = log(PHOS.h))

write_out(ABTS, p_meta("ABTS.csv"))
write_out(enz2, p_meta("enz2.csv"))
```

## 2.6 ENZYME MODELS AND SUPPLEMENTARY FIGURE S3

Gamma GLMM (log link) for phenol oxidase, Gaussian GLMM on log-transformed
hydrolase activity. Selection follows the porewater gas models in section 2.10:
the temperature by drawdown interaction is kept only where the LRT supports it,
and the random intercept is dropped when its variance is estimated at zero.

`CO2_Treatment` is fitted as a fixed effect here and is not in the dbRDA. The
dbRDA conditions on enclosure with `Condition(Plot)`, which removes all
among-enclosure variation, and CO2 is assigned per enclosure, so it is removed
along with it and cannot be estimated. These models put enclosure in as a random
intercept instead, which models the among-enclosure variation rather than
sweeping it out, so an enclosure-level fixed effect stays estimable.

```{r enzyme-models-and-fig}

enz_ABTS <- left_join(ABTS, metadata_df, by = c("Plot", "Year", "Depth")) |>
  filter(!is.na(Warming_Treatment)) |>
  group_by(Name) |>
  mutate(mean_ActivtypH = mean(ActivtypH, na.rm = TRUE)) |>
  ungroup() |>
  distinct(Sample, .keep_all = TRUE)

POX_LABEL <- "Phenol oxidase (Gamma, log link)"
pox_lrt <- lrt_interaction(
  mean_ActivtypH ~ Temp_avg_top + WT_drawdown_cm + CO2_Treatment +
    Temp_avg_top:WT_drawdown_cm + (1 | Plot),
  mean_ActivtypH ~ Temp_avg_top + WT_drawdown_cm + CO2_Treatment + (1 | Plot),
  enz_ABTS, family = Gamma(link = "log"), label = "Phenol oxidase")
pox_fixed <- drop_singular_re(pox_lrt$model, enz_ABTS, POX_LABEL)
pox <- list(model = pox_fixed$model, dropped_re = pox_fixed$dropped_re,
            lrt_p = pox_lrt$lrt_p)
register_model(tidy_glmmtmb(pox$model, POX_LABEL))
print(tidy_glmmtmb(pox$model, POX_LABEL))

fit_enz <- function(resp, label) {
  f <- function(rhs) as.formula(paste(resp, "~", rhs))
  r <- lrt_interaction(
    f("Temp_avg_top + WT_drawdown_cm + CO2_Treatment + Temp_avg_top:WT_drawdown_cm + (1|Plot)"),
    f("Temp_avg_top + WT_drawdown_cm + CO2_Treatment + (1|Plot)"),
    enz2, family = gaussian(), label = label)
  fixed <- drop_singular_re(r$model, enz2, label)
  register_model(tidy_glmmtmb(fixed$model, label))
  list(model = fixed$model, dropped_re = fixed$dropped_re, lrt_p = r$lrt_p)
}
res_BG   <- fit_enz("BG.h_log",   "beta-glucosidase (log)")
res_CB   <- fit_enz("CB.h_log",   "Cellobiohydrolase (log)")
res_NAG  <- fit_enz("NAG.h_log",  "beta-N-acetylglucosaminidase (log)")
res_PHOS <- fit_enz("PHOS.h_log", "Acid phosphatase (log)")

FIG_S3A <- ggplot(enz_ABTS, aes(x = WT_drawdown_cm, y = mean_ActivtypH)) +
  geom_point(aes(fill = Warming_Treatment.ord, shape = Year), colour = "black", size = 4) +
  geom_smooth(method = "lm", color = "black", linetype = "dashed", linewidth = 1, se = FALSE) +
  scale_shape_manual(values = c(22, 24)) + scale_fill_manual(values = allGroupsColors) +
  guides(fill = guide_legend("Temperature\ntreatment", override.aes = list(shape = 21))) +
  xlab("Water table drawdown (cm)") +
  ylab(expression(Phenol ~ oxidase ~ activity ~ "(" * mu * "mol/kg*h)")) +
  theme_bw() +
  theme(legend.position = "none", axis.text = element_text(size = 16),
        axis.title = element_text(size = 16))

enz2.long <- melt(enz2, id = c("WT_drawdown", "Warming_Treatment.ord", "Year"),
                  measure = c("BG.h", "CB.h", "NAG.h", "PHOS.h"))
colnames(enz2.long)[4] <- "Hydrolase"
levels(enz2.long$Hydrolase) <- c("beta-glucosidase", "cellobiohydrolase",
                                 "beta-N-acetylglucosaminidase", "acid phosphatase")

FIG_S3B <- ggplot(enz2.long, aes(x = WT_drawdown * 100, y = value, fill = Hydrolase)) +
  geom_smooth(aes(color = Hydrolase), method = "lm", se = FALSE,
              linetype = "dashed", linewidth = 1) +
  geom_point(aes(fill = Hydrolase, shape = Year), size = 5) +
  scale_shape_manual(values = c(22, 24)) +
  guides(shape = "none",
         fill  = guide_legend("Hydrolase", nrow = 4, byrow = TRUE,
                              override.aes = list(shape = 21))) +
  scale_fill_npg() + scale_color_npg() +
  scale_y_continuous(trans = "log", labels = scales::label_number(accuracy = 0.01)) +
  ylab(expression(Hydrolase ~ activity ~ "(" * mu * "mol/kg*h)")) +
  xlab("Water table drawdown (cm)") +
  theme_bw() +
  theme(legend.position = "bottom", axis.text = element_text(size = 16),
        axis.title = element_text(size = 16))

save_fig(ggarrange(FIG_S3A, FIG_S3B, ncol = 2, labels = c("A", "B"),
                   font.label = list(size = 18)),
         p_fig("Figure_S3_enzyme_activity.pdf"), width = 10, height = 5)
```

## 2.7 POREWATER PHENOLICS

```{r load-phenolics}

Spruce.phenolics1 <- read.csv(p_raw_req("Spruce_phenolics.csv")) |>
  separate(Date, c("Year", "Month", "Day"), sep = "-") |>
  filter(Year %in% c("2020", "2021", "2022"), Depth %in% c("25")) |>
  mutate(Plot = as.factor(gsub("[A-Za-z]+(\\d+)", "\\1", Plot)), Year = as.factor(Year)) |>
  dplyr::select(Year, Plot, Gallic_acid_conc_mg_L) |>
  group_by(Plot, Year) |>
  summarise(across(everything(), \(x) mean(x, na.rm = TRUE)), .groups = "drop") |>
  left_join(metadata_df, by = c("Plot", "Year")) |>
  na.omit() |>
  as.data.frame()

cat("Gallic acid Shapiro P =",
    round(shapiro.test(Spruce.phenolics1$Gallic_acid_conc_mg_L)$p.value, 3), "\n")

phen <- lrt_interaction(
  Gallic_acid_conc_mg_L ~ Temp_avg_top + WT_drawdown_cm + Temp_avg_top:WT_drawdown_cm + (1|Plot),
  Gallic_acid_conc_mg_L ~ Temp_avg_top + WT_drawdown_cm + (1|Plot),
  Spruce.phenolics1, label = "Gallic acid")
phen_fix <- drop_singular_re(phen$model, Spruce.phenolics1, "Gallic acid")
step_phenol <- phen_fix$model
register_model(tidy_glmmtmb(step_phenol, "Gallic-acid-equivalent reducing capacity"))
print(tidy_glmmtmb(step_phenol, "Gallic-acid-equivalent reducing capacity"))
```

## 2.8 METABOLOMICS: NOSC

NOSC = -((-z + 4C + H - 3N - 2O + 5P - 2S) / C) + 4, with z = 0, from the KEGG
formula. Formulae that will not parse, or that contain no carbon, return NA and
are excluded. The sample-level value is the relative-abundance-weighted mean.

```{r load-metabolomics-nosc}

metabolites_phylo <- read_metab_ps()

if (!identical(taxa_names(metabolites_phylo),
               unname(as.character(tax_table(metabolites_phylo)[, "Feature"]))))
  stop("taxa_names() and the Feature column of the metabolite tax table are not ",
       "the same vector.")

calculate_nosc <- function(formula, z = 0) {
  if (is.na(formula) || formula == "") return(NA_real_)
  el <- str_match_all(formula, "([A-Z][a-z]*)(\\d*)")[[1]]
  if (nrow(el) == 0) return(NA_real_)
  n <- as.numeric(el[, 3]); n[is.na(n)] <- 1
  cnt <- setNames(n, el[, 2])
  g <- function(s) if (s %in% names(cnt)) cnt[[s]] else 0
  C <- g("C"); if (C == 0) return(NA_real_)
  -((-z + 4*C + g("H") - 3*g("N") - 2*g("O") + 5*g("P") - 2*g("S")) / C) + 4
}

metabolites_phylo_n <- transform_sample_counts(metabolites_phylo, function(x) x / sum(x))
tax_table_metab      <- as.data.frame(tax_table(metabolites_phylo_n))
tax_table_metab$NOSC <- vapply(tax_table_metab$KEGG_formula, calculate_nosc, numeric(1))
tax_table(metabolites_phylo_n) <- as.matrix(tax_table_metab)
tax_table(metabolites_phylo)   <- as.matrix(tax_table_metab)

cat("Features with a usable NOSC:", sum(!is.na(tax_table_metab$NOSC)),
    "of", nrow(tax_table_metab), "\n")

rel_mat <- as(otu_table(metabolites_phylo_n), "matrix")
if (!taxa_are_rows(metabolites_phylo_n)) rel_mat <- t(rel_mat)   # features x samples
nosc_vec <- tax_table_metab$NOSC[match(rownames(rel_mat), rownames(tax_table_metab))]
usable   <- !is.na(nosc_vec)
if (!any(usable))
  stop("No metabolite feature has a usable NOSC. Check that KEGG_formula is ",
       "populated and that tax_table rownames match the otu_table.")
sample_nosc <- colSums(rel_mat[usable, , drop = FALSE] * nosc_vec[usable])
cat(sprintf("NOSC: %d of %d features usable, covering %.1f%% of relative abundance\n",
            sum(usable), length(usable),
            100 * mean(colSums(rel_mat[usable, , drop = FALSE]))))

sample_df <- data.frame(sample_data(metabolites_phylo_n))
sample_df <- sample_df[, !grepl("^NOSC", names(sample_df)), drop = FALSE]
sample_df$NOSC <- unname(sample_nosc[rownames(sample_df)])
if (all(is.na(sample_df$NOSC)))
  stop("NOSC did not map onto the samples. sample_data rownames are: ",
       paste(head(rownames(sample_df), 3), collapse = ", "),
       "; otu_table sample names are: ",
       paste(head(colnames(rel_mat), 3), collapse = ", "))
sample_data(metabolites_phylo_n) <- sample_df
sample_data(metabolites_phylo)   <- sample_df

# ------------------------------------------------------------------------------
# 2.8b  NOSC restricted to Level 1-3 annotations
# ------------------------------------------------------------------------------

# Levels 1-3: features with at least a computational structure proposal.
# Excluded: L4 (formula only, no name) and unannotated features.
HIGH_CONF       <- "^L1$|^L2$|^L3$"
HIGH_CONF_LABEL <- "Level 1-3"

conf_vec <- tax_table_metab$Annotation_Confidence[
  match(rownames(rel_mat), rownames(tax_table_metab))]
conf_lab <- ifelse(is.na(conf_vec) | conf_vec %in% c("", "NA"), "unannotated", conf_vec)

nosc_by_conf <- data.frame(
    Annotation_Confidence = conf_lab,
    has_formula           = usable,
    rel_abundance         = rowMeans(rel_mat)) |>
  dplyr::group_by(Annotation_Confidence) |>
  dplyr::summarise(n_features        = dplyr::n(),
                   n_with_NOSC       = sum(has_formula),
                   pct_with_NOSC     = round(100 * mean(has_formula), 1),
                   mean_rel_abund    = signif(sum(rel_abundance), 3),
                   .groups = "drop") |>
  dplyr::arrange(dplyr::desc(n_with_NOSC))
cat("\n=== NOSC coverage by annotation confidence ===\n")
print(as.data.frame(nosc_by_conf))

high <- usable & grepl(HIGH_CONF, conf_lab)
cat(sprintf("NOSC features at %s: %d of %d usable (%.1f%%)\n",
            HIGH_CONF_LABEL, sum(high), sum(usable), 100 * sum(high) / sum(usable)))

if (sum(high) >= 5) {
  sample_nosc_high <- colSums(rel_mat[high, , drop = FALSE] * nosc_vec[high])
  wt_high <- colSums(rel_mat[high, , drop = FALSE])
  sample_nosc_high_norm <- sample_nosc_high / wt_high

  sens <- data.frame(
    Sample          = names(sample_nosc),
    NOSC_all        = as.numeric(sample_nosc),
    NOSC_L1L3       = as.numeric(sample_nosc_high[names(sample_nosc)]),
    NOSC_L1L3_norm  = as.numeric(sample_nosc_high_norm[names(sample_nosc)]))
  r_raw  <- cor(sens$NOSC_all, sens$NOSC_L1L3,      use = "complete.obs")
  r_norm <- cor(sens$NOSC_all, sens$NOSC_L1L3_norm, use = "complete.obs")
  cat(sprintf("Sample NOSC, all formulae vs Level 1-3 only: r = %.3f (raw), %.3f (renormalised)\n",
              r_raw, r_norm))

  sens_dat <- sample_df
  sens_dat$NOSC_L1L3 <- sens$NOSC_L1L3_norm[match(rownames(sens_dat), sens$Sample)]
  sens_fit <- try(glmmTMB::glmmTMB(NOSC_L1L3 ~ Temp_avg_top + WT_drawdown_cm,
                                   data = sens_dat), silent = TRUE)
  if (!inherits(sens_fit, "try-error")) {
    cat("\n--- NOSC (Level 1-3 only) ~ Temp + WT ---\n")
    print(tidy_glmmtmb(sens_fit, "Sample NOSC (Level 1-3 annotations only)"))
    register_model(tidy_glmmtmb(sens_fit, "Sample NOSC (Level 1-3 annotations only)"))
  }
} else {
  message("Fewer than 5 NOSC features at Level 1-3; sensitivity analysis skipped.")
}

write_out(sample_df, p_meta("metadata_with_NOSC.csv"))
saveRDS(metabolites_phylo, p_ps("PhyloseqObj_metabolites_withNOSC.RDS"))

nosc_data <- sample_df |> mutate(Plot = as.factor(Plot))
nosc <- lrt_interaction(
  NOSC ~ Temp_avg_top + WT_drawdown_cm + Temp_avg_top:WT_drawdown_cm + (1|Plot),
  NOSC ~ Temp_avg_top + WT_drawdown_cm + (1|Plot),
  nosc_data, label = "NOSC")
nosc_fix  <- drop_singular_re(nosc$model, nosc_data, "NOSC")
step_nosc <- nosc_fix$model
register_model(tidy_glmmtmb(step_nosc,
  paste0("Sample NOSC", if (nosc_fix$dropped_re) " (enclosure RE dropped, singular)" else "")))
print(tidy_glmmtmb(step_nosc, "Sample NOSC"))
```

## 2.9 MAASLIN2: METABOLITE RESPONSES

Primary model: both predictors in one Maaslin2 fit with the enclosure random
effect, `Temp_avg_top + WT_drawdown_cm + (1|Plot)`. Supplementary Data 3 and 4,
the FDR summary and the class-level tests come from this fit. The two
single-predictor fixed-effects fits are a sensitivity check.

```{r maaslin-metabolites}

# ------------------------------------------------------------------------------
# 2.9.1  Feature table and annotations
# ------------------------------------------------------------------------------

ps_met  <- read_metab_ps(quiet = TRUE)
keep    <- taxa_sums(ps_met) > 0
if (any(!keep)) ps_met <- prune_taxa(keep, ps_met)
cat("Metabolite features tested:", ntaxa(ps_met), "\n")

taxdf_met <- data.frame(tax_table(ps_met))
taxdf_met$feature <- rownames(taxdf_met)

# Mode, retention time and precursor m/z out of the feature ID; every compound
# name labelled with how many features carry it. The diastereomer alias and the
# "M" sentinel were handled in read_metab_ps(), so they are not repeated here.
taxdf_met <- cbind(taxdf_met, parse_feature_id(taxdf_met$feature))
taxdf_met <- add_name_multiplicity(taxdf_met)

print(as.data.frame(METAB_MSI_CENSUS), row.names = FALSE)

# --- Compound Discoverer per-feature values -----------------------------------
# Supplied as a keyed CSV in 00_inputs/ and left-joined at the point of use.
annot_path <- p_raw(METAB_ANNOT_UPDATE_FILE)
if (file.exists(annot_path)) {
  annot_upd <- readr::read_csv(annot_path, show_col_types = FALSE)
  stopifnot("feature" %in% names(annot_upd))


  if ("Area_CV_percent_CD" %in% names(annot_upd)) {
    taxdf_met <- dplyr::left_join(
      taxdf_met, annot_upd[, c("feature", "Area_CV_percent_CD")], by = "feature")
    taxdf_met$Area_CV_percent    <- round(taxdf_met$Area_CV_percent_CD, 1)
    taxdf_met$Area_CV_percent_CD <- NULL
    cov_n <- sum(!is.na(taxdf_met$Area_CV_percent))
    cat(sprintf("Area CV%%: supplied for %d of %d features (%.1f%%)\n",
                cov_n, nrow(taxdf_met), 100 * cov_n / nrow(taxdf_met)))
    if (cov_n < nrow(taxdf_met))
      message("  Area CV% is partial. The published CV figure and by-level table ",
              "must state their denominator, or wait for the full CD export.")
  }
} else {
  message(METAB_ANNOT_UPDATE_FILE, " not in 00_inputs/. Supplementary Data 3 and 4 ",
          "will be written without Area_CV_percent.")
}

met_tax_cols <- c("feature", "Acquisition_mode", "RT_min", "precursor_mz",
                  "Final_name", "Name_n_features", "Name_status",
                  "Kingdom", "Superclass", "Class", "Subclass",
                  "Parent.Level.1", "Parent.Level.2", "Parent.Level.3",
                  "Parent.Level.4", "Parent.Level.5", "Annotation_Confidence")
if ("Area_CV_percent" %in% names(taxdf_met))
  met_tax_cols <- c(met_tax_cols, "Area_CV_percent")

cat(sprintf("Acquisition modes present: %s\n",
            paste(sort(unique(na.omit(taxdf_met$Acquisition_mode))), collapse = ", ")))

# Census of compound names carried by more than one feature.
dup_names <- taxdf_met |>
  dplyr::filter(!is.na(Name_n_features), Name_n_features > 1) |>
  dplyr::select(Final_name, feature, Acquisition_mode, RT_min, precursor_mz,
                Annotation_Confidence, Name_n_features) |>
  dplyr::arrange(dplyr::desc(Name_n_features), Final_name, RT_min)
# The Supplementary Methods quote this count. The denominator is every feature in
# the table, not the significant subset; the unnamed sentinel is already NA here.
cat(sprintf(paste0("Compound names carried by >1 feature: %d names, %d features ",
                   "(max %d per name), out of %d named features and %d features total\n"),
            dplyr::n_distinct(dup_names$Final_name), nrow(dup_names),
            if (nrow(dup_names)) max(dup_names$Name_n_features) else 0L,
            sum(!is.na(taxdf_met$Final_name)), nrow(taxdf_met)))

# ------------------------------------------------------------------------------
# 2.9.1b  Normalisation identity check
# ------------------------------------------------------------------------------

stored_mat <- as(otu_table(ps_met), "matrix")
if (!taxa_are_rows(ps_met)) stored_mat <- t(stored_mat)
cat(sprintf("Stored intensities: %d features x %d samples, range %.4g to %.4g, %d zeros\n",
            nrow(stored_mat), ncol(stored_mat), min(stored_mat), max(stored_mat),
            sum(stored_mat == 0)))
cat("  No zeros and a", sprintf("%.1f", max(stored_mat) / min(stored_mat)),
    "fold total range: these are log2 values, not peak areas.\n")

raw_auc_path <- p_raw(RAW_AUC_FILE)
if (!file.exists(raw_auc_path))
  stop("A feature-wise coefficient of variation requires the raw peak areas. ",
       "The stored intensities are log2 values and a CV computed on them is ",
       "meaningless. Copy ", RAW_AUC_FILE, " into 00_inputs/ and re-knit, or ",
       "delete this block and the CV panel from the manuscript.")

# The first column of raw_auc_table.csv holds the feature IDs and has no header.
# Take it positionally.
raw_tbl <- readr::read_csv(raw_auc_path, show_col_types = FALSE)
raw_auc <- as.matrix(as.data.frame(raw_tbl)[, -1, drop = FALSE])
rownames(raw_auc) <- as.character(raw_tbl[[1]])
storage.mode(raw_auc) <- "double"
cat(sprintf("  Raw AUC table: %d features x %d samples, range %.4g to %.4g, %d zeros\n",
            nrow(raw_auc), ncol(raw_auc), min(raw_auc), max(raw_auc),
            sum(raw_auc == 0)))

# Check that the stored intensities equal log2(2 * M * x / m_j).
shared_f <- intersect(rownames(raw_auc), rownames(stored_mat))
shared_s <- intersect(colnames(raw_auc), colnames(stored_mat))
if (length(shared_f) < 100 || length(shared_s) < 20)
  stop("Only ", length(shared_f), " feature(s) and ", length(shared_s),
       " sample(s) are shared between ", RAW_AUC_FILE, " and ", METAB_PS_FILE,
       ". Check the identifier conventions on both. Stored sample names are: ",
       paste(utils::head(colnames(stored_mat), 3), collapse = ", "),
       "; raw table sample names are: ",
       paste(utils::head(colnames(raw_auc), 3), collapse = ", "), ".")
m_j  <- colMeans(raw_auc)
M    <- mean(m_j)
recon <- log2(2 * M * sweep(raw_auc[shared_f, shared_s, drop = FALSE], 2,
                            m_j[shared_s], "/"))
norm_err <- max(abs(recon - stored_mat[shared_f, shared_s, drop = FALSE]))
cat(sprintf("  Mean-normalisation check on %d x %d shared cells: max |error| = %.3g\n",
            length(shared_f), length(shared_s), norm_err))
if (norm_err > 1e-6)
  warning("The stored intensities are not log2(2 * M * x / m_j) to within 1e-6. ",
          "The Methods description of the normalisation is wrong; max error ",
          signif(norm_err, 4), ".")


# ------------------------------------------------------------------------------
# 2.9.2  Molecular weight from the KEGG formula
# ------------------------------------------------------------------------------

ATOMIC_MASS <- c(C = 12.000, H = 1.00783, N = 14.00307, O = 15.99491,
                 P = 30.97376, S = 31.97207, Cl = 34.96885, Br = 78.91834,
                 F = 18.99840, I = 126.90447, Na = 22.98977, K = 38.96371)

calculate_mw <- function(formula) {
  if (is.na(formula) || formula == "") return(NA_real_)
  el <- stringr::str_match_all(formula, "([A-Z][a-z]*)(\\d*)")[[1]]
  if (nrow(el) == 0) return(NA_real_)
  n <- as.numeric(el[, 3]); n[is.na(n)] <- 1
  if (any(!el[, 2] %in% names(ATOMIC_MASS))) return(NA_real_)
  sum(ATOMIC_MASS[el[, 2]] * n)
}
taxdf_met$MW <- vapply(taxdf_met$KEGG_formula, calculate_mw, numeric(1))

# ------------------------------------------------------------------------------
# 2.9.2b  Phenolic pool split by aromatic ring count
# ------------------------------------------------------------------------------

PHENOLIC_SINGLE_RING <- c(
  "Benzene and substituted derivatives",
  "Phenols",
  "Phenol ethers",
  "Cinnamic acids and derivatives",
  "Cinnamaldehydes",
  "Benzaldehydes",
  "Benzoic acids and derivatives",
  "Phenylpropanoic acids",
  "Anisoles",
  "Benzenesulfonic acids and derivatives",
  "Toluenes",
  "Phenylacetic acids")

PHENOLIC_MULTI_RING <- c(
  "Flavonoids",
  "Isoflavonoids",
  "2-arylbenzofuran flavonoids",
  "Neoflavonoids",
  "Stilbenes",
  "Naphthalenes",
  "Naphthofurans",
  "Coumarins and derivatives",
  "Isocoumarins and derivatives",
  "Benzopyrans",
  "Benzofurans",
  "Linear 1,3-diarylpropanoids",
  "Diarylheptanoids",
  "Anthracenes",
  "Anthracyclines",
  "Xanthones",
  "Depsides and depsidones",
  "Tannins",
  "Indanes",
  "Fluorenes",
  "Biphenyls and derivatives")

taxdf_met$Ring_class <- dplyr::case_when(
  taxdf_met$Superclass == "Lignans, neolignans and related compounds" ~ "Multi-ring",
  taxdf_met$Class %in% PHENOLIC_MULTI_RING                            ~ "Multi-ring",
  taxdf_met$Class %in% PHENOLIC_SINGLE_RING                           ~ "Single-ring",
  TRUE                                                                ~ NA_character_)

taxdf_met$Poly_struct_set <- ifelse(
  taxdf_met$Superclass %in% POLY_SUPERCLASSES & !is.na(taxdf_met$Ring_class),
  paste("Phenolic -", taxdf_met$Ring_class), NA_character_)

# Free amino acids separated from peptides. ClassyFire puts both under the class
# "Carboxylic acids and derivatives" and the subclass "Amino acids, peptides, and
# analogues", so a test at class level mixes them. Parent.Level.1 splits them.
# Every other feature is the background, so "Amino acids" is tested against the
# rest of the metabolome rather than against peptides alone.
if (!"Parent.Level.1" %in% names(taxdf_met))
  stop("taxdf_met has no Parent.Level.1 column, so free amino acids cannot be ",
       "separated from peptides. Columns present: ",
       paste(names(taxdf_met), collapse = ", "))
taxdf_met$AA_set <- dplyr::case_when(
  taxdf_met$Parent.Level.1 %in% "Amino acids and derivatives" ~ "Amino acids",
  taxdf_met$Parent.Level.1 %in% "Peptides"                    ~ "Peptides",
  TRUE                                                        ~ "Rest of metabolome")

cat("\n=== Amino acid set composition ===\n")
print(taxdf_met |>
  dplyr::filter(AA_set != "Rest of metabolome") |>
  dplyr::count(AA_set, Class, Subclass, Parent.Level.2, name = "n_features") |>
  dplyr::arrange(AA_set, dplyr::desc(n_features)) |>
  as.data.frame(), row.names = FALSE)
cat("Features: amino acids", sum(taxdf_met$AA_set == "Amino acids"),
    "| peptides", sum(taxdf_met$AA_set == "Peptides"),
    "| background", sum(taxdf_met$AA_set == "Rest of metabolome"), "\n")

# Carbohydrates. Figure 5 enriches glycolysis, carbon metabolism and pyruvate
# metabolism, and every WS-OX path to a methanogen in Figure 3 runs through a
# carbohydrate. The ClassyFire class "Organooxygen compounds" is a poor proxy
# because it also holds alcohols, polyols and carbonyls, so the set is taken at
# subclass level.
if (!"Subclass" %in% names(taxdf_met))
  stop("taxdf_met has no Subclass column; the carbohydrate set cannot be built.")
taxdf_met$Carb_set <- ifelse(
  taxdf_met$Subclass %in% "Carbohydrates and carbohydrate conjugates",
  "Carbohydrates", "Rest of metabolome")

cat("\n=== Carbohydrate set ===\n")
cat("Features: carbohydrates", sum(taxdf_met$Carb_set == "Carbohydrates"),
    "| background", sum(taxdf_met$Carb_set == "Rest of metabolome"), "\n")
if (sum(taxdf_met$Carb_set == "Carbohydrates") < 8) {
  message("Fewer than 8 carbohydrate features. The competitive test is reported ",
          "but should not be interpreted.")
} else {
  print(taxdf_met |>
    dplyr::filter(Carb_set == "Carbohydrates") |>
    dplyr::count(Class, Subclass, Annotation_Confidence, name = "n_features") |>
    dplyr::arrange(dplyr::desc(n_features)) |>
    as.data.frame(), row.names = FALSE)
}

poly_all <- dplyr::filter(taxdf_met, Superclass %in% POLY_SUPERCLASSES)
unassigned <- sort(unique(poly_all$Class[is.na(poly_all$Ring_class)]))
if (length(unassigned))
  message("Phenolic Classes with no ring assignment (",
          sum(is.na(poly_all$Ring_class)), " features): ",
          paste(unassigned, collapse = "; "),
          ". Add each to PHENOLIC_SINGLE_RING or PHENOLIC_MULTI_RING.")

ring_summary <- poly_all |>
  dplyr::mutate(Ring_class = ifelse(is.na(Ring_class), "Unassigned", Ring_class)) |>
  dplyr::group_by(Ring_class) |>
  dplyr::summarise(n_features = dplyr::n(),
                   n_classes  = dplyr::n_distinct(Class),
                   median_MW  = round(median(MW, na.rm = TRUE), 1),
                   q25_MW     = round(quantile(MW, .25, na.rm = TRUE), 1),
                   q75_MW     = round(quantile(MW, .75, na.rm = TRUE), 1),
                   .groups    = "drop")
cat("\n=== Phenolic pool, structural split ===\n")
print(as.data.frame(ring_summary))

# ------------------------------------------------------------------------------
# 2.9.3  Primary model
# ------------------------------------------------------------------------------

if (REFIT_MAASLIN) {
  fit_maaslin(ps_met, "MaAslin_metab_multi",
              fixed = c("Temp_avg_top", "WT_drawdown_cm"), random = "Plot",
              transform = "NONE")
}
MaAslin_metab_multi <- readRDS(maaslin_path("MaAslin_metab_multi.rds"))
stopifnot(all(c("Temp_avg_top", "WT_drawdown_cm") %in%
              MaAslin_metab_multi$results$metadata))

# ------------------------------------------------------------------------------
# 2.9.4  Supplementary Data 3 and 4
# ------------------------------------------------------------------------------

build_suppl_metab <- function(res, predictor_name, tag) {
  res$results |>
    dplyr::rename(feature = 1) |>
    dplyr::filter(metadata == predictor_name, pval < 0.05) |>
    add_effect_size() |>
    dplyr::mutate(metadata = tag) |>
    dplyr::left_join(taxdf_met[, c(met_tax_cols, "MW", "Ring_class")], by = "feature") |>
    dplyr::select(feature, Acquisition_mode, RT_min, precursor_mz,
                  metadata, coef, stderr, CI_lower, CI_upper, pval, qval,
                  evidence, N, dplyr::any_of(c("N.not.zero", "N.not.0")),
                  Final_name, Name_n_features, Name_status,
                  Kingdom, Superclass, Class, Subclass,
                  dplyr::starts_with("Parent.Level"), Annotation_Confidence,
                  dplyr::any_of("Area_CV_percent"), MW, Ring_class) |>
    dplyr::arrange(coef)
}

# Column definitions carried alongside Supplementary Data 3 and 4. Acquisition
# mode, retention time and precursor m/z are explicit columns rather than being
# encoded only in the feature ID, so repeated compound names can be assessed as
# isomers, adducts or duplicates.
SD34_HEADER_NOTE <- data.frame(column = c(
  "feature", "Acquisition_mode", "RT_min", "precursor_mz",
  "Final_name", "Name_n_features", "Name_status", "Annotation_Confidence",
  "Area_CV_percent"),
  definition = c(
    "Feature identifier, '<acquisition mode>_<retention time in min>_<precursor m/z>'.",
    "Chromatographic method and polarity the feature was detected in.",
    "Chromatographic retention time, minutes.",
    "Precursor m/z.",
    paste("Putative compound name. A name is an assignment, not an identification,",
          "unless Annotation_Confidence is L1. Stereoisomeric assignments that",
          "in-silico methods cannot separate are reported once, without a",
          "stereodescriptor."),
    "Number of features in the full table carrying this same compound name.",
    paste("Whether the name is unique to this feature, shared with others, or the",
          "feature is unnamed. Features sharing a name are not replicates; compare",
          "RT_min and precursor_mz to judge whether they are isomers or adducts."),
    paste("Metabolomics Standards Initiative level. L1, match to an authentic",
          "standard on precursor m/z (5 ppm), MS/MS spectrum and retention time",
          "(within 0.5 min), standards run on the same instrument, column and",
          "method as the samples. L2, full MS/MS match to mzCloud. L3, structure",
          "proposed computationally from predicted formula and compound class",
          "(SIRIUS, CSI:FingerID, CANOPUS, ClassyFire, ChemSpider). L4, molecular",
          "formula only, no putative name. L3 and L4 names are the best-matching",
          "candidate structure, not an identification."),
    paste("Compound Discoverer coefficient of variation of peak area for this",
          "compound across the sample set. This is dispersion across biological",
          "samples spanning a 0 to 9 degrees C warming gradient, three",
          "water-table regimes and three years, and is expected to be large for",
          "responsive features. It is not the pooled-QC relative standard",
          "deviation: the <30% QC RSD and <=25% correction-factor thresholds",
          "reported in the Supplementary Methods were applied to replicate",
          "injections of the pooled QC extract, which is a different quantity."))
)
# The definition list must match the columns actually written, or a reader will
# look for a column that is not there.
if (!"Area_CV_percent" %in% names(taxdf_met))
  SD34_HEADER_NOTE <- SD34_HEADER_NOTE[SD34_HEADER_NOTE$column != "Area_CV_percent", ]
write_out(SD34_HEADER_NOTE, p_sd("Suppl_Data_03_04_column_definitions.csv"))

Suppl_Data_03 <- build_suppl_metab(MaAslin_metab_multi, "WT_drawdown_cm", "WT_drawdown")
Suppl_Data_04 <- build_suppl_metab(MaAslin_metab_multi, "Temp_avg_top",   "Temp_avg_top")
write_out(Suppl_Data_03, p_sd("Suppl_Data_03_metabolites_WT.csv"))
write_out(Suppl_Data_04, p_sd("Suppl_Data_04_metabolites_temp.csv"))

# ------------------------------------------------------------------------------
# 2.9.5  Class-level competitive test
# ------------------------------------------------------------------------------

class_level_competitive <- function(res, predictor_name, label, tax,
                                    level = "Superclass", min_n = 5) {
  df <- res$results |>
    dplyr::rename(feature = 1) |>
    dplyr::filter(metadata == predictor_name) |>
    dplyr::left_join(tax[, c("feature", level)], by = "feature") |>
    dplyr::filter(!is.na(.data[[level]]), .data[[level]] != "")

  df |>
    dplyr::group_by(set = .data[[level]]) |>
    dplyr::filter(dplyr::n() >= min_n) |>
    dplyr::group_modify(~{
      inset  <- .x
      outset <- dplyr::filter(df, .data[[level]] != .y$set[1])
      p_bg   <- mean(outset$coef > 0)
      data.frame(
        n_features  = nrow(inset),
        n_increase  = sum(inset$coef > 0),
        n_decrease  = sum(inset$coef < 0),
        median_coef = median(inset$coef),
        median_coef_background = median(outset$coef),
        p_KS     = suppressWarnings(ks.test(inset$pval, "punif")$p.value),
        p_wilcox = suppressWarnings(wilcox.test(inset$coef, outset$coef)$p.value),
        p_sign   = if (p_bg > 0 && p_bg < 1)
                     binom.test(sum(inset$coef > 0), nrow(inset), p = p_bg)$p.value
                   else NA_real_)
    }) |>
    dplyr::ungroup() |>
    dplyr::mutate(predictor = label, level = level,
                  q_KS     = p.adjust(p_KS,     method = "BH"),
                  q_wilcox = p.adjust(p_wilcox, method = "BH"),
                  q_sign   = p.adjust(p_sign,   method = "BH")) |>
    dplyr::arrange(p_wilcox)
}

class_comp <- purrr::pmap_dfr(
  list(pred  = c("WT_drawdown_cm", "Temp_avg_top", "WT_drawdown_cm", "Temp_avg_top"),
       lab   = c("WT_drawdown", "Temp_avg_top", "WT_drawdown", "Temp_avg_top"),
       lev   = c("Superclass", "Superclass", "Class", "Class"),
       min_n = c(5, 5, 8, 8)),
  function(pred, lab, lev, min_n)
    class_level_competitive(MaAslin_metab_multi, pred, lab, taxdf_met, lev, min_n))
print(as.data.frame(class_comp), digits = 3)

# Single-ring versus multi-ring phenolics, each set the other's background.
class_comp_struct <- dplyr::bind_rows(
  class_level_competitive(MaAslin_metab_multi, "WT_drawdown_cm", "WT_drawdown",
                          taxdf_met, "Poly_struct_set"),
  class_level_competitive(MaAslin_metab_multi, "Temp_avg_top", "Temp_avg_top",
                          taxdf_met, "Poly_struct_set"))
print(as.data.frame(class_comp_struct), digits = 3)

# Free amino acids and peptides, each tested against everything else.
class_comp_AA <- dplyr::bind_rows(
  class_level_competitive(MaAslin_metab_multi, "WT_drawdown_cm", "WT_drawdown",
                          taxdf_met, "AA_set"),
  class_level_competitive(MaAslin_metab_multi, "Temp_avg_top", "Temp_avg_top",
                          taxdf_met, "AA_set")) |>
  dplyr::filter(set != "Rest of metabolome")
print(as.data.frame(class_comp_AA), digits = 3)

# Carbohydrates against the rest of the metabolome.
class_comp_carb <- dplyr::bind_rows(
  class_level_competitive(MaAslin_metab_multi, "WT_drawdown_cm", "WT_drawdown",
                          taxdf_met, "Carb_set"),
  class_level_competitive(MaAslin_metab_multi, "Temp_avg_top", "Temp_avg_top",
                          taxdf_met, "Carb_set")) |>
  dplyr::filter(set != "Rest of metabolome")
print(as.data.frame(class_comp_carb), digits = 3)

# Supplementary Data 5: ClassyFire superclass and class tests plus the
# phenolic ring-count split, in one sheet.
write_out(
  dplyr::bind_rows(class_comp, class_comp_struct, class_comp_AA, class_comp_carb) |>
    dplyr::transmute(
      compound_set           = set,
      grouping               = dplyr::recode(level,
                                 Poly_struct_set = "Phenolic ring count",
                                 AA_set          = "Amino acid vs peptide",
                                 Carb_set        = "Carbohydrates",
                                 .default = level),
      predictor              = predictor,
      n_features             = n_features,
      n_increase             = n_increase,
      n_decrease             = n_decrease,
      median_coef            = median_coef,
      median_coef_background = median_coef_background,
      p_KS = p_KS, q_KS = q_KS,
      p_wilcox = p_wilcox, q_wilcox = q_wilcox,
      p_sign = p_sign, q_sign = q_sign) |>
    dplyr::arrange(grouping, predictor, q_wilcox),
  p_sd("Suppl_Data_05_classlevel_competitive.csv"))

# ------------------------------------------------------------------------------
# 2.9.6  Class-level test on class mean intensity, with the enclosure random effect
# ------------------------------------------------------------------------------

class_intensity <- {
  met_df   <- sample_matrix(ps_met)
  met_norm <- met_df / rowMeans(met_df, na.rm = TRUE)
  met_norm$SampleID <- rownames(met_norm)
  met_norm
} |>
  dplyr::left_join(
    data.frame(sample_data(ps_met)) |>
      dplyr::select(SampleID = Sample, Temp_avg_top, WT_drawdown_cm, Plot, Year),
    by = "SampleID") |>
  tidyr::pivot_longer(cols = -c(SampleID, Temp_avg_top, WT_drawdown_cm, Plot, Year),
                      names_to = "feature", values_to = "NormalizedIntensity") |>
  dplyr::inner_join(taxdf_met[, c("feature", "Superclass", "Class",
                                  "Poly_struct_set", "AA_set", "Carb_set")],
                    by = "feature")

fit_class_model <- function(dat, set_col) {
  dat |>
    dplyr::filter(!is.na(.data[[set_col]]), .data[[set_col]] != "") |>
    dplyr::group_by(SampleID, Temp_avg_top, WT_drawdown_cm, Plot, Year,
                    set = .data[[set_col]]) |>
    dplyr::summarise(MeanIntensity = mean(NormalizedIntensity, na.rm = TRUE),
                     n_features = dplyr::n(), .groups = "drop") |>
    dplyr::filter(n_features >= 5) |>
    dplyr::group_by(set) |>
    dplyr::group_modify(~{
      d <- .x
      # Predictors are standardised for conditioning and the slopes divided back
      # by the standard deviations, so coefficients stay on the per-unit scale.
      sd_t <- stats::sd(d$Temp_avg_top)
      sd_w <- stats::sd(d$WT_drawdown_cm)
      d$Temp_z <- as.numeric(scale(d$Temp_avg_top))
      d$WT_z   <- as.numeric(scale(d$WT_drawdown_cm))

      usable <- function(m) {
        if (inherits(m, "try-error") || is.null(m)) return(FALSE)
        cf <- try(summary(m)$coefficients$cond, silent = TRUE)
        !inherits(cf, "try-error") && all(c("Temp_z", "WT_z") %in% rownames(cf)) &&
          !any(is.nan(cf[, "Std. Error"])) && !any(is.na(cf[, "Std. Error"]))
      }

      m <- try(glmmTMB(MeanIntensity ~ Temp_z + WT_z + (1 | Plot), data = d),
               silent = TRUE)
      fit_type <- "mixed, scaled predictors"

      if (!usable(m)) {
        m <- try(glmmTMB(MeanIntensity ~ Temp_z + WT_z + (1 | Plot), data = d,
                         control = glmmTMBControl(optimizer = optim,
                                                  optArgs = list(method = "BFGS"))),
                 silent = TRUE)
        fit_type <- "mixed, scaled, BFGS"
      }
      if (!usable(m)) {
        m <- try(glmmTMB(MeanIntensity ~ Temp_z + WT_z, data = d), silent = TRUE)
        fit_type <- "fixed effects only (enclosure RE not estimable)"
      }
      if (!usable(m)) return(data.frame())

      cf  <- summary(m)$coefficients$cond
      r2m <- suppressWarnings(try(r.squaredGLMM(m)[1, "R2m"], silent = TRUE))
      data.frame(
        n_features      = d$n_features[1],
        coef_Temp       = cf["Temp_z", "Estimate"]   / sd_t,
        se_Temp         = cf["Temp_z", "Std. Error"] / sd_t,
        p_Temp          = cf["Temp_z", "Pr(>|z|)"],
        coef_WT         = cf["WT_z",   "Estimate"]   / sd_w,
        se_WT           = cf["WT_z",   "Std. Error"] / sd_w,
        p_WT            = cf["WT_z",   "Pr(>|z|)"],
        coef_Temp_perSD = cf["Temp_z", "Estimate"],
        coef_WT_perSD   = cf["WT_z",   "Estimate"],
        R2m             = if (inherits(r2m, "try-error")) NA_real_ else r2m,
        fit_type        = fit_type,
        converged       = TRUE)
    }) |>
    dplyr::ungroup() |>
    dplyr::mutate(q_Temp = p.adjust(p_Temp, "BH"), q_WT = p.adjust(p_WT, "BH"))
}

class_selfcontained        <- fit_class_model(class_intensity, "Superclass")
class_selfcontained_struct <- fit_class_model(class_intensity, "Poly_struct_set")
class_selfcontained_AA     <- fit_class_model(class_intensity, "AA_set")
class_selfcontained_carb   <- fit_class_model(class_intensity, "Carb_set")
print(as.data.frame(class_selfcontained),        digits = 3)
print(as.data.frame(class_selfcontained_struct), digits = 3)

cat("\n=== Class-model fit types ===\n")
print(table(class_selfcontained$fit_type))
if (any(class_selfcontained$fit_type != "mixed, scaled predictors"))
  message("Classes not fitted with the mixed model: ",
          paste(class_selfcontained$set[
            class_selfcontained$fit_type != "mixed, scaled predictors"],
            collapse = ", "))
n_missing <- dplyr::n_distinct(class_intensity$Superclass[
  !is.na(class_intensity$Superclass) & class_intensity$Superclass != ""]) -
  nrow(class_selfcontained)
if (n_missing > 0)
  warning(n_missing, " class model(s) still failed to fit and are absent from the table.")

print(as.data.frame(class_selfcontained), digits = 3, row.names = FALSE)
# class_selfcontained_struct is the class-mean model on the single-ring versus
# multi-ring split. The manuscript reports the ring split only through the
# competitive tests in Supplementary Data 5, so this fit stays in the session and
# is not written out.
print(as.data.frame(class_selfcontained_AA), digits = 3)
print(as.data.frame(class_selfcontained_AA), digits = 3, row.names = FALSE)
print(as.data.frame(class_selfcontained_carb), digits = 3)
print(as.data.frame(class_selfcontained_carb), digits = 3, row.names = FALSE)

# ------------------------------------------------------------------------------
# 2.9.7  Sensitivity: single-predictor fixed-effects fits
# ------------------------------------------------------------------------------

if (REFIT_MAASLIN) {
  fit_maaslin(ps_met, "MaAslinLM_metab_WT",   fixed = "WT_drawdown_cm", random = NULL,
              transform = "NONE")
  fit_maaslin(ps_met, "MaAslinLM_metab_temp", fixed = "Temp_avg_top",   random = NULL,
              transform = "NONE")
}
MaAslinLM_metab_WT   <- readRDS(maaslin_path("MaAslinLM_metab_WT.rds"))
MaAslinLM_metab_temp <- readRDS(maaslin_path("MaAslinLM_metab_temp.rds"))

fdr_tbl <- dplyr::bind_rows(
  fdr_summary(MaAslin_metab_multi,  "Primary: Temp + WT + (1|Plot)"),
  fdr_summary(MaAslinLM_metab_WT,   "Sensitivity: WT only, fixed effects"),
  fdr_summary(MaAslinLM_metab_temp, "Sensitivity: Temp only, fixed effects"))
print(as.data.frame(fdr_tbl), digits = 3)

print(dplyr::bind_rows(
    class_level_competitive(MaAslinLM_metab_WT,   "WT_drawdown_cm", "WT_drawdown",
                            taxdf_met, "Superclass"),
    class_level_competitive(MaAslinLM_metab_temp, "Temp_avg_top",   "Temp_avg_top",
                            taxdf_met, "Superclass")) |> as.data.frame(), digits = 3)
```

## 2.10 POREWATER GAS MODELS AND SUPPLEMENTARY FIGURE S2

```{r porewater-gas-models-and-fig}

CH4.pw <- read.csv(p_raw_req("CH4-Surface_Porewater_2017-2021.csv"))[
  c("Year","Month","Plot","Depth","d13CH4","d13CO2","alpha","CH4mM","CO2mM","corrCO2CH4")]
pw2022 <- read.csv(p_raw_req("sprucePorewater2022.csv"))[
  c("year","month","site","depth","d13CH4","d13CO2","alpha","CH4mM","CO2mM","corrCO2CH4")]
colnames(pw2022) <- colnames(CH4.pw)
pw2022$Plot <- gsub("Plot(\\d+)", "P\\1", pw2022$Plot)

PWgas <- rbind(CH4.pw, pw2022) |>
  filter(grepl("P\\d+", Plot)) |>
  mutate(Plot = as.integer(gsub("P0*(\\d+)", "\\1", Plot)), Year = as.factor(Year),
         CH4mM = as.numeric(CH4mM), CO2mM = as.numeric(CO2mM)) |>
  filter(Year %in% c("2020","2021","2022"), Depth < 50, Depth > 10) |>
  group_by(Year, Plot, Depth) |>
  summarise(CH4mM = mean(CH4mM, na.rm = TRUE), CO2mM = mean(CO2mM, na.rm = TRUE),
            corrCO2CH4 = mean(corrCO2CH4, na.rm = TRUE), .groups = "drop") |>
  left_join(sample_df |> dplyr::select(-any_of("Depth")) |>
              mutate(Year = as.factor(Year), Plot = as.integer(as.character(Plot))),
            by = c("Year", "Plot")) |>
  mutate(CO2_CH4_ratio = CO2mM / CH4mM, Plot = as.factor(Plot))

write_out(PWgas, p_meta("PW_gas_metadata.csv"))

fit_pw <- function(response, label) {
  cat(label, "Shapiro P =",
      round(shapiro.test(na.omit(PWgas[[response]]))$p.value, 3), "\n")
  f <- function(rhs) as.formula(paste(response, "~", rhs))
  r <- lrt_interaction(f("Temp_avg_top * WT_drawdown_cm + (1|Plot)"),
                       f("Temp_avg_top + WT_drawdown_cm + (1|Plot)"),
                       PWgas, label = label)
  fixed <- drop_singular_re(r$model, PWgas, label)
  register_model(tidy_glmmtmb(fixed$model, label))
  list(model = fixed$model, dropped_re = fixed$dropped_re, lrt_p = r$lrt_p)
}
res_ch4   <- fit_pw("CH4mM",         "Porewater CH4 (mM)")
res_co2   <- fit_pw("CO2mM",         "Porewater CO2 (mM)")
res_ratio <- fit_pw("CO2_CH4_ratio", "Porewater CO2:CH4 ratio")

scatter_pw <- function(x, y, xlab, ylab, legend_pos) {
  ggplot(PWgas, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point(aes(fill = Warming_Treatment.ord, shape = Year), colour = "black", size = 8) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", linewidth = 1) +
    scale_fill_manual(values = allGroupsColors) + scale_shape_manual(values = c(21,22,24)) +
    guides(fill = guide_legend("Temperature\ntreatment", override.aes = list(shape = 21))) +
    xlab(xlab) + ylab(ylab) + theme_bw() +
    theme(legend.position = legend_pos, axis.text = element_text(size = 14),
          axis.title = element_text(size = 16))
}

FIG_S2C <- interactions::interact_plot(
    lm(CO2_CH4_ratio ~ Temp_avg_top * WT_drawdown_cm, data = PWgas),
    pred = Temp_avg_top, modx = WT_drawdown_cm, plot.points = TRUE, point.size = 8,
    y.label = "CO2:CH4 ratio", x.label = "Temperature (ºC, May-Aug)",
    legend.main = "Water table drawdown (cm)") +
  theme_bw() +
  theme(legend.position = "right", axis.text = element_text(size = 14),
        axis.title = element_text(size = 16))

save_fig(ggarrange(
    scatter_pw("Temp_avg_top",  "CO2mM", "Temperature (ºC, May-Aug)", "Pore water CO2 (mM)", "none"),
    scatter_pw("WT_drawdown_cm","CH4mM", "Water table drawdown (cm)", "Pore water CH4 (mM)", "right"),
    FIG_S2C, ncol = 3, labels = c("A","B","C"), font.label = list(size = 18)),
  p_fig("Figure_S2_porewater_gas.pdf"), width = 18, height = 6)
```

## 3. dbRDA COMMUNITY ORDINATIONS — Figure 1, Supplementary Figures S4 and S9

Condition on enclosure for repeated sampling, and on depth for the amplicons.
Terms tested sequentially by permutation, 999 permutations, `by = "terms"`.
Each permutation test runs once and the result is reused for the figure
annotation and for Supplementary Data 7.

```{r dbrda-run}

DBRDA_PERM <- 999

run_dbrda <- function(dis, sampledf, condition = "Condition(Plot)",
                      label, expected_terms, seed = 1234) {
  cond <- paste(condition, collapse = " + ")
  f <- function(rhs) as.formula(paste("dis ~", rhs, "+", cond))

  full     <- vegan::dbrda(f("Temp_avg_top + WT_drawdown_cm"), data = sampledf)
  vif_full <- vif.cca(full)
  cat("\n=== ", label, ": VIF, full model ===\n"); print(vif_full)

  set.seed(seed)
  a_temp <- anova(vegan::dbrda(f("Temp_avg_top"),   data = sampledf), permutations = DBRDA_PERM)
  set.seed(seed)
  a_wt   <- anova(vegan::dbrda(f("WT_drawdown_cm"), data = sampledf), permutations = DBRDA_PERM)
  set.seed(seed)
  a_int  <- anova(vegan::dbrda(f("Temp_avg_top * WT_drawdown_cm"), data = sampledf),
                  by = "terms", permutations = DBRDA_PERM)

  int_row <- grep(":", rownames(a_int))
  p_int   <- if (length(int_row)) a_int$`Pr(>F)`[int_row[1]] else NA_real_
  cat(label, ": Temp alone P =", a_temp$`Pr(>F)`[1],
      "| WT alone P =", a_wt$`Pr(>F)`[1],
      "| interaction P =", p_int, "\n")

  chosen <- if (!is.na(p_int) && p_int < 0.05) {
    "Temp_avg_top * WT_drawdown_cm"
  } else {
    keep <- c("Temp_avg_top", "WT_drawdown_cm")[
      c(a_temp$`Pr(>F)`[1] < 0.05, a_wt$`Pr(>F)`[1] < 0.05)]
    if (length(keep) == 0) "Temp_avg_top + WT_drawdown_cm" else paste(keep, collapse = " + ")
  }
  if (!identical(chosen, expected_terms))
    message(label, ": selection gives '", chosen, "', expected '", expected_terms,
            "'. The selected model is the one fitted and plotted.")

  final <- vegan::dbrda(f(chosen), data = sampledf)
  set.seed(seed)
  a_final <- anova(final, permutations = DBRDA_PERM)
  r2 <- RsquareAdj(final)

  vif_sel <- vif.cca(final)
  cat("=== ", label, ": VIF, selected model (", chosen, ") ===\n", sep = "")
  print(vif_sel)
  pred_in <- intersect(names(vif_sel), c("Temp_avg_top", "WT_drawdown_cm"))
  if (any(vif_sel[pred_in] > 10, na.rm = TRUE))
    warning(label, ": VIF > 10 in the selected model for ",
            paste(pred_in[vif_sel[pred_in] > 10], collapse = ", "))

  list(label = label, model = final, terms = chosen, selected = chosen,
       submitted_terms = expected_terms,
       vif = vif_sel, vif_full = vif_full,
       anova = a_final, adj_R2 = r2$adj.r.squared,
       n = nrow(sampledf), p_temp = a_temp$`Pr(>F)`[1], p_wt = a_wt$`Pr(>F)`[1],
       p_interaction = p_int,
       stat = paste0("R2adj=", round(r2$adj.r.squared, 3),
                     ", F=", round(a_final$F[1], 2),
                     ", ", if (a_final$`Pr(>F)`[1] < 0.001) "p<0.001"
                           else paste0("p=", round(a_final$`Pr(>F)`[1], 3)),
                     ", n=", nrow(sampledf)))
}

# axis2 is taken from the fitted model: a single-predictor model has one
# constrained axis, so the second axis is MDS1.
#
# ARROW_FRAC controls biplot arrow length. vegan returns biplot scores on the
# constrained-axis scale, which in some panels is much larger than the spread of
# the site scores, so an unscaled arrow runs off the plot. Every arrow in a panel
# is multiplied by one common factor chosen so the longest reaches ARROW_FRAC of
# that panel's site-score range. Relative arrow lengths and all angles are
# preserved within a panel; lengths are not comparable between panels, which is
# already true of biplot scores.
ARROW_FRAC <- 0.65

dbrda_panel <- function(fit, sampledf, fill_var, axis1 = "dbRDA1", axis2 = NULL,
                        arrow_labels, flip_x = FALSE, high, low,
                        title = NULL, arrow_frac = ARROW_FRAC) {
  site_axes <- colnames(scores(fit$model, display = "sites"))
  if (is.null(axis2))
    axis2 <- if (length(site_axes) > 1 && grepl("^dbRDA", site_axes[2])) site_axes[2] else "MDS1"
  smry <- summary(fit$model)
  prop <- smry$cont$importance["Proportion Explained", ]
  d1   <- cbind(data.frame(scores(fit$model, display = "sites")),
                Year = as.factor(sampledf$Year),
                Temp_avg_top = sampledf$Temp_avg_top,
                WT_drawdown_cm = sampledf$WT_drawdown_cm)
  ARROW_LABELS <- c(Temp_avg_top = "Temp", WT_drawdown_cm = "Water table\ndrawdown")
  d2 <- as.data.frame(scores(fit$model, display = "bp"))
  d2 <- d2[rownames(d2) %in% names(ARROW_LABELS), , drop = FALSE]
  if (!missing(arrow_labels) && length(arrow_labels) == nrow(d2)) {
    rownames(d2) <- arrow_labels
  } else {
    rownames(d2) <- unname(ARROW_LABELS[rownames(d2)])
  }

  # Rescale the arrows. Half-range on each axis, so an arrow drawn from the
  # origin at arrow_frac = 1 would just touch the edge of the site cloud.
  half_x <- max(abs(range(d1[[axis1]], na.rm = TRUE)))
  half_y <- max(abs(range(d1[[axis2]], na.rm = TRUE)))
  reach  <- max(sqrt((d2[[axis1]] / half_x)^2 + (d2[[axis2]] / half_y)^2), na.rm = TRUE)
  if (is.finite(reach) && reach > 0) {
    k <- arrow_frac / reach
    d2[[axis1]] <- d2[[axis1]] * k
    d2[[axis2]] <- d2[[axis2]] * k
  }

  s <- if (flip_x) -1 else 1

  ggplot(d1, aes(x = s * .data[[axis1]], y = .data[[axis2]])) +
    geom_point(aes(fill = .data[[fill_var]], shape = Year), color = "black", size = 6) +
    scale_shape_manual(values = c(21, 22, 24)) +
    scale_fill_gradient(high = high, low = low) +
    geom_hline(yintercept = 0, linetype = "dotted") +
    geom_vline(xintercept = 0, linetype = "dotted") +
    geom_segment(data = d2, aes(x = 0, xend = s * .data[[axis1]], y = 0, yend = .data[[axis2]]),
                 color = "grey20", arrow = arrow(length = unit(0.01, "npc")), linewidth = 2) +
    geom_text(data = d2, aes(x = s * .data[[axis1]], y = .data[[axis2]],
                             label = rownames(d2),
                             hjust = 0.5 * (1 - sign(s * .data[[axis1]])),
                             vjust = 0.5 * (1 - sign(.data[[axis2]]))),
              color = "grey20", size = 4.5, lineheight = 0.9) +
    annotate("text", x = Inf, y = -Inf, label = fit$stat,
             hjust = 1.05, vjust = -0.5, color = "grey40", size = 4) +
    # Extra x expansion so the "Water table drawdown" label, which is drawn
    # outward from the arrow tip, is not clipped at the panel edge.
    scale_x_continuous(expand = expansion(mult = 0.14)) +
    scale_y_continuous(expand = expansion(mult = 0.10)) +
    labs(title = title) +
    xlab(paste0(axis1, " (", round(prop[axis1] * 100, 1), "%)")) +
    ylab(paste0(axis2, " (", round(prop[axis2] * 100, 1), "%)")) +
    theme_bw() +
    theme(legend.position = "none", axis.text = element_text(size = 16),
          axis.title = element_text(size = 16),
          plot.title = element_text(size = 20, hjust = 0.5,
                                    margin = margin(b = 10)),
          plot.margin = margin(t = 12, r = 8, b = 6, l = 8))
}

# --- Datasets ------------------------------------------------------------------

peat16S.phy.RA <- readRDS(p_raw_req("peat16S_phy.rds")) |>
  transform_sample_counts(function(x) x / sum(x)) |>
  subset_samples(!is.na(Warming))
peatITS.phy.RA <- readRDS(p_raw_req("peatITS.phy.rds")) |>
  transform_sample_counts(function(x) x / sum(x)) |>
  subset_samples(!is.na(Warming))
metabolites.phy.RA <- transform_sample_counts(metabolites_phylo, function(x) x / sum(x))
droughtMAG.ps          <- readRDS(p_raw_req(MAG_ABUNDANCE_RDS))
droughtMAG.activity.ps <- readRDS(p_raw_req(MAG_ACTIVITY_RDS))

# Derive WT_drawdown_cm where it is missing and join the remaining model
# predictors from the canonical metadata on Year + Plot.
repair_mag_metadata <- function(ps, label) {
  sd_now  <- data.frame(sample_data(ps), check.names = FALSE)
  keep_rn <- rownames(sd_now)
  needed  <- c("WT_drawdown_cm", "AvgTemp10cm", "Temp_avg_top", "NOSC")
  before  <- setdiff(needed, names(sd_now))

  if (!"WT_drawdown_cm" %in% names(sd_now) && "WT_drawdown" %in% names(sd_now))
    sd_now$WT_drawdown_cm <- sd_now$WT_drawdown * 100

  still <- setdiff(needed, names(sd_now))
  if (length(still)) {
    if (!all(c("Year", "Plot") %in% names(sd_now)))
      stop(label, ": cannot join metadata, sample_data has no Year/Plot. Present: ",
           paste(sort(names(sd_now)), collapse = ", "))
    src <- sample_df |>
      dplyr::mutate(.y = as.character(Year), .p = as.character(Plot)) |>
      dplyr::select(.y, .p, dplyr::any_of(still)) |>
      dplyr::distinct(.y, .p, .keep_all = TRUE)
    sd_now$.y <- as.character(sd_now$Year)
    sd_now$.p <- as.character(sd_now$Plot)
    n_before  <- nrow(sd_now)
    sd_now    <- dplyr::left_join(sd_now, src, by = c(".y", ".p"))
    if (nrow(sd_now) != n_before)
      stop(label, ": metadata join duplicated rows (", n_before, " -> ", nrow(sd_now), ")")
    sd_now$.y <- NULL; sd_now$.p <- NULL
    rownames(sd_now) <- keep_rn
  }

  after <- setdiff(needed, names(sd_now))
  cat(sprintf("  %-10s missing before: %-32s after: %s\n", label,
              if (length(before)) paste(before, collapse = ", ") else "none",
              if (length(after))  paste(after,  collapse = ", ") else "none"))
  for (v in intersect(needed, names(sd_now))) {
    n_na <- sum(is.na(sd_now[[v]]))
    if (n_na) cat(sprintf("             %s has %d/%d NA\n", v, n_na, nrow(sd_now)))
  }

  sample_data(ps) <- sample_data(sd_now)
  ps
}

cat("\n=== MAG metadata repair ===\n")
droughtMAG.ps          <- repair_mag_metadata(droughtMAG.ps,          "abundance")
droughtMAG.activity.ps <- repair_mag_metadata(droughtMAG.activity.ps, "activity")

cat("\n=== MAG builds in use ===\n")
cat(" abundance:", MAG_ABUNDANCE_RDS, "|", nsamples(droughtMAG.ps),
    "samples,", ntaxa(droughtMAG.ps), "taxa\n")
cat(" activity :", MAG_ACTIVITY_RDS, "|", nsamples(droughtMAG.activity.ps),
    "samples,", ntaxa(droughtMAG.activity.ps), "taxa\n")

dis_16S <- phyloseq::distance(peat16S.phy.RA,     method = "bray")
dis_ITS <- phyloseq::distance(peatITS.phy.RA,     method = "bray")
dis_met <- phyloseq::distance(metabolites.phy.RA, method = "bray")
dis_abd <- phyloseq::distance(droughtMAG.ps,          method = "bray")
dis_act <- phyloseq::distance(droughtMAG.activity.ps, method = "bray")

sd_16S <- data.frame(sample_data(peat16S.phy.RA))
sd_ITS <- data.frame(sample_data(peatITS.phy.RA))
sd_met <- data.frame(sample_data(metabolites.phy.RA))
sd_abd <- data.frame(sample_data(droughtMAG.ps))
sd_act <- data.frame(sample_data(droughtMAG.activity.ps))

fit_16S <- run_dbrda(dis_16S, sd_16S, c("Condition(Depth)","Condition(Plot)"),
                     "16S amplicons",  "Temp_avg_top + WT_drawdown_cm")
fit_ITS <- run_dbrda(dis_ITS, sd_ITS, c("Condition(Depth)","Condition(Plot)"),
                     "ITS amplicons",  "Temp_avg_top + WT_drawdown_cm")
fit_met <- run_dbrda(dis_met, sd_met, "Condition(Plot)",
                     "Metabolites",    "WT_drawdown_cm")
fit_abd <- run_dbrda(dis_abd, sd_abd, "Condition(Plot)",
                     "MAG abundance",  "Temp_avg_top + WT_drawdown_cm")
fit_act <- run_dbrda(dis_act, sd_act, "Condition(Plot)",
                     "MAG activity",   "Temp_avg_top * WT_drawdown_cm")

DBRDA_FITS <- list(fit_act, fit_met, fit_16S, fit_ITS, fit_abd)
```

```{r dbrda-figures, fig.width=12, fig.height=13.5}

FIG1A <- dbrda_panel(fit_act, sd_act, "WT_drawdown_cm",
                     title = "Genome activity",
                     high = "#FF6633", low = "#56B4E9")
FIG1B <- dbrda_panel(fit_met, sd_met, "WT_drawdown_cm", flip_x = TRUE,
                     title = "Metabolite composition",
                     high = "#FF6633", low = "#56B4E9")
FIG1C <- dbrda_panel(fit_16S, sd_16S, "Temp_avg_top",
                     title = "Microbial community\n(16S rRNA gene)",
                     high = "#BD0026", low = "#fcffb4")
FIG1D <- dbrda_panel(fit_ITS, sd_ITS, "Temp_avg_top",
                     title = "Fungal community (ITS)",
                     high = "#BD0026", low = "#fcffb4")
FIG_S4 <- dbrda_panel(fit_abd, sd_abd, "Temp_avg_top",
                       title = "Genome abundance",
                       high = "#BD0026", low = "#fcffb4")

# Panel C carries a two-line title; pad the one-line titles so the four plot
# areas in Figure 1 still start at the same height.
PANEL_TITLE_LINES <- 2
pad_title <- function(p, n = PANEL_TITLE_LINES) {
  t <- p$labels$title
  if (is.null(t) || !nzchar(t)) return(p)
  extra <- n - lengths(regmatches(t, gregexpr("\n", t, fixed = TRUE)))[1] - 1
  if (extra > 0) p <- p + labs(title = paste0(t, strrep("\n", extra)))
  p
}

save_fig(FIG1A,   p_fig("Figure_1A_MAG_activity.pdf"),          width = 6, height = 6.2)
save_fig(FIG1B,   p_fig("Figure_1B_metabolites.pdf"),           width = 6, height = 6.7)
save_fig(FIG1C,   p_fig("Figure_1C_16S.pdf"),          width = 6, height = 6.9)
save_fig(FIG1D,   p_fig("Figure_1D_ITS.pdf"),          width = 6, height = 6.7)
save_fig(FIG_S4, p_fig("Figure_S4_MAG_abundance_dbRDA.pdf"),    width = 8, height = 6.7)
save_fig(ggarrange(pad_title(FIG1A), pad_title(FIG1B),
                   pad_title(FIG1C), pad_title(FIG1D),
                   labels = c("A","B","C","D"),
                   font.label = list(size = 25), ncol = 2, nrow = 2, legend = "none"),
         p_fig("Figure_1_dbRDA.pdf"), height = 13.5, width = 12)

# Supplementary Data 7: VIF for the selected model of each dataset, with the
# full two-predictor VIF alongside.
write_out(
  purrr::map_dfr(DBRDA_FITS, \(f) data.frame(
    model = f$label, vif_from = "selected model", terms = f$selected,
    as.list(f$vif))) |>
  dplyr::bind_rows(
    purrr::map_dfr(DBRDA_FITS, \(f) data.frame(
      model = f$label, vif_from = "full model",
      terms = "Temp_avg_top + WT_drawdown_cm",
      as.list(f$vif_full)))),
  p_sd("Suppl_Data_07_dbRDA_VIF.csv"))
print(purrr::map_dfr(DBRDA_FITS, \(f) data.frame(
    model = f$label, terms_fitted = f$terms, terms_selected = f$selected,
    adj_R2 = f$adj_R2, F_stat = f$anova$F[1], p_value = f$anova$`Pr(>F)`[1],
    p_temp_marginal = f$p_temp, p_WT_marginal = f$p_wt,
    p_interaction = f$p_interaction, n = f$n)), digits = 3, row.names = FALSE)
```

```{r dbrda-varpart, fig.width=16, fig.height=4}

varpart_one <- function(dis, sampledf, label) {
  vp <- varpart(dis, sampledf[, "Temp_avg_top", drop = FALSE],
                     sampledf[, "WT_drawdown_cm", drop = FALSE])
  cat("\n=== VarPart:", label, "===\n"); print(vp)
  vp
}
vp_MAGactivity <- varpart_one(dis_act, sd_act, "MAG activity")
vp_metabolites <- varpart_one(dis_met, sd_met, "Metabolites")
vp_16S         <- varpart_one(dis_16S, sd_16S, "16S amplicons")

# Partial permutation tests. Each predictor is tested with the other one moved
# into the conditioning term, so the test matches the unique fraction reported by
# varpart. The marginal tests in run_dbrda() test each predictor on its own and
# do not.
partial_dbrda <- function(dis, sampledf, condition, label, seed = 1234) {
  cond <- paste(c(condition), collapse = " + ")
  one <- function(pred, other) {
    f <- as.formula(paste("dis ~", pred, "+ Condition(", other, ") +", cond))
    set.seed(seed)
    a <- anova(vegan::dbrda(f, data = sampledf), permutations = DBRDA_PERM)
    r <- RsquareAdj(vegan::dbrda(f, data = sampledf))$adj.r.squared
    data.frame(model = label, predictor = pred,
               conditioned_on = paste(other, cond, sep = " + "),
               partial_adj_R2 = r, F_stat = a$F[1], p_partial = a$`Pr(>F)`[1],
               n = nrow(sampledf), stringsAsFactors = FALSE)
  }
  dplyr::bind_rows(one("Temp_avg_top",   "WT_drawdown_cm"),
                   one("WT_drawdown_cm", "Temp_avg_top"))
}

PARTIAL_TESTS <- dplyr::bind_rows(
  partial_dbrda(dis_act, sd_act, "Condition(Plot)",                          "MAG activity"),
  partial_dbrda(dis_met, sd_met, "Condition(Plot)",                          "Metabolites"),
  partial_dbrda(dis_16S, sd_16S, c("Condition(Depth)", "Condition(Plot)"),   "16S amplicons"),
  partial_dbrda(dis_ITS, sd_ITS, c("Condition(Depth)", "Condition(Plot)"),   "ITS amplicons"),
  partial_dbrda(dis_abd, sd_abd, "Condition(Plot)",                          "MAG abundance"))

cat("\n=== Partial dbRDA tests (each predictor conditioned on the other) ===\n")
print(as.data.frame(PARTIAL_TESTS), digits = 3, row.names = FALSE)
# Not written on its own. These values are published beside the marginal tests in
# Table_dbRDA_marginal_vs_partial.csv, which is the table the Results quote.

# Marginal and partial P values side by side.
print(
  purrr::map_dfr(DBRDA_FITS, \(f) data.frame(
      model = f$label, predictor = c("Temp_avg_top", "WT_drawdown_cm"),
      p_marginal = c(f$p_temp, f$p_wt))) |>
    dplyr::left_join(dplyr::select(PARTIAL_TESTS, model, predictor,
                                   partial_adj_R2, p_partial),
                     by = c("model", "predictor")), digits = 3, row.names = FALSE)

pdf(p_fig("Figure_S9_variation_partitioning.pdf"), width = 16, height = 4)
par(mfrow = c(1, 3))
plot(vp_MAGactivity, digits = 2, Xnames = c("Temp","Water\ntable"),
     main = "MAG Activity",  bg = c("tomato","steelblue"))
plot(vp_metabolites, digits = 2, Xnames = c("Temp","Water\ntable"),
     main = "Metabolites",   bg = c("tomato","steelblue"))
plot(vp_16S,         digits = 2, Xnames = c("Temp","Water\ntable"),
     main = "16S Amplicons", bg = c("tomato","steelblue"))
dev.off()
FILE_LOG <- c(FILE_LOG, p_fig("Figure_S9_variation_partitioning.pdf"))

# Fractions are derived from vp$part$fract, whose row names state the model each
# one came from: X1 is temperature, X2 is water table.
#
#   unique_temp = R2(X1+X2) - R2(X2 alone)
#   unique_WT   = R2(X1+X2) - R2(X1 alone)
#   shared      = R2(X1 alone) + R2(X2 alone) - R2(X1+X2)
#   unexplained = 1 - R2(X1+X2)

varpart_fractions <- function(vp, nm) {
  fr  <- vp$part$fract
  col <- grep("^Adj", colnames(fr), value = TRUE)[1]
  rn  <- rownames(fr)
  has1 <- grepl("X1", rn); has2 <- grepl("X2", rn)
  i_ab  <- which(has1 & !has2)   # temperature alone
  i_bc  <- which(has2 & !has1)   # water table alone
  i_abc <- which(has1 &  has2)   # both
  stopifnot(length(i_ab) == 1, length(i_bc) == 1, length(i_abc) == 1)

  ab  <- fr[i_ab,  col]
  bc  <- fr[i_bc,  col]
  abc <- fr[i_abc, col]

  out <- data.frame(model = nm,
                    temp_alone   = ab,
                    WT_alone     = bc,
                    both         = abc,
                    unique_temp  = abc - bc,
                    unique_WT    = abc - ab,
                    shared       = ab + bc - abc,
                    unexplained  = 1 - abc,
                    stringsAsFactors = FALSE)
  stopifnot(isTRUE(all.equal(out$unique_temp + out$unique_WT + out$shared, abc,
                             tolerance = 1e-8)))
  out
}

VP_LIST <- list("MAG Activity" = vp_MAGactivity, "Metabolites" = vp_metabolites,
                "16S Amplicons" = vp_16S)

cat("\n=== varpart fract tables, as labelled by vegan ===\n")
for (nm in names(VP_LIST)) {
  cat("\n--", nm, "--\n"); print(VP_LIST[[nm]]$part$fract)
  cat("   indfract rows, in file order:\n"); print(VP_LIST[[nm]]$part$indfract)
}

VARPART <- purrr::imap_dfr(VP_LIST, varpart_fractions)

# The partial test P value belongs with the unique fraction it corresponds to.
pt <- PARTIAL_TESTS |>
  dplyr::mutate(.m = tolower(model)) |>
  dplyr::select(.m, predictor, p_partial) |>
  tidyr::pivot_wider(names_from = predictor, values_from = p_partial,
                     names_prefix = "p_unique_")
VARPART <- VARPART |>
  dplyr::mutate(.m = tolower(model)) |>
  dplyr::left_join(pt, by = ".m") |>
  dplyr::select(-.m) |>
  dplyr::rename(p_unique_temp = p_unique_Temp_avg_top,
                p_unique_WT   = p_unique_WT_drawdown_cm) |>
  dplyr::relocate(p_unique_temp, .after = unique_temp) |>
  dplyr::relocate(p_unique_WT,   .after = unique_WT)
if (any(is.na(VARPART$p_unique_temp) | is.na(VARPART$p_unique_WT)))
  stop("Partial test P values did not join onto every varpart row. ",
       "Check that VP_LIST names match PARTIAL_TESTS$model.")

cat("\n=== Variation partitioning ===\n"); print(VARPART, row.names = FALSE)

# The raw vegan fractions behind the summary above and Supplementary Figure S9.
print(purrr::imap_dfr(VP_LIST, \(vp, nm) {
    ind <- vp$part$indfract
    col <- grep("^Adj", colnames(ind), value = TRUE)[1]
    data.frame(model = nm, fraction = rownames(ind),
               adj_R2 = suppressWarnings(as.numeric(ind[[col]])),
               stringsAsFactors = FALSE)}), row.names = FALSE)
```

## 4. MAG-LEVEL MAASLIN2 — Figure 2

```{r maaslin-MAG-level}

# psmelt drops the CAMPER columns when the tax table is wide. Put them back.
get_camper_annot <- function(ps) {
  tt <- data.frame(tax_table(ps))
  dplyr::select(tt, Gene, all_of(grep("^CAMPER", colnames(tt), value = TRUE)))
}
psmelt_with_camper <- function(ps) psmelt(ps) |> left_join(get_camper_annot(ps), by = "Gene")

MAGactivity.ps <- prune_taxa(taxa_sums(droughtMAG.activity.ps) > 0, droughtMAG.activity.ps)
taxdf_MAG      <- data.frame(tax_table(droughtMAG.ps))

# Supplementary Data 6: taxonomy, genome quality and phenol-active gene counts
# for all MAGs. The counts are the numerator behind the statement in the Results
# that every MAG encodes at least one phenol-active enzyme, so they travel with
# the table a reader would check it against.
suppl_data_06 <- taxdf_MAG |>
  dplyr::left_join(read_csv(p_meta("df_CAMPER.csv"), show_col_types = FALSE) |>
                     dplyr::select(SPRUCE_ID, anoxic, oxic,
                                   Specialist_phenol_decomposition),
                   by = "SPRUCE_ID") |>
  dplyr::mutate(anoxic = tidyr::replace_na(anoxic, 0L),
                oxic   = tidyr::replace_na(oxic,   0L),
                phenol_active_genes = anoxic + oxic,
                Specialist_phenol_decomposition =
                  tidyr::replace_na(Specialist_phenol_decomposition, "None"))

cat(sprintf("Supplementary Data 6: %d MAGs, %d with at least one phenol-active gene (%.1f%%)\n",
            nrow(suppl_data_06), sum(suppl_data_06$phenol_active_genes > 0),
            100 * mean(suppl_data_06$phenol_active_genes > 0)))
write_out(suppl_data_06, p_sd("Suppl_Data_06_MAG_taxonomy_quality.csv"))

# --- MAG-level Maaslin2 results -----------------------------------------------

if (REFIT_MAASLIN) {
  fit_maaslin(MAGactivity.ps, "MaAslinLM_ActivityMAG",      "WT_drawdown_cm", NULL)
  fit_maaslin(MAGactivity.ps, "MaAslinLM_MAGactivity_temp", "AvgTemp10cm",    NULL)
  fit_maaslin(droughtMAG.ps,  "MaAslinLM_MAG_temp",         "AvgTemp10cm",    NULL)
  fit_maaslin(MAGactivity.ps, "MaAslinLM_ActivityMAG_NOSC", "NOSC",           NULL)
}
df_CoefWT           <- read_maaslin(maaslin_path("MaAslinLM_ActivityMAG.rds"),      "SPRUCE_ID", TRUE)
df_CoefTempActivity <- read_maaslin(maaslin_path("MaAslinLM_MAGactivity_temp.rds"), "SPRUCE_ID", TRUE)
df_CoefTemp         <- read_maaslin(maaslin_path("MaAslinLM_MAG_temp.rds"),         "SPRUCE_ID", TRUE)
df_CoefNOSC         <- read_maaslin(maaslin_path("MaAslinLM_ActivityMAG_NOSC.rds"), "SPRUCE_ID", TRUE)


cat("\nMAG abundance ~ Temp: ",
    sum(df_CoefTemp$pval < 0.05), " at P<0.05 | ",
    sum(df_CoefTemp$qval < 0.05), " at q<0.05 | ",
    sum(df_CoefTemp$qval < 0.10), " at q<0.10 | ",
    sum(df_CoefTemp$qval < 0.25), " at q<0.25 | min q = ",
    signif(min(df_CoefTemp$qval, na.rm = TRUE), 3), "\n", sep = "")

# --- Supplementary Data 8: clustered MAG responses ----------------------------

Clustered_df <- read_csv(p_raw_req(CLUSTER_CSV), show_col_types = FALSE)

for (grp in list(list(ids = WSOX_IDS, cl = "Warming-stimulated oxic"),
                 list(ids = WSAN_IDS, cl = "Warming-stimulated anoxic"),
                 list(ids = WI_IDS,   cl = "Warming-inhibited"))) {
  from_csv <- Clustered_df$SPRUCE_ID[Clustered_df$Cluster == grp$cl]
  if (!setequal(grp$ids, from_csv))
    stop(grp$cl, " in ", CLUSTER_CSV, " no longer matches the hardcoded vector. ",
         "Only in file: ", paste(setdiff(from_csv, grp$ids), collapse = ", "),
         " | only in code: ", paste(setdiff(grp$ids, from_csv), collapse = ", "))
}

Suppl_Data_08 <- taxdf_MAG |>
  filter(SPRUCE_ID %in% Clustered_df$SPRUCE_ID) |>
  left_join(dplyr::select(Clustered_df, SPRUCE_ID, Cluster), by = "SPRUCE_ID") |>
  left_join(dplyr::select(df_CoefTemp, SPRUCE_ID, coef_Temp = coef, stderr_Temp = stderr,
                          CI_lower_Temp = CI_lower, CI_upper_Temp = CI_upper,
                          pval_Temp = pval, qval_Temp = qval, evidence_Temp = evidence),
            by = "SPRUCE_ID") |>
  left_join(dplyr::select(df_CoefWT, SPRUCE_ID, coef_WT = coef, stderr_WT = stderr,
                          CI_lower_WT = CI_lower, CI_upper_WT = CI_upper,
                          pval_WT = pval, qval_WT = qval, evidence_WT = evidence),
            by = "SPRUCE_ID") |>
  left_join(dplyr::select(df_CoefNOSC, SPRUCE_ID, coef_NOSC = coef, stderr_NOSC = stderr,
                          pval_NOSC = pval, qval_NOSC = qval, evidence_NOSC = evidence),
            by = "SPRUCE_ID") |>
  dplyr::select(Drought_ID, Domain, Phylum, Class, Order, Family, Genus, Species,
                SPRUCE_ID, Cluster,
                starts_with("coef_Temp"), starts_with("stderr_Temp"),
                starts_with("CI_lower_Temp"), starts_with("CI_upper_Temp"),
                pval_Temp, qval_Temp, evidence_Temp,
                coef_WT, stderr_WT, CI_lower_WT, CI_upper_WT, pval_WT, qval_WT, evidence_WT,
                coef_NOSC, stderr_NOSC, pval_NOSC, qval_NOSC, evidence_NOSC) |>
  arrange(Cluster, SPRUCE_ID)
write_out(Suppl_Data_08, p_sd("Suppl_Data_08_MAG_responses.csv"))

# Figure 3 node set: the P<0.05 subset of the fixed-effects abundance model.
Fig3_node_set <- filter(df_CoefTemp, pval <= 0.05)$SPRUCE_ID
cat("Figure 3 node set:", length(Fig3_node_set), "warming-responsive MAGs\n")
if (!setequal(Fig3_node_set, WARMING_RESPONSIVE_IDS))
  warning("Figure 3 node set no longer equals the three warming clusters. ",
          "Symmetric difference: ",
          paste(union(setdiff(Fig3_node_set, WARMING_RESPONSIVE_IDS),
                      setdiff(WARMING_RESPONSIVE_IDS, Fig3_node_set)), collapse = ", "))
```

```{r fig2, fig.width=15, fig.height=15}

# --- Figure 2A: genome response tree -------------------------------------------------

Taxa_Phenol_active_annotations <- read_csv(p_raw_req("Taxa_Phenol_active_annotations.csv"), show_col_types = FALSE)
colnames(Taxa_Phenol_active_annotations)[1:3] <- c("Drought_ID","anoxic_potential","oxic_potential")

tax_INFO <- taxdf_MAG |>
  left_join(dplyr::select(Clustered_df, SPRUCE_ID, Cluster), by = "SPRUCE_ID") |>
  left_join(read_csv(p_meta("df_CAMPER.csv"), show_col_types = FALSE), by = "SPRUCE_ID") |>
  left_join(dplyr::select(df_CoefWT,   SPRUCE_ID, coef_waterdrawdown = coef), by = "SPRUCE_ID") |>
  left_join(dplyr::select(df_CoefTemp, SPRUCE_ID, coef_temp = coef),          by = "SPRUCE_ID") |>
  left_join(Taxa_Phenol_active_annotations, by = "Drought_ID") |>
  mutate(Circle = 1,
         Cluster = as.factor(replace_na(Cluster, "None")),
         Talent_specialist = ifelse(Talent_Status == "none", "none", Specialist))
rownames(tax_INFO) <- tax_INFO$SPRUCE_ID

required_cols <- c("Circle","Phylum","Talent_specialist","coef_waterdrawdown","coef_temp","anoxic","oxic")
if (length(setdiff(required_cols, colnames(tax_INFO))))
  stop("Missing columns in tax_INFO: ",
       paste(setdiff(required_cols, colnames(tax_INFO)), collapse = ", "))

# Node 509 is the root of this tree file.
phy_tree(droughtMAG.ps) <- ape::root(phy_tree(droughtMAG.ps), node = 509)
tre_full <- phy_tree(droughtMAG.ps)

orderMAGs   <- unique(get_taxa_name(ggtree(tre_full)))
orderPhylum <- unique(taxdf_MAG[match(orderMAGs, taxdf_MAG$SPRUCE_ID), "Phylum"])
tax_INFO$Phylum <- factor(tax_INFO$Phylum, levels = rev(orderPhylum))
write_out(data.frame(Phylum = levels(tax_INFO$Phylum),
                     color  = unname(pals::glasbey())[seq_along(levels(tax_INFO$Phylum))]),
          p_meta("phylum_color_df.csv"))

tre_full_w_INFO <- full_join(as_tibble(tre_full), tax_INFO, by = c("label" = "SPRUCE_ID")) |>
  tidytree::as.treedata()

p3 <- ggtree(tre_full_w_INFO, layout = "fan", open.angle = 8, size = 0.1) +
  geom_tiplab(aes(label = ifelse(Cluster != "None", label, NA), color = Cluster),
              size = 3, offset = 0.1, align = TRUE) +
  scale_color_manual(values = cluster_colors) +
  theme(legend.position = "none") +
  geom_fruit(geom = geom_bar, mapping = aes(y = SPRUCE_ID, x = Circle, fill = Phylum),
             size = 1, pwidth = 0.1, offset = 0.55, stat = "identity", orientation = "y",
             axis.param = list(axis = "none")) +
  scale_fill_manual(values = unname(pals::glasbey())) +
  ggnewscale::new_scale_fill() +
  geom_fruit(geom = geom_star, mapping = aes(y = SPRUCE_ID, fill = Talent_specialist),
             size = 3, stroke = NA, starshape = 1, starstroke = 0, inherit.aes = FALSE) +
  scale_fill_manual(values = specialist_colors) +
  ggnewscale::new_scale_fill()

FIG_2A <- p3 +
  geom_fruit(geom = geom_bar, mapping = aes(y = SPRUCE_ID, x = Circle, fill = coef_waterdrawdown),
             size = 0.05, pwidth = 0.1, stat = "identity", orientation = "y",
             axis.param = list(axis = "none")) +
  scale_fill_gradient2(low = "forestgreen", mid = "white", high = "darkgoldenrod4",
                       midpoint = 0, na.value = "grey90") +
  labs(fill = "Change in relative activity\nwith water table drawdown") +
  ggnewscale::new_scale_fill() +
  geom_fruit(geom = geom_bar, mapping = aes(y = SPRUCE_ID, x = Circle, fill = coef_temp),
             size = 0.05, pwidth = 0.1, stat = "identity", orientation = "y",
             axis.param = list(axis = "none")) +
  scale_fill_gradient2(low = "#9eb6d0", mid = "white", high = "#e0301e",
                       midpoint = 0, na.value = "grey90") +
  labs(fill = "Change in relative abundance\nwith warming")
save_fig(FIG_2A, p_fig("Figure_2A_genome_response_tree.pdf"), height = 15, width = 15)

# --- Figure 2B: porewater CO2 vs genome group ---------------------------------

# PWgas has one row per enclosure-year-depth. Collapse to one value per
# enclosure-year before joining to a genome table.
dup_depths <- PWgas |> count(Year, Plot) |> filter(n > 1) |> nrow()
if (dup_depths > 0)
  message(dup_depths, " enclosure-years have porewater measured at more than one ",
          "depth; averaging across depth before joining.")
PWgas_sub <- PWgas |>
  group_by(Year, Plot) |>
  dplyr::summarise(across(c(CH4mM, CO2mM), \(x) mean(x, na.rm = TRUE)), .groups = "drop") |>
  mutate(Year = as.factor(Year), Plot = as.factor(Plot))

# Figure 2B is relative abundance. The axis label is built from the same
# constant as the data.
FIG2B_MEASURE <- "abundance"
fig2b_ps      <- droughtMAG.ps
FIG2B_AXIS    <- paste("Total relative", FIG2B_MEASURE, "of")

summarise_genome_group <- function(ids, ps = fig2b_ps, label = "genome group") {
  keep <- prune_taxa(taxa_names(ps) %in% ids, ps)
  if (!setequal(taxa_names(keep), ids))
    stop(label, ": genomes missing from the object: ",
         paste(setdiff(ids, taxa_names(keep)), collapse = ", "))
  out <- psmelt(keep) |>
    dplyr::select(Abundance, SPRUCE_ID, Warming, Year, CO2_Treatment, AvgTemp10cm,
                  Temp_avg_top, WT_drawdown, Warming_Treatment.ord, Plot) |>
    mutate(WT_drawdown_cm = WT_drawdown * 100, Year = as.factor(Year), Plot = as.factor(Plot)) |>
    left_join(PWgas_sub, by = c("Year", "Plot")) |>
    filter(!is.na(CO2mM)) |>
    group_by(Warming, Year, CO2_Treatment, AvgTemp10cm, Temp_avg_top,
             WT_drawdown_cm, Warming_Treatment.ord, Plot, CO2mM) |>
    dplyr::summarize(sum_Abundance = sum(Abundance, na.rm = TRUE),
                     n_genomes = dplyr::n(), .groups = "drop")
  if (!all(out$n_genomes == length(ids)))
    stop(label, ": incomplete genome sums, n_genomes ranges ",
         paste(range(out$n_genomes), collapse = "-"), " but should be ", length(ids))
  out
}

df_wsox <- summarise_genome_group(WSOX_IDS, label = "WS-OX")
df_wsan <- summarise_genome_group(WSAN_IDS, label = "WS-AN")
df_wi   <- summarise_genome_group(WI_IDS,   label = "WI")

# Holm across the three groups.
co2_group_models <- purrr::imap_dfr(
  list("WS-OX" = df_wsox, "WS-AN" = df_wsan, "WI" = df_wi),
  \(d, nm) {
    m <- lm(CO2mM ~ sum_Abundance, data = d)
    s <- summary(m)
    data.frame(measure = FIG2B_MEASURE, group = nm, n = nobs(m),
               slope = coef(m)[2], adj_R2 = s$adj.r.squared,
               p_value = s$coefficients[2, 4],
               x_min = min(d$sum_Abundance), x_max = max(d$sum_Abundance))
  }) |>
  mutate(p_holm = p.adjust(p_value, method = "holm"))

cat("\n=== Porewater CO2 vs genome-group total relative abundance ===\n")
print(as.data.frame(co2_group_models), digits = 4, row.names = FALSE)
if (!any(co2_group_models$p_holm < 0.05))
  cat("NOTE: no group survives Holm correction across the three tests.\n")

scatter_co2 <- function(df, xlab, legend_pos = "none") {
  ggplot(df, aes(x = sum_Abundance, y = CO2mM)) +
    geom_point(aes(fill = Warming_Treatment.ord, shape = Year), colour = "black", size = 8) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", linewidth = 1) +
    scale_fill_manual(values = allGroupsColors) + scale_shape_manual(values = c(21, 22, 24)) +
    guides(fill = guide_legend("Temperature\ntreatment", override.aes = list(shape = 21))) +
    xlab(xlab) + ylab("Pore water CO2 (mM)") + theme_bw() +
    theme(legend.position = legend_pos, axis.text = element_text(size = 16),
          axis.title = element_text(size = 18))
}

fig2b_C <- scatter_co2(df_wi, paste(FIG2B_AXIS, "WI"), "right")
save_legend(fig2b_C, legend_path(p_fig("Figure_2B_CO2_genome_scatter.pdf")),
            width = 3, height = 5)
save_fig(ggarrange(
    scatter_co2(df_wsan, paste(FIG2B_AXIS, "WS-AN")),
    scatter_co2(df_wsox, paste(FIG2B_AXIS, "WS-OX")),
    fig2b_C + theme(legend.position = "none"),
    ncol = 3, labels = c("A","B","C"), font.label = list(size = 18)),
  p_fig("Figure_2B_CO2_genome_scatter.pdf"), width = 15, height = 5)

# Sensitivity: the WS-OX relationship on the two core genomes only.
cat("WS-OX core (SPRUCE 397, 2183) adj R2 =",
    round(summary(lm(CO2mM ~ sum_Abundance,
      data = summarise_genome_group(c("SPRUCE 397","SPRUCE 2183"),
                                    fig2b_ps, "core WS-OX")))$adj.r.squared, 3), "\n")

# --- Figure 2C: phylogenetic tree ---------------------------------------------

FIG_2C <- p3 +
  geom_fruit(geom = geom_bar, mapping = aes(y = SPRUCE_ID, x = anoxic), fill = "#7CB9E8",
             stat = "identity", orientation = "y",
             axis.params = list(axis = "x", text.angle = -90, text.size = 1.5,
                                line.size = 0, vjust = 0.5)) +
  geom_fruit(geom = geom_bar, mapping = aes(y = SPRUCE_ID, x = oxic), fill = "#F88379",
             stat = "identity", orientation = "y",
             axis.params = list(axis = "x", text.angle = -90, text.size = 1.5,
                                line.size = 0, vjust = 0.5))
save_fig(FIG_2C, p_fig("Figure_2C_phylogenetic_tree.pdf"), height = 10, width = 10)
```

## 5. CO-OCCURRENCE NETWORK (SpiecEasi) — Figure 3

`RUN_NETWORK <- TRUE` fits SpiecEasi and writes `se_FULL.RDS` and
`assoMat_FULL.Rds` into `08_network/`. `FALSE` reloads those two files and
rebuilds the graph, the linking tables and the figure from them.

Everything this section writes goes to `08_network/`, except Supplementary
Data 10, which also goes to `04_suppl_data/`.

```{r network-build}

# Refit SpiecEasi RESAMPLE_N times on random 80% subsets of the metabolite
# features and record, for each edge of the displayed network, the fraction of
# replicates in which it is recovered. This tests sensitivity to the retained
# metabolite set; it is not a bootstrap over samples.
resample_edge_recovery <- function(target_edges, mag_mat, met_mat, n_rep) {
  recovered <- replicate(n_rep, {
    keep_cols <- sample(seq_len(ncol(met_mat)), round(0.8 * ncol(met_mat)))
    se_r <- multi.spiec.easi(list(mag_mat, met_mat[, keep_cols, drop = FALSE]),
      method = "mb", lambda.min.ratio = 1e-3, nlambda = 20,
      pulsar.params = list(rep.num = 20, ncores = 1))
    m <- as.matrix(symBeta(getOptBeta(se_r), mode = "ave"))
    dimnames(m) <- list(c(colnames(mag_mat), colnames(met_mat)[keep_cols]),
                        c(colnames(mag_mat), colnames(met_mat)[keep_cols]))
    gg <- graph_from_adjacency_matrix(m, mode = "undirected",
                                      weighted = TRUE, diag = FALSE)
    e  <- apply(ends(gg, E(gg)), 1, \(x) paste(sort(x), collapse = "|"))
    target_edges %in% e
  })
  data.frame(edge = target_edges, recovery_frequency = rowMeans(recovered))
}

PS_helper     <- function(x) (x - mean(x)) / sqrt(sd(x, na.rm = TRUE))
pareto_scale  <- function(x) data.frame(apply(x, 2, PS_helper))
log_helper    <- function(x, min.val) log2((x + sqrt(x^2 + min.val^2)) / 2)
log_transform <- function(x) {
  x_nz    <- x[, which(apply(x, 2, sum) != 0)]
  min.val <- min(abs(x_nz[x_nz != 0])) / 10
  data.frame(apply(x_nz, 2, log_helper, min.val))
}
get_sample_matrix <- function(ps) {
  otu <- as(otu_table(ps), "matrix")
  if (taxa_are_rows(ps)) otu <- t(otu)
  otu
}

# ---- Metabolites -------------------------------------------------------------

# Same reader as sections 2.8 and 2.9, so the network is built on exactly the
# metabolite set the supplementary tables describe: unnamed sentinel filtered,
# MSI levels on the four-level scheme.
metab_net_ps <- read_metab_ps(quiet = TRUE)
met_clean <- data.frame(otu_table(metab_net_ps, taxa_are_rows = TRUE)) |>
  log_transform() |> pareto_scale()
metabolites_clean <- metab_net_ps
metabolites_clean@otu_table <- otu_table(as.matrix(met_clean), taxa_are_rows = TRUE)

# Above the 30th percentile of taxa_sums, then restricted to features with an
# annotation confidence and a name.
threshold <- quantile(taxa_sums(metabolites_clean), ABUND_QUANTILE)
metabolites_phylo_sub <- metabolites_clean |>
  prune_taxa(taxa_sums(metabolites_clean) > threshold, x = _) |>
  subset_taxa(!is.na(Annotation_Confidence)) |>
  subset_taxa(!is.na(Final_name))
cat(sprintf("  Metabolites: %d (from %d)\n",
            ntaxa(metabolites_phylo_sub), ntaxa(metab_net_ps)))
# The "M" sentinel is removed upstream in read_metab_ps(), so !is.na(Final_name)
# already excludes it. Assert rather than filter again, so a regression shows up
# here instead of silently changing the network input.
stopifnot(!any(as.data.frame(tax_table(metabolites_phylo_sub))$Final_name %in% "M"))
saveRDS(metabolites_phylo_sub, p_net("metabolites_phylo_Filtered.RDS"))

# ---- MAGs --------------------------------------------------------------------

MAGactivity.net.ps <- readRDS(p_raw_req(MAG_ACTIVITY_RDS))
sampledf_net <- data.frame(sample_data(MAGactivity.net.ps))
sample_names(MAGactivity.net.ps) <- paste0("P", sampledf_net$Plot, "_", sampledf_net$Year)

# Methanogens by KEGG module presence.
methanogen_modules <- c("Methanogen_CO2_M00567", "Methanogen_Acetate_M00357",
                        "Methanogen_Methanol_M00356", "Methanogen_Methylamine_M00563")
mods    <- read_csv(p_raw_req(MODULE_CSV), show_col_types = FALSE)
present <- intersect(methanogen_modules, names(mods))
methanogens <- mods$SPRUCE_ID[rowSums(!is.na(mods[, present, drop = FALSE]) &
                                      mods[, present, drop = FALSE] != "") > 0]
stopifnot(setequal(methanogens, METHANOGEN_IDS))

targets <- Clustered_df$SPRUCE_ID[Clustered_df$Cluster %in%
  c("Warming-stimulated anoxic", "Warming-stimulated oxic", "Warming-inhibited")]
ps_MAG_sub <- subset_taxa(MAGactivity.net.ps, SPRUCE_ID %in% c(targets, methanogens))
cat(sprintf("  MAGs: %d (%d warming-responsive + %d methanogens)\n",
            ntaxa(ps_MAG_sub), length(targets), length(methanogens)))
saveRDS(ps_MAG_sub, p_net("ps_MAG_FilteredSensitive.RDS"))

# ---- Align -------------------------------------------------------------------

mag_mat <- get_sample_matrix(ps_MAG_sub)
met_mat <- get_sample_matrix(metabolites_phylo_sub)
shared  <- sort(intersect(rownames(mag_mat), rownames(met_mat)))
mag_mat <- mag_mat[shared, , drop = FALSE]
met_mat <- met_mat[shared, , drop = FALSE]
stopifnot(identical(rownames(mag_mat), rownames(met_mat)))
cat(sprintf("  Samples: %d | MAGs: %d | Metabolites: %d\n",
            nrow(mag_mat), ncol(mag_mat), ncol(met_mat)))

# ---- SpiecEasi ---------------------------------------------------------------

if (RUN_NETWORK) {
  se_FULL <- multi.spiec.easi(list(mag_mat, met_mat), method = "mb",
    lambda.min.ratio = 1e-3, nlambda = 20,
    pulsar.params = list(rep.num = 20, ncores = 1))
  assoMat_FULL <- as.matrix(symBeta(getOptBeta(se_FULL), mode = "ave"))
  dimnames(assoMat_FULL) <- list(c(colnames(mag_mat), colnames(met_mat)),
                                 c(colnames(mag_mat), colnames(met_mat)))
  diag(assoMat_FULL) <- 1
  saveRDS(se_FULL,      p_net("se_FULL.RDS"))
  saveRDS(assoMat_FULL, p_net("assoMat_FULL.Rds"))
} else {
  se_FULL      <- readRDS(net_path("se_FULL.RDS"))
  assoMat_FULL <- readRDS(net_path("assoMat_FULL.Rds"))
}

# StARS stability is a whole-network quantity, not a per-edge quantity.
cat(sprintf("  StARS stability: %.4f | edges: %d\n",
            getStability(se_FULL), sum(getRefit(se_FULL)) / 2))
```

```{r network-graph, fig.width=14, fig.height=14}

all_MAGs <- c(targets, methanogens)
g_full <- graph_from_adjacency_matrix(assoMat_FULL, mode = "undirected",
                                      weighted = TRUE, diag = FALSE)
g_full <- set_edge_attr(g_full, "color",
  value = ifelse(edge_attr(g_full, "weight") > 0, "forestgreen", "#D2042D"))

ee   <- ends(g_full, E(g_full))
keep_e <- which(degree(g_full, ee[,1]) > DEGREE_THRESHOLD |
                degree(g_full, ee[,2]) > DEGREE_THRESHOLD)
g_sub <- subgraph_from_edges(g_full, keep_e)
ee2   <- ends(g_sub, E(g_sub))
g_sub2 <- induced_subgraph(g_sub,
  unique(c(ee2[which(ee2[,1] %in% all_MAGs | ee2[,2] %in% all_MAGs), ])))
cat(sprintf("  After degree filter (>%d): %d edges\n", DEGREE_THRESHOLD, ecount(g_sub)))

G1 <- Clustered_df$SPRUCE_ID[Clustered_df$Cluster == "Warming-stimulated anoxic"]
G2 <- Clustered_df$SPRUCE_ID[Clustered_df$Cluster == "Warming-stimulated oxic"]
G3 <- Clustered_df$SPRUCE_ID[Clustered_df$Cluster == "Warming-inhibited"]

get_edge_weight <- function(a, b, m)
  if (a %in% rownames(m) && b %in% colnames(m)) m[a, b] else NA_real_

connections_df <- purrr::map_dfr(V(g_sub2)$name[V(g_sub2)$name %in% targets], function(smag) {
  metabs <- setdiff(neighbors(g_sub2, smag)$name, all_MAGs)
  purrr::map_dfr(metabs, function(metab) {
    meths <- intersect(neighbors(g_sub2, metab)$name, methanogens)
    if (!length(meths)) return(NULL)
    w1 <- get_edge_weight(smag, metab, assoMat_FULL)
    purrr::map_dfr(meths, function(meth) {
      w2 <- get_edge_weight(metab, meth, assoMat_FULL)
      data.frame(genome = smag, metabolite = metab, methanogen = meth,
                 weight_genome_metab = w1, weight_metab_methanogen = w2,
                 sign_genome_metab     = ifelse(is.na(w1), NA, ifelse(w1 > 0, "positive", "negative")),
                 sign_metab_methanogen = ifelse(is.na(w2), NA, ifelse(w2 > 0, "positive", "negative")))
    })
  })
})
if (nrow(connections_df) == 0) stop("No genome-metabolite-methanogen paths found.")
cat(sprintf("  Genome-methanogen connections: %d\n", nrow(connections_df)))

df_metab <- tibble::rownames_to_column(data.frame(tax_table(metabolites_phylo_sub)), "metabolite")
df_metab$Final_name <- apply_name_aliases(df_metab$Final_name)
df_MAG   <- data.frame(tax_table(ps_MAG_sub))

write_out(connections_df, p_net("Linking_Metabolites.csv"))
write_out(left_join(connections_df, df_metab, by = "metabolite"),
          p_net("Linking_Metabolites_INFO.csv"))

path_verts <- unique(c(connections_df$genome, connections_df$metabolite, connections_df$methanogen))
g_net <- induced_subgraph(g_sub2, path_verts)
drop_same_type <- function(g, type) {
  ee <- ends(g, E(g))
  t1 <- ifelse(ee[,1] %in% all_MAGs, "MAG", "Metabolite")
  t2 <- ifelse(ee[,2] %in% all_MAGs, "MAG", "Metabolite")
  delete_edges(g, which(t1 == type & t2 == type))
}
g_net <- drop_same_type(drop_same_type(g_net, "Metabolite"), "MAG")
cat(sprintf("  Final: %d nodes | %d edges\n", vcount(g_net), ecount(g_net)))
saveRDS(g_net, p_net("igraph_Fig3_network.RDS"))

# ---- Edge recovery under node resampling -------------------------------------
# The stored result in 00_inputs/ is valid only for the edge set it was computed
# on. If the metabolite filter, the abundance quantile or the SpiecEasi fit
# changes the displayed edges, the knit either recomputes (RUN_RESAMPLE) or stops.

fig3_edges  <- apply(ends(g_net, E(g_net)), 1, \(x) paste(sort(x), collapse = "|"))
stab_path   <- p_raw("Check_Fig3_edge_stability.csv")
stab_stored <- if (file.exists(stab_path))
  read_csv(stab_path, show_col_types = FALSE) else NULL
stab_stale  <- is.null(stab_stored) || !setequal(stab_stored$edge, fig3_edges)

if (stab_stale && !RUN_RESAMPLE)
  stop("Check_Fig3_edge_stability.csv does not describe the current Figure 3 edge set. ",
       if (is.null(stab_stored)) "The file is missing. " else paste0(
         length(setdiff(stab_stored$edge, fig3_edges)), " edge(s) only in the file, ",
         length(setdiff(fig3_edges, stab_stored$edge)), " only in the graph. "),
       "Set RUN_RESAMPLE <- TRUE and re-knit; budget roughly ", RESAMPLE_N,
       " x the SpiecEasi fit time.")

if (stab_stale) {
  message("Edge set changed. Refitting SpiecEasi ", RESAMPLE_N,
          " times on random 80% subsets of the metabolite set.")
  edge_stability <- resample_edge_recovery(fig3_edges, mag_mat, met_mat, RESAMPLE_N)
  write.csv(edge_stability, stab_path, row.names = FALSE)
} else {
  edge_stability <- stab_stored
}

metab_name_all <- setNames(df_metab$Final_name, df_metab$metabolite)
edge_stability <- edge_stability |>
  tidyr::separate(edge, c("node_a", "node_b"), sep = "\\|", remove = FALSE) |>
  dplyr::mutate(
    MAG             = ifelse(node_a %in% all_MAGs, node_a, node_b),
    metabolite      = ifelse(node_a %in% all_MAGs, node_b, node_a),
    metabolite_name = unname(metab_name_all[metabolite]),
    edge_type       = ifelse(MAG %in% methanogens,
                             "metabolite-methanogen", "MAG-metabolite")) |>
  dplyr::select(edge, MAG, metabolite, metabolite_name, edge_type, recovery_frequency) |>
  dplyr::arrange(dplyr::desc(recovery_frequency))

rf <- edge_stability$recovery_frequency
cat(sprintf(
  "\nFigure 3 edge recovery, %d resampling replicates: %d edges, median %.2f, mean %.2f, range %.2f to %.2f\n",
  RESAMPLE_N, length(rf), median(rf), mean(rf), min(rf), max(rf)))
for (thr in c(0.5, 0.7, 0.8, 0.9))
  cat(sprintf("  recovered in >= %.0f%% of replicates: %d of %d\n",
              100 * thr, sum(rf >= thr), length(rf)))

# ---- Node metadata and plot --------------------------------------------------

vn <- V(g_net)$name
metab_superclass <- setNames(df_metab$Superclass, df_metab$metabolite)

classify_metab <- function(superclass) dplyr::case_when(
  superclass == "Lipids and lipid-like molecules"      ~ "Lipids and derivatives",
  superclass %in% c("Organic acids and derivatives", "Organic nitrogen compounds",
                    "Organoheterocyclic compounds", "Nucleosides, nucleotides, and analogues",
                    "Alkaloids and derivatives")       ~ "Amino acids and derivatives",
  superclass %in% NETWORK_PHENOLIC_SUPERCLASSES        ~ "Phenolic and polyphenolic compounds",
  TRUE                                                 ~ "Other organic compounds")

node_group <- dplyr::case_when(
  vn %in% methanogens ~ "Methanogen",
  vn %in% G1 ~ "Warming-stimulated anoxic-preferring",
  vn %in% G2 ~ "Warming-stimulated oxic-preferring",
  vn %in% G3 ~ "Warming-inhibited",
  TRUE ~ classify_metab(metab_superclass[vn]))
node_type  <- ifelse(vn %in% all_MAGs, "MAG", "Metabolite")

mag_lookup   <- setNames(paste0(df_MAG$SPRUCE_ID, ": ", df_MAG$Phylum), df_MAG$SPRUCE_ID)
metab_lookup <- setNames(df_metab$Final_name, df_metab$metabolite)

# Two independent reasons a Figure 3 node should not carry a compound name, both
# applied:
#
#   1. MSI level. L3 and L4 assignments are computational proposals; the name is
#      the best-matching candidate structure, not an identification. Printing it
#      on a figure asserts more than the data support. Only L1 and L2 features
#      (MSI_NAMEABLE) keep their name.
#   2. Biological implausibility. A handful of L1/L2 annotations are implausible
#      in an ombrotrophic bog regardless of the spectral evidence. Estriol is the
#      clearest case: it passes as L2 and is a mammalian steroid.
#
# Everything else is labelled by ClassyFire class plus the feature identifier, so
# the node is traceable to a row of Supplementary Data 10 without asserting a
# structure. The candidate name stays in Supplementary Data 10.
UNEXPECTED_ANNOTATIONS <- c(
  "Estriol", "17-Amino-5alpha-androstan-11-ol", "cis-Dehydroretinal",
  "Pfaffic acid", "Decarboxyhypothamnolic acid", "Ramalinic acid A")

# Two compound names are too long for a label box. Shortened for the figure
# only; the full name is in Supplementary Data 10. The kavalactone pair is the
# NAME_ALIASES diastereomer, so both features carry the same shortened label.
FIG3_NAME_SHORT <- c(
  "5-Hydroxy-4-methoxy-6-(2-phenylethyl)-5,6-dihydro-2H-pyran-2-one (stereochemistry unresolved)" =
    "Kavalactone\n(stereochemistry unresolved)")

# Some ClassyFire strings are too long to sit on a node. Shortened for the figure
# only; the full class is in Supplementary Data 10.
CLASS_LABEL_SHORT <- c(
  "Lignans, neolignans and related compounds" = "Lignan-like",
  "Steroids and steroid derivatives"          = "Steroid",
  "Prenol lipids"                             = "Prenol lipid",
  "Depsides and depsidones"                   = "Depsidone")

blankna <- function(x) ifelse(is.na(x) | x %in% c("", "NA"), NA_character_, x)
class_label <- function(feature) {
  r <- df_metab[match(feature, df_metab$metabolite), , drop = FALSE]
  lab <- dplyr::coalesce(blankna(r$Subclass), blankna(r$Class),
                         blankna(r$Superclass))
  short <- unname(CLASS_LABEL_SHORT[lab])
  lab <- ifelse(is.na(short), lab, short)
  ifelse(is.na(lab), "unclassified feature", paste0(lab, " (putative)"))
}

metab_msi     <- setNames(df_metab$Annotation_Confidence, df_metab$metabolite)
metab_display <- unname(metab_lookup[vn])
is_metab      <- !(vn %in% all_MAGs)

short_hit <- !is.na(metab_display) & metab_display %in% names(FIG3_NAME_SHORT)
metab_display[short_hit] <- unname(FIG3_NAME_SHORT[metab_display[short_hit]])

msi_vn   <- unname(metab_msi[vn])
low_conf <- is_metab & (is.na(msi_vn) | !(msi_vn %in% MSI_NAMEABLE))
implaus  <- is_metab & !is.na(metab_display) &
            metab_display %in% UNEXPECTED_ANNOTATIONS
reclass  <- low_conf | implaus

reclass_reason <- dplyr::case_when(
  low_conf & implaus ~ "MSI L3/L4 and implausible in an ombrotrophic bog",
  low_conf           ~ "MSI L3/L4 or unannotated: name is a candidate structure",
  implaus            ~ "implausible in an ombrotrophic bog",
  TRUE               ~ NA_character_)

if (any(reclass)) {
  # Class on the first line, feature ID on the second, so the node stays
  # traceable to a row of Supplementary Data 10.
  metab_display[reclass] <- paste0(class_label(vn[reclass]), "\n", vn[reclass])
  cat(sprintf("  Figure 3: %d of %d metabolite label(s) replaced by class + feature ID\n",
              sum(reclass), sum(is_metab)))
  print(table(reclass_reason[reclass]))
}

# The network filter requires a non-missing Final_name, so this should never
# fire; if the filter is ever loosened, the node falls back to its feature ID
# rather than printing the string "NA".
no_label <- is_metab & is.na(metab_display)
if (any(no_label)) {
  warning(sum(no_label), " metabolite node(s) reached Figure 3 with no label; ",
          "using the feature ID.")
  metab_display[no_label] <- vn[no_label]
}

# Every metabolite label carries its MSI level, so a reader can see at a glance
# how much of the network rests on computational annotation.
msi_tag <- ifelse(is.na(msi_vn), "unannotated", msi_vn)
metab_display[is_metab] <- paste0(metab_display[is_metab], " [", msi_tag[is_metab], "]")

display_name <- ifelse(vn %in% all_MAGs, unname(mag_lookup[vn]), metab_display)
display_name[is.na(display_name)] <- vn[is.na(display_name)]

fermenter_ids <- c("SPRUCE 124", "SPRUCE 675")
all_levels <- c("Methanogen", "Warming-stimulated anoxic-preferring",
                "Warming-stimulated oxic-preferring", "Warming-inhibited",
                "Amino acids and derivatives", "Phenolic and polyphenolic compounds",
                "Lipids and derivatives", "Other organic compounds")
g_net <- set_vertex_attr(g_net, "group",        value = factor(node_group, levels = all_levels))
g_net <- set_vertex_attr(g_net, "node_type",    value = node_type)
g_net <- set_vertex_attr(g_net, "display_name", value = display_name)
g_net <- set_vertex_attr(g_net, "shape_group",  value = dplyr::case_when(
  vn %in% fermenter_ids ~ "Fermenter MAG", vn %in% all_MAGs ~ "MAG", TRUE ~ "Metabolite"))
print(table(node_group, useNA = "always"))

# Tripartite layout: genomes left, metabolites centre, methanogens right.
#
# The columns are pushed further apart than the node spacing needs, to open a
# gutter between the metabolite column and the methanogen column for the
# compound-name boxes. X_LABEL_RIGHT is the right edge every metabolite label
# is aligned to; the labels are right-justified there and grow leftward, so
# their right edges form a straight line whatever the name length.
X_MAG        <- -6
X_METAB      <-  0
X_METHANOGEN <-  6
X_LABEL_RIGHT <- 5.1     # right edge of the compound-name boxes
X_MAG_LABEL   <- -6.4    # MAG labels sit left of their column, right-aligned
X_METH_LABEL  <-  6.4    # methanogen labels sit right of theirs, left-aligned

# The metabolite column sets the vertical scale, because it has the most nodes
# and its labels are the two-line ones. METAB_ROW_GAP has to clear the height of
# a two-line box or the grey boxes overlap; it is the number to change if they
# ever do. The other two columns are then spread over the same total height, so
# the panel has no dead space and the MAG labels are not crowded.
METAB_ROW_GAP <- 2.1
y_half <- 0.5 * METAB_ROW_GAP * (sum(node_type == "Metabolite") - 1)

layout_mat <- matrix(NA_real_, nrow = vcount(g_net), ncol = 2, dimnames = list(vn, NULL))
for (grp in list(list(v = vn[node_type == "MAG" & !(vn %in% methanogens)], x = X_MAG),
                 list(v = vn[node_type == "Metabolite"],                   x = X_METAB),
                 list(v = vn[vn %in% methanogens],                         x = X_METHANOGEN))) {
  n <- length(grp$v)
  layout_mat[grp$v, 1] <- grp$x
  layout_mat[grp$v, 2] <- if (n == 1) 0 else seq(-y_half, y_half, length.out = n)
}

# Label frames, built off the same layout so a label can never drift from its
# node. Kept out of the graph object because their x is fixed by column, not by
# the node position, which is what makes the right edges line up.
lab_df <- data.frame(name = vn, y = layout_mat[vn, 2], label = display_name,
                     stringsAsFactors = FALSE)
lab_metab <- lab_df[node_type == "Metabolite", ]
lab_mag   <- lab_df[node_type == "MAG" & !(vn %in% methanogens), ]
lab_meth  <- lab_df[vn %in% methanogens, ]

node_colors <- c(
  "Methanogen" = "#D740F2", "Warming-stimulated anoxic-preferring" = "#CA9F01",
  "Warming-stimulated oxic-preferring" = "#538F6B", "Warming-inhibited" = "#0066CC",
  "Amino acids and derivatives" = "#E07040",
  "Phenolic and polyphenolic compounds" = "#7B3030",
  "Lipids and derivatives" = "#B8964A", "Other organic compounds" = "#999999")

FIG3 <- ggraph(g_net, layout = layout_mat[vn, ]) +
  geom_edge_link(aes(color = color), alpha = 0.6, width = 1.2) +
  scale_edge_color_identity() +
  geom_node_point(data = \(x) x[x$node_type == "Metabolite", ],
                  aes(fill = group, colour = shape_group), shape = 24, size = 5, stroke = 0.5) +
  geom_node_point(data = \(x) x[x$node_type == "MAG" & !(x$name %in% fermenter_ids), ],
                  aes(fill = group, colour = shape_group), shape = 21, size = 8, stroke = 0.8) +
  geom_node_point(data = \(x) x[x$node_type == "MAG" & x$name %in% fermenter_ids, ],
                  aes(fill = group, colour = shape_group), shape = 22, size = 8, stroke = 0.8) +
  scale_fill_manual(values = node_colors, breaks = all_levels[1:4],
    name = "Genome classification",
    guide = guide_legend(order = 1, nrow = 2,
      override.aes = list(shape = 21, size = 4, stroke = 0.8, color = "black"))) +
  # Compound names: grey box, placed in the gutter to the right of the
  # metabolite column, right-justified at X_LABEL_RIGHT so every box ends on the
  # same vertical line. The box is opaque so a name stays legible where an edge
  # runs under it. inherit.aes = FALSE keeps these off the node aesthetics.
  geom_label(data = lab_metab, inherit.aes = FALSE,
             aes(x = X_LABEL_RIGHT, y = y, label = label),
             hjust = 1, vjust = 0.5, size = 3.2, lineheight = 0.95,
             colour = "grey15", fill = "grey90", alpha = 1,
             label.size = 0, label.r = unit(0.06, "lines"),
             label.padding = unit(0.16, "lines")) +
  # A leader from each metabolite node to its box, so a reader can follow a name
  # back to its triangle across the gutter.
  geom_segment(data = lab_metab, inherit.aes = FALSE,
               aes(x = X_METAB + 0.12, xend = X_LABEL_RIGHT, y = y, yend = y),
               colour = "grey75", linewidth = 0.2, linetype = "dotted") +
  # MAGs and methanogens keep plain text, outside their own columns, so nothing
  # competes with the compound-name gutter.
  geom_text(data = lab_mag, inherit.aes = FALSE,
            aes(x = X_MAG_LABEL, y = y, label = label),
            hjust = 1, size = 3.4, colour = "grey15") +
  geom_text(data = lab_meth, inherit.aes = FALSE,
            aes(x = X_METH_LABEL, y = y, label = label),
            hjust = 0, size = 3.4, colour = "grey15") +
  scale_colour_manual(values = c(MAG = "black", `Fermenter MAG` = "black", Metabolite = "black"),
    name = "Node shape",
    guide = guide_legend(order = 2, override.aes = list(
      shape = c(21, 22, 24), size = c(4, 4, 3), fill = "grey60", stroke = 0.8, linetype = 0))) +
  # The label gutter and the two outer text columns sit outside the node
  # coordinates, so the panel has to be told to make room for them.
  coord_cartesian(xlim = c(X_MAG_LABEL - 4.2, X_METH_LABEL + 4.2), clip = "off") +
  theme_void() + theme(legend.position = "bottom",
                       plot.margin = margin(6, 6, 6, 6))
save_fig(FIG3, p_net("Figure_3_network.pdf"), height = 14, width = 14)
save_fig(FIG3, p_fig("Figure_3_network.pdf"), height = 14, width = 14)

# ---- Supplementary Data 10 ----------------------------------------------------

# Each row is a genome-metabolite-methanogen path, so it spans two edges and
# carries two recovery frequencies. Edge keys are built the same way as in
# edge_stability above: the node pair sorted alphabetically and "|"-separated,
# so the join is on feature IDs and never on display names.
edge_key <- function(a, b) purrr::map2_chr(a, b, \(x, y) paste(sort(c(x, y)), collapse = "|"))

# What Figure 3 prints on each metabolite node, keyed on the feature ID. Carried
# into Supplementary Data 10 so that a claim in the Results can be traced from the
# figure label to the candidate name, the MSI level and the two edge recoveries
# without joining anything by hand.
fig3_label <- setNames(gsub("\n", " | ", metab_display[is_metab]), vn[is_metab])

Suppl_Data_10 <- connections_df |>
  mutate(connected_sensitive_tax = unname(mag_lookup[genome]),
         connected_methanogen    = methanogen,
         Metabolite_feature_ID   = metabolite,
         Metabolite_name         = unname(metab_lookup[metabolite]),
         Figure_3_label          = unname(fig3_label[metabolite]),
         key_genome_metab        = edge_key(genome,     metabolite),
         key_metab_methanogen    = edge_key(metabolite, methanogen)) |>
  left_join(dplyr::select(df_metab, metabolite, Kingdom, Superclass,
                          Class.metab = Class, Subclass, starts_with("Parent.Level")),
            by = "metabolite") |>
  # Mode, RT, m/z, MSI level and name multiplicity, so the network table carries
  # the same annotation bookkeeping as Supplementary Data 3 and 4.
  left_join(dplyr::select(taxdf_met, metabolite = feature, Acquisition_mode,
                          RT_min, precursor_mz, Annotation_Confidence,
                          Name_n_features, Name_status),
            by = "metabolite") |>
  left_join(dplyr::select(edge_stability,
                          key_genome_metab      = edge,
                          recovery_genome_metab = recovery_frequency),
            by = "key_genome_metab") |>
  left_join(dplyr::select(edge_stability,
                          key_metab_methanogen      = edge,
                          recovery_metab_methanogen = recovery_frequency),
            by = "key_metab_methanogen") |>
  mutate(Classification_network  = classify_metab(Superclass),
         edge_genome_metab       = sign_genome_metab,
         edge_metab_methanogen   = sign_metab_methanogen,
         weight_genome_metab     = round(weight_genome_metab, 4),
         weight_metab_methanogen = round(weight_metab_methanogen, 4)) |>
  dplyr::select(connected_sensitive_tax, connected_methanogen,
                Metabolite_feature_ID, Acquisition_mode, RT_min, precursor_mz,
                Metabolite_name, Figure_3_label, Annotation_Confidence,
                Name_n_features, Name_status,
                Kingdom, Superclass, Class.metab, Subclass, starts_with("Parent.Level"),
                Classification_network,
                edge_genome_metab,     weight_genome_metab,     recovery_genome_metab,
                edge_metab_methanogen, weight_metab_methanogen, recovery_metab_methanogen) |>
  mutate(across(where(is.character), \(x) na_if(x, ""))) |>
  arrange(connected_sensitive_tax, connected_methanogen, Metabolite_name)

# Every path edge must appear in edge_stability. An NA here means the Figure 3
# edge set and the resampling run have diverged, which the setequal guard above
# should already have caught.
if (anyNA(Suppl_Data_10$recovery_genome_metab) ||
    anyNA(Suppl_Data_10$recovery_metab_methanogen))
  stop("Supplementary Data 10: ",
       sum(is.na(Suppl_Data_10$recovery_genome_metab)), " genome-metabolite and ",
       sum(is.na(Suppl_Data_10$recovery_metab_methanogen)), " metabolite-methanogen ",
       "edge(s) have no recovery frequency. Rerun the resampling chunk.")

cat(sprintf("  Supplementary Data 10: %d paths, %d edge recovery values joined\n",
            nrow(Suppl_Data_10), 2 * nrow(Suppl_Data_10)))

write_out(Suppl_Data_10, p_net("Suppl_Data_10_linked_metabolites.csv"), na = "NA")
write_out(Suppl_Data_10, p_sd("Suppl_Data_10_linked_metabolites.csv"),  na = "NA")

# Network summary numbers quoted in the text and the Figure 3 legend.
fig3_summary <- data.frame(
    stars_stability      = getStability(se_FULL),
    edges_full           = sum(getRefit(se_FULL)) / 2,
    degree_threshold     = DEGREE_THRESHOLD,
    abundance_quantile   = ABUND_QUANTILE,
    n_MAGs_input         = ncol(mag_mat),
    n_metabolites_input  = ncol(met_mat),
    n_samples            = nrow(mag_mat),
    fig3_nodes           = vcount(g_net),
    fig3_edges           = ecount(g_net),
    n_connections        = nrow(connections_df),
    fig3_metabolites     = sum(is_metab),
    fig3_metab_named     = sum(is_metab & !reclass),
    fig3_metab_by_class  = sum(reclass),
    fig3_metab_L1        = sum(is_metab & msi_tag == "L1"),
    fig3_metab_L2        = sum(is_metab & msi_tag == "L2"),
    fig3_metab_L3        = sum(is_metab & msi_tag == "L3"),
    fig3_metab_L4        = sum(is_metab & msi_tag == "L4"),
    edge_recovery_median = median(edge_stability$recovery_frequency),
    edge_recovery_min    = min(edge_stability$recovery_frequency),
    edge_recovery_max    = max(edge_stability$recovery_frequency),
    resample_replicates  = RESAMPLE_N)
print(t(fig3_summary))
```

`Check_Fig3_edge_stability.csv` is produced by `resample_edge_recovery()`, defined
in `network-build` and called from `network-graph` whenever the stored file does
not describe the edge set actually plotted. It is not a separate chunk any more:
keeping the resampling code in one place is what stops the recovery frequencies
and the figure describing different graphs. The result is written back into
`00_inputs/` so that `WIPE_OUTPUTS` cannot delete it, and is reused on subsequent
knits for as long as the edge set is unchanged.

## 6. GENE-LEVEL RESULTS — Figure 4, Supplementary Figures S5, S6 and S7

Predictor names are used exactly as the Maaslin2 objects store them,
`AvgTemp10cm` and `WT_drawdown_cm`.

```{r gene-level-setup}

TEMP_GENE <- "AvgTemp10cm"
WT_GENE   <- "WT_drawdown_cm"

# Maaslin2 takes its feature names from taxa_names(); every join below is on the
# Gene column of the tax table. They have to be the same vector.
if (!identical(taxa_names(Genes_withMAG_NORMALIZED),
               unname(as.character(tax_table(Genes_withMAG_NORMALIZED)[, "Gene"]))))
  stop("taxa_names() and the Gene column of the gene tax table are not the same ",
       "vector. Every gene-level join below is on Gene and would return NA ",
       "annotations.")

subset_genes <- function(ids, camper_only = TRUE) {
  ps <- prune_taxa(rownames(tax_table(Genes_withMAG_NORMALIZED))[
    tax_table(Genes_withMAG_NORMALIZED)[, "SPRUCE_ID"] %in% ids], Genes_withMAG_NORMALIZED)
  if (camper_only) ps <- subset_taxa(ps, CAMPER_id != "" & !is.na(CAMPER_id))
  ps
}
```

```{r fig-s-cytochrome, fig.width=9, fig.height=3}

# --- Supplementary Figure S5 and Supplementary Data 9 -------------------------

Thermo_ox.genes.ps <- subset_genes(WSOX_IDS, camper_only = FALSE)
tax_ox_all <- data.frame(tax_table(Thermo_ox.genes.ps))
target_cyt <- rownames(tax_ox_all)[grep("cytochrome c", tax_ox_all$pfam_hits,
                                        ignore.case = TRUE)]
cat("Cytochrome c genes in WS-OX genomes:", length(target_cyt), "\n")

if (REFIT_MAASLIN) {
  fit_maaslin(prune_taxa(target_cyt, Thermo_ox.genes.ps), "RE_Thermo_cytochrome",
              fixed = c("AvgTemp10cm", "WT_drawdown_cm"), random = "Plot")
}

Suppl_Data_09 <- readRDS(maaslin_path("RE_Thermo_cytochrome.rds"))$results |>
  dplyr::rename(Gene = 1) |>
  add_effect_size() |>
  left_join(tax_ox_all, by = "Gene") |>
  filter(metadata == TEMP_GENE) |>
  arrange(pval)
write_out(Suppl_Data_09, p_sd("Suppl_Data_09_cytochrome_c_temperature.csv"))

df_Thermo_ox4 <- prune_taxa(target_cyt, Thermo_ox.genes.ps) |>
  psmelt_with_camper() |>
  mutate(WT_drawdown_cm = WT_drawdown * 100) |>
  group_by(SPRUCE_ID, CAMPER_gene_description, Year, AvgTemp10cm,
           WT_drawdown_cm, Warming_Treatment.ord) |>
  dplyr::summarise(mean_Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") |>
  mutate(Year = as.factor(Year))

save_fig(
  ggplot(df_Thermo_ox4, aes(x = AvgTemp10cm, y = mean_Abundance)) +
    geom_point(aes(fill = SPRUCE_ID, shape = Year), colour = "black", size = 5, alpha = 0.8) +
    geom_smooth(method = "lm", color = "black", linetype = "dashed", linewidth = 1, se = FALSE) +
    facet_wrap(~CAMPER_gene_description, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = scico(length(unique(df_Thermo_ox4$SPRUCE_ID)), palette = "bamako")) +
    scale_shape_manual(values = c(21, 22, 24)) +
    guides(fill = guide_legend("Genome", override.aes = list(shape = 21))) +
    xlab("Temperature (ºC, May-Aug)") +
    ylab("Relative activity of gene\n(TAD80/ribosomal expression)") +
    theme_bw() +
    theme(legend.position = "right", axis.text = element_text(size = 12),
          axis.title = element_text(size = 16), strip.text = element_text(size = 12),
          strip.background = element_blank()),
  p_fig("Figure_S5_cytochrome_c.pdf"), height = 3, width = 9)
```

```{r fig4, fig.width=8, fig.height=4}

# --- Figure 4: phenol-active gene responses in WS-OX and WS-AN genomes --------

read_gene_maaslin <- function(rds, ps) {
  tax <- data.frame(tax_table(ps))
  stopifnot("Gene" %in% names(tax))
  readRDS(rds)$results |>
    dplyr::rename(Gene = 1) |>
    add_effect_size() |>
    left_join(tax, by = "Gene")
}

ps_ox <- subset_genes(WSOX_IDS)
ps_an <- subset_genes(WSAN_IDS)
if (REFIT_MAASLIN) {
  fit_maaslin(ps_ox, "RE_Thermo_CAMPERgenes",     c("AvgTemp10cm", "WT_drawdown_cm"), "Plot")
  fit_maaslin(ps_an, "RE_ThermoANOX_CAMPERgenes", c("AvgTemp10cm", "WT_drawdown_cm"), "Plot")
}
df_Coef_Therm <- read_gene_maaslin(maaslin_path("RE_Thermo_CAMPERgenes.rds"),     ps_ox)
df_Coef_AN    <- read_gene_maaslin(maaslin_path("RE_ThermoANOX_CAMPERgenes.rds"), ps_an)
stopifnot(all(c(TEMP_GENE, WT_GENE) %in% df_Coef_Therm$metadata),
          all(c(TEMP_GENE, WT_GENE) %in% df_Coef_AN$metadata))

# Genes moving in one direction, per genome, per oxygen class, per predictor.
# The grid is built explicitly so a genome with no significant genes sits at (0, 0).
count_gene_responses <- function(df, direction = c("up", "down"), genomes = NULL) {
  direction <- match.arg(direction)
  grid <- expand.grid(
    SPRUCE_ID     = if (is.null(genomes)) sort(unique(df$SPRUCE_ID)) else sort(genomes),
    metadata      = sort(unique(df$metadata)),
    CAMPER_oxygen = c("oxic", "anoxic"),
    stringsAsFactors = FALSE)
  df |>
    filter(pval < 0.05, if (direction == "up") coef > 0 else coef < 0,
           CAMPER_oxygen %in% c("oxic", "anoxic")) |>
    count(SPRUCE_ID, metadata, CAMPER_oxygen, name = "n_genes") |>
    right_join(grid, by = c("SPRUCE_ID", "metadata", "CAMPER_oxygen")) |>
    mutate(n_genes = tidyr::replace_na(n_genes, 0L)) |>
    tidyr::pivot_wider(names_from = CAMPER_oxygen, values_from = n_genes,
                       names_prefix = "genes_")
}

camper_scatter <- function(df, title, colors, axis_prefix, legend_pos = "none") {
  ggplot(df, aes(x = genes_oxic, y = genes_anoxic)) +
    geom_point(aes(fill = taxonomy), colour = "black", pch = 21, size = 5) +
    scale_fill_manual(values = colors) +
    scale_x_continuous(breaks = scales::breaks_pretty()) +
    scale_y_continuous(breaks = scales::breaks_pretty()) +
    guides(fill = guide_legend("Genome", override.aes = list(shape = 21))) +
    xlab(paste0("Oxygen-dependent genes\n", axis_prefix)) +
    ylab(paste0("Oxygen-independent genes\n", axis_prefix)) +
    ggtitle(title) + theme_bw() +
    theme(legend.position = legend_pos, axis.text = element_text(size = 16),
          axis.title = element_text(size = 16))
}

add_taxonomy <- function(counts, tax) {
  counts |>
    left_join(distinct(dplyr::select(tax, SPRUCE_ID, Phylum, Genus)), by = "SPRUCE_ID") |>
    mutate(taxonomy = paste0(SPRUCE_ID, ": ", Phylum,
                             ifelse(Genus == "" | is.na(Genus), "", paste0(" g__", Genus))))
}

FIG4B_GENOMES <- c("SPRUCE 124", "SPRUCE 2057", "SPRUCE 2060", "SPRUCE 2148",
                   "SPRUCE 2155", "SPRUCE 2191", "SPRUCE 275")
if (length(setdiff(FIG4B_GENOMES, df_Coef_AN$SPRUCE_ID)))
  warning("Figure 4B genomes absent from the WS-AN results: ",
          paste(setdiff(FIG4B_GENOMES, df_Coef_AN$SPRUCE_ID), collapse = ", "))

genes_df_OX <- add_taxonomy(count_gene_responses(df_Coef_Therm, "up", WSOX_IDS),
                            data.frame(tax_table(ps_ox)))
genes_df_AN <- add_taxonomy(count_gene_responses(df_Coef_AN, "up", FIG4B_GENOMES),
                            data.frame(tax_table(ps_an)))

green   <- scico(max(7, n_distinct(genes_df_OX$taxonomy)), palette = "bamako")
yellows <- c("#663329","#A54543","#D85F4D","#E48751","#ECAC54","#F7D971","#FFFECB")

fig4a_A <- camper_scatter(filter(genes_df_OX, metadata == WT_GENE),   "Water table drawdown", green, "increasing")
fig4a_B <- camper_scatter(filter(genes_df_OX, metadata == TEMP_GENE), "Temperature",          green, "increasing", "right")
save_legend(fig4a_B, legend_path(p_fig("Figure_4A_WSOX_phenolic_genes.pdf")), width = 3.5, height = 5)
save_fig(ggarrange(fig4a_A, fig4a_B + theme(legend.position = "none"),
    ncol = 2, labels = c("A","B"), font.label = list(size = 18)),
  p_fig("Figure_4A_WSOX_phenolic_genes.pdf"), width = 8, height = 4)

fig4b_A <- camper_scatter(filter(genes_df_AN, metadata == WT_GENE),   "Water table drawdown", yellows, "increasing")
fig4b_B <- camper_scatter(filter(genes_df_AN, metadata == TEMP_GENE), "Temperature",          yellows, "increasing", "right")
save_legend(fig4b_B, legend_path(p_fig("Figure_4B_WSAN_phenolic_genes.pdf")), width = 3.5, height = 5)
save_fig(ggarrange(fig4b_A, fig4b_B + theme(legend.position = "none"),
    ncol = 2, labels = c("A","B"), font.label = list(size = 18)),
  p_fig("Figure_4B_WSAN_phenolic_genes.pdf"), width = 8, height = 4)

# Gene counts quoted in the Results.
cat("\nWS-OX upregulated phenol-active genes by predictor:\n")
print(genes_df_OX |> group_by(metadata) |>
        summarise(total = sum(genes_oxic + genes_anoxic), .groups = "drop"))
cat("WS-OX downregulated:\n")
print(count_gene_responses(df_Coef_Therm, "down", WSOX_IDS) |> group_by(metadata) |>
        summarise(total = sum(genes_oxic + genes_anoxic), .groups = "drop"))
cat("WS-AN upregulated:\n")
print(genes_df_AN |> group_by(metadata) |>
        summarise(total = sum(genes_oxic + genes_anoxic), .groups = "drop"))

# Supplementary Data 12 and 13: the gene-level tables behind Figure 4.
write_out(filter(df_Coef_Therm, pval < 0.05) |> arrange(metadata, pval),
          p_sd("Suppl_Data_12_WSOX_CAMPER_genes.csv"))
write_out(filter(df_Coef_AN,    pval < 0.05) |> arrange(metadata, pval),
          p_sd("Suppl_Data_13_WSAN_CAMPER_genes.csv"))

# --- Supplementary Data 11: hydrogenases in WS-AN genomes, HydDB annotated ---

ps_an_hyd <- prune_taxa(
  rownames(tax_table(load_hyddb_ps()))[
    tax_table(load_hyddb_ps())[, "SPRUCE_ID"] %in% WSAN_IDS],
  load_hyddb_ps())
if (REFIT_MAASLIN) {
  fit_maaslin(ps_an_hyd, "RE_ThermoANOX_hydrogenases",
              fixed = c("AvgTemp10cm", "WT_drawdown_cm"), random = "Plot")
}

tax_an_hyd <- data.frame(tax_table(ps_an_hyd))
hyd_cols   <- grep("^HydDB", names(tax_an_hyd), value = TRUE)
if (!length(hyd_cols))
  warning("No HydDB columns in ", HYDDB_PS_FILE,
          ". Supplementary Data 11 will have no hydrogenase annotation.")

Suppl_Data_11 <- readRDS(maaslin_path("RE_ThermoANOX_hydrogenases.rds"))$results |>
  dplyr::rename(Gene = 1) |>
  add_effect_size() |>
  left_join(tax_an_hyd, by = "Gene") |>
  filter(pval < 0.05,
         if (length(hyd_cols)) !is.na(.data[[hyd_cols[1]]]) & .data[[hyd_cols[1]]] != ""
         else grepl("hydrogenase", kegg_hit, ignore.case = TRUE)) |>
  arrange(metadata, pval)
write_out(Suppl_Data_11, p_sd("Suppl_Data_11_hydrogenases_HydDB.csv"))
```

```{r fig-s-aminocarb, fig.width=12, fig.height=15}

# --- Supplementary Figures S6 and S7 ------------------------------------------

Select_taxa <- read.csv(net_path("Linking_Metabolites_INFO.csv")) |>
  pull(genome) |> unique() |> setdiff(c("SPRUCE 190", "SPRUCE 689"))
# Three genes, each in a pathway enriched in Figure 5: K00382 (dihydrolipoyl
# dehydrogenase, branched-chain amino acid degradation), K01915 (glutamine
# synthetase, alanine/aspartate/glutamate metabolism) and K00873 (pyruvate
# kinase, glycolysis, pyruvate metabolism and carbon metabolism).
target_ko <- c("K00382", "K01915", "K00873")   # dihydrolipoamide DH, GlnA, pyruvate kinase

subset_data_df_plot <- subset_genes(Select_taxa, camper_only = FALSE) |>
  psmelt_with_camper() |>
  filter(ko_id %in% target_ko) |>
  group_by(SPRUCE_ID, ko_id, pfam_hits, Phylum, Genus, Year, AvgTemp10cm,
           Temp_avg_top, WT_drawdown, Warming_Treatment.ord) |>
  dplyr::summarise(mean_Abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") |>
  mutate(taxonomy = paste0(SPRUCE_ID, ": ", Phylum, " g__", Genus), Year = as.factor(Year))

colors_aminocarb <- c("#663329","#A54543","#D85F4D","#E48751","#ECAC54","#F7D971","#FFFECB",
                       "#003A46","#15483A","#355E26","#988C02","#D0B74B","#FFE5AC","#C5DBA0")

save_fig(
  merge(expand.grid(taxonomy = unique(subset_data_df_plot$taxonomy), ko_id = target_ko),
        subset_data_df_plot, by = c("taxonomy","ko_id"), all.x = TRUE) |>
    ggplot(aes(x = AvgTemp10cm, y = mean_Abundance)) +
    geom_point(aes(fill = taxonomy, shape = Year), colour = "black", size = 5, alpha = 0.8) +
    facet_wrap(~taxonomy + ko_id, scales = "free_y", ncol = 3) +
    scale_fill_manual(values = colors_aminocarb) +
    scale_shape_manual(values = c(21, 22, 24)) +
    xlab("Temperature (ºC, May-Aug)") +
    ylab("Relative activity of gene\n(TAD80/ribosomal expression)") +
    theme_bw() +
    theme(legend.position = "none", axis.text = element_text(size = 12),
          axis.title = element_text(size = 16), strip.text = element_blank(),
          strip.background = element_blank()),
  p_fig("Figure_S6_central_metabolism_genes.pdf"), height = 15, width = 12)

compound_class <- data.frame(tax_table(metabolites_phylo))

# Three panels: peptides, free amino acids, carbohydrates. ClassyFire files
# peptides and free amino acids together under the class "Carboxylic acids and
# derivatives", so grouping by Class would merge the first two panels. The split
# is taken at Parent.Level.1, matching the class-level tests in section 2.9.5,
# and the panels are ordered so the peptide result and the free amino acid null
# sit next to each other.
if (!all(c("Parent.Level.1", "Subclass") %in% names(compound_class)))
  stop("The metabolite tax table needs Parent.Level.1 and Subclass to build the ",
       "peptide, amino acid and carbohydrate panels of Supplementary Figure S7.")
compound_class$Metab_set <- dplyr::case_when(
  compound_class$Parent.Level.1 %in% "Peptides"                            ~ "Peptides",
  compound_class$Parent.Level.1 %in% "Amino acids and derivatives"         ~ "Amino acids",
  compound_class$Subclass %in% "Carbohydrates and carbohydrate conjugates" ~ "Carbohydrates",
  TRUE                                                                     ~ NA_character_)
cat("\nSupplementary Figure S7 features: peptides",
    sum(compound_class$Metab_set %in% "Peptides"),
    "| amino acids", sum(compound_class$Metab_set %in% "Amino acids"),
    "| carbohydrates", sum(compound_class$Metab_set %in% "Carbohydrates"), "\n")

amino_carb_summary <- {
  met_norm <- sample_matrix(metabolites_phylo)
  met_norm <- met_norm / rowMeans(met_norm, na.rm = TRUE)
  met_norm$SampleID <- rownames(met_norm)
  met_norm
} |>
  left_join(data.frame(sample_data(metabolites_phylo)) |>
              dplyr::select(SampleID = Sample, AvgTemp10cm, Warming_Treatment.ord,
                            Year, WT_drawdown_cm),
            by = "SampleID") |>
  pivot_longer(cols = -c(SampleID, AvgTemp10cm, Warming_Treatment.ord, Year, WT_drawdown_cm),
               names_to = "Feature", values_to = "NormalizedIntensity") |>
  inner_join(filter(compound_class, !is.na(Metab_set)), by = "Feature") |>
  group_by(SampleID, AvgTemp10cm, Warming_Treatment.ord, Year, WT_drawdown_cm, Metab_set) |>
  dplyr::summarise(MeanIntensity = mean(NormalizedIntensity, na.rm = TRUE), .groups = "drop") |>
  mutate(Metab_set = factor(Metab_set,
                            levels = c("Peptides", "Amino acids", "Carbohydrates")),
         Year      = as.factor(Year))

if (nrow(amino_carb_summary) == 0)
  stop("No peptide, amino acid or carbohydrate feature matched the tax table. ",
       "Check that make.names() has not altered the feature IDs.")
if (dplyr::n_distinct(amino_carb_summary$Metab_set) != 3)
  stop("Supplementary Figure S7 expects three panels; found: ",
       paste(levels(droplevels(amino_carb_summary$Metab_set)), collapse = ", "))

save_fig(
  ggplot(amino_carb_summary, aes(x = AvgTemp10cm, y = MeanIntensity, shape = Year)) +
    scale_shape_manual(values = c(21, 22, 24)) +
    geom_point(aes(fill = Warming_Treatment.ord), size = 8, colour = "black") +
    geom_smooth(aes(linetype = Year), method = "lm", se = FALSE, linewidth = 1, color = "black") +
    facet_wrap(~Metab_set) +
    scale_fill_manual(values = allGroupsColors) +
    scale_linetype_manual(values = c("2020" = "solid","2021" = "dashed","2022" = "dotted")) +
    guides(fill = guide_legend("Temperature\ntreatment", override.aes = list(shape = 21))) +
    labs(x = "Temperature at 10 cm (May-Aug, °C)", y = "Mean Normalized Intensity") +
    theme_bw() +
    theme(legend.position = "right", axis.text = element_text(size = 12),
          axis.title = element_text(size = 16), strip.text = element_text(size = 16),
          strip.background = element_blank()),
  p_fig("Figure_S7_peptide_amino_acid_carbohydrate.pdf"), width = 14, height = 5)

aa_carb_models <- purrr::imap_dfr(
  list("Peptides (set mean intensity)"                             = "Peptides",
       "Amino acids, free, peptides excluded (set mean intensity)" = "Amino acids",
       "Carbohydrates (set mean intensity)"                        = "Carbohydrates"),
  \(cls, nm) {
    m <- stats::step(lm(MeanIntensity ~ AvgTemp10cm * WT_drawdown_cm,
                        data = filter(amino_carb_summary, Metab_set == cls)), trace = 0)
    s <- summary(m)
    data.frame(response = nm, formula = paste(deparse(formula(m)), collapse = " "),
               term = rownames(s$coefficients), s$coefficients,
               adj_R2 = s$adj.r.squared, n = nobs(m), row.names = NULL)
  })
# Printed as a cross-check only; the reported values come from the class-mean
# mixed models above.
print(aa_carb_models)
```

## 7. KEGG PATHWAY ENRICHMENT — Figure 5 (EXPLORATORY)

Exploratory. The gene-level tests behind the query set are close to, or below,
the 5% null expectation, and the chunk prints the counts it is based on.
`universe` is the set of KO IDs detected in these genomes. Correction is `holm`,
with BH reported alongside.

```{r clusterprofiler-fig5, fig.width=10, fig.height=8}

Select_taxa_net <- read.csv(net_path("Linking_Metabolites_INFO.csv")) |>
  pull(genome) |> unique() |> setdiff(c("SPRUCE 190", "SPRUCE 689"))
Thermo_ox_net.genes.ps <- subset_genes(Select_taxa_net, camper_only = FALSE)
taxdf_net <- data.frame(tax_table(Thermo_ox_net.genes.ps))

if (REFIT_MAASLIN && REFIT_LARGE) {
  fit_maaslin(Thermo_ox_net.genes.ps, "RE_networked_ALL",
              fixed = "AvgTemp10cm", random = "Plot")
}
MaAslinLM_net <- readRDS(maaslin_path("RE_networked_ALL.rds"))
df_Coef_net   <- MaAslinLM_net$results |> dplyr::rename(Gene = 1) |>
  add_effect_size() |> left_join(taxdf_net, by = "Gene")
stopifnot(TEMP_GENE %in% df_Coef_net$metadata)

n_tested <- nrow(filter(df_Coef_net, metadata == TEMP_GENE))
n_p05    <- sum(df_Coef_net$metadata == TEMP_GENE & df_Coef_net$pval < 0.05, na.rm = TRUE)
cat(sprintf("Gene-level: %d tested, %d at P<0.05 (%.1f%%, null expectation 5%%), min q = %.3f\n",
            n_tested, n_p05, 100 * n_p05 / n_tested,
            min(df_Coef_net$qval[df_Coef_net$metadata == TEMP_GENE], na.rm = TRUE)))

df_ko <- df_Coef_net |>
  filter(metadata == TEMP_GENE, !is.na(ko_id), !ko_id %in% c("", "NA")) |>
  mutate(diffexpressed = case_when(coef > 0 & pval < 0.05 ~ "Warm",
                                   coef < 0 & pval < 0.05 ~ "Ambient",
                                   TRUE ~ "NO"))

KO_UNIVERSE <- unique(df_ko$ko_id)
cat("Background universe:", length(KO_UNIVERSE), "unique KO IDs\n")

pathways   <- keggList("pathway", "ko")
kegg_names <- data.frame(
  ID          = sub("path:", "", names(pathways)),
  Description = sub(" - Reference pathway", "", pathways),
  stringsAsFactors = FALSE)
TERM2GENE <- clusterProfiler::download_KEGG("ko")$KEGGPATHID2EXTID |>
  dplyr::select(term = from, gene = to) |>
  left_join(kegg_names, by = c("term" = "ID")) |>
  mutate(term = Description) |>
  dplyr::select(term, gene) |>
  filter(!is.na(term))

ora <- clusterProfiler::enricher(
  gene          = df_ko$ko_id[df_ko$diffexpressed == "Warm"],
  universe      = KO_UNIVERSE,
  TERM2GENE     = TERM2GENE,
  pAdjustMethod = "holm",
  pvalueCutoff  = 1, qvalueCutoff = 1)

if (is.null(ora) || nrow(ora@result) == 0)
  stop("enricher() returned nothing. Check that ko_id values match TERM2GENE.")
res_df <- ora@result |>
  mutate(p_adjust_BH = p.adjust(pvalue, "BH"), p_adjust_holm = p.adjust(pvalue, "holm"))
print(head(res_df[order(res_df$pvalue), c("ID","Description","Count","pvalue","p_adjust_BH","p_adjust_holm")], 20), row.names = FALSE)

cat("Pathways at raw P<0.05 and >=10 genes:",
    sum(res_df$pvalue < 0.05 & res_df$Count >= 10), "\n")
cat("Surviving Holm P<0.05:", sum(res_df$p_adjust_holm < 0.05), "\n")
cat("Surviving BH q<0.05:  ", sum(res_df$p_adjust_BH   < 0.05), "\n")

res_df_filt <- filter(res_df, pvalue < 0.05, Count >= 10)

if (nrow(res_df_filt) == 0) {
  warning("No KEGG pathway passes the Figure 5 display filter.")
} else {
  FIG5 <- ggplot(arrange(res_df_filt, desc(Count)),
                 aes(x = fct_reorder(Description, Count), y = Count, fill = pvalue)) +
    geom_col() + coord_flip() +
    scale_fill_gradient(name = "P-value", high = "#9ecae1", low = "#08306b") +
    labs(x = "KEGG Pathway", y = "Number of genes with increased transcription",
         caption = paste0("Exploratory. Query set defined at nominal P < 0.05; ",
                          n_p05, "/", n_tested, " genes (",
                          sprintf("%.1f%%", 100 * n_p05 / n_tested),
                          ") reach that threshold, against a 5% null expectation.")) +
    theme_bw() +
    theme(axis.text = element_text(size = 12), axis.title = element_text(size = 16),
          plot.caption = element_text(size = 9, hjust = 0))
  save_fig(FIG5, p_fig("Figure_5_KEGG_enrichment.pdf"), height = 8, width = 10)
}

# --- Supplementary Data 14 ---------------------------------------------------
Suppl_Data_14 <- arrange(df_Coef_net, metadata, pval)
cat("Supplementary Data 14: ", nrow(Suppl_Data_14), " rows, ",
    sum(Suppl_Data_14$pval < 0.05, na.rm = TRUE), " at P<0.05\n", sep = "")
if (sum(Suppl_Data_14$pval < 0.05, na.rm = TRUE) == 0)
  stop("Supplementary Data 14 contains no gene at P < 0.05.")
write_out(Suppl_Data_14, p_sd("Suppl_Data_14_networked_genes.csv"))
```

## 8. CAMPER GENES, FULL COMMUNITY

FDR census across the whole community. Not submitted as a numbered sheet.

```{r maaslin-all-camper}

if (REFIT_MAASLIN && REFIT_LARGE) {
  fit_maaslin(subset_taxa(Genes_withMAG_NORMALIZED, CAMPER_id != "" & !is.na(CAMPER_id)),
              "RE_ALL_CAMPERgenes",
              fixed = c("AvgTemp10cm", "WT_drawdown_cm"), random = "Plot")
}
df_Coef_ALLCAMPER <- readRDS(maaslin_path("RE_ALL_CAMPERgenes.rds"))$results |>
  dplyr::rename(Gene = 1) |> add_effect_size()
print(fdr_summary(list(results = df_Coef_ALLCAMPER), "All CAMPER genes ~ Temp + WT + (1|Plot)"))
```

## 9. METHANOGEN ACTIVITY — Figure 6

```{r methanogen, fig.width=12, fig.height=6}

# --- Supplementary Figure S8 --------------------------------------------------

matching_mcrA <- rownames(tax_table(Genes_withMAG_NORMALIZED))[
  tax_table(Genes_withMAG_NORMALIZED)[, "ko_id"] %in% "K00399"]
methanogen_gene_ps <- prune_taxa(
  intersect(matching_mcrA, rownames(tax_table(subset_genes(METHANOGEN_IDS, FALSE)))),
  Genes_withMAG_NORMALIZED)
active_mcrA <- unique(data.frame(tax_table(methanogen_gene_ps))$SPRUCE_ID)
cat("Methanogen MAGs expressing mcrA:", paste(active_mcrA, collapse = ", "), "\n")

# --- Share of mcrA expression held by the four oxygen-sensitive methanogens ----
# Denominator: every gene annotated K00399 in any MAG. Numerator: the K00399
# genes in METHANOGEN_IDS. Abundance is ribosome-normalised gene activity.

mcrA_all_ps <- prune_taxa(matching_mcrA, Genes_withMAG_NORMALIZED)

mcrA_share <- psmelt(mcrA_all_ps) |>
  dplyr::mutate(is_methanogen = SPRUCE_ID %in% METHANOGEN_IDS) |>
  dplyr::group_by(SPRUCE_ID, is_methanogen) |>
  dplyr::summarise(total_activity = sum(Abundance, na.rm = TRUE),
                   n_genes        = dplyr::n_distinct(OTU),
                   n_samples_nonzero = sum(Abundance > 0, na.rm = TRUE),
                   .groups = "drop") |>
  dplyr::arrange(dplyr::desc(total_activity)) |>
  dplyr::mutate(pct_of_total = 100 * total_activity / sum(total_activity))

MCRA_PCT_METHANOGENS <- sum(mcrA_share$pct_of_total[mcrA_share$is_methanogen])

cat(sprintf(
  "mcrA (K00399): %d genes across %d MAGs. The %d methanogen MAGs hold %.1f%% of total normalised mcrA activity.\n",
  length(matching_mcrA), nrow(mcrA_share), sum(mcrA_share$is_methanogen),
  MCRA_PCT_METHANOGENS))

if (nrow(mcrA_share) > sum(mcrA_share$is_methanogen))
  cat("  Non-methanogen MAGs carrying K00399: ",
      paste(mcrA_share$SPRUCE_ID[!mcrA_share$is_methanogen], collapse = ", "), "\n", sep = "")

print(as.data.frame(mcrA_share), digits = 3, row.names = FALSE)


df_Meth_PW <- prune_taxa(taxa_names(droughtMAG.activity.ps) %in% active_mcrA,
                         droughtMAG.activity.ps) |>
  psmelt() |>
  mutate(WT_drawdown_cm = WT_drawdown * 100, Year = as.factor(Year), Plot = as.factor(Plot)) |>
  group_by(Warming, Year, Plot, AvgTemp10cm, Temp_avg_top, WT_drawdown_cm,
           Warming_Treatment.ord) |>
  dplyr::summarize(sum_Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop") |>
  left_join(dplyr::select(PWgas_sub, Year, Plot, CH4mM), by = c("Year", "Plot")) |>
  filter(!is.na(CH4mM))

model_meth_CH4 <- glmmTMB(CH4mM ~ sum_Abundance + (1 | Plot), family = gaussian(),
                          data = df_Meth_PW)
meth_fix <- drop_singular_re(model_meth_CH4, df_Meth_PW, "CH4 ~ methanogen activity")
register_model(tidy_glmmtmb(meth_fix$model, "Porewater CH4 ~ total methanogen activity"))
print(tidy_glmmtmb(meth_fix$model, "Porewater CH4 ~ total methanogen activity"))

save_fig(
  ggplot(df_Meth_PW, aes(x = sum_Abundance, y = CH4mM)) +
    geom_point(aes(fill = Warming_Treatment.ord), colour = "black", size = 8, pch = 21) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", linewidth = 1) +
    scale_fill_manual(values = allGroupsColors) +
    guides(fill = guide_legend("Temperature\ntreatment", override.aes = list(shape = 21))) +
    xlab("Total relative activity of methanogens") + ylab("Porewater CH4 (mM)") +
    theme_bw() +
    theme(legend.position = "right", axis.text = element_text(size = 12),
          axis.title = element_text(size = 16)),
  p_fig("Figure_S8_methanogen_activity.pdf"), width = 8, height = 7)

# --- Annual CH4 emission by year, Supplementary Data 1 and Figure S8C ---------
# Modelled annual emission per enclosure from Cory et al. (2025). Restricted to
# the ten study enclosures and the three sampled years, and fitted on the log
# scale because emission spans more than an order of magnitude between enclosures.
#
# Temperature enters as a fixed effect. Warming is assigned per enclosure, so a
# year-only model with (1|Plot) hides the whole warming gradient inside the
# random intercept: enclosure variance falls from 0.46 to 0.05 once temperature
# is fitted, meaning most of what separates enclosures is the treatment. The
# year by temperature interaction is kept when the LRT supports it, as elsewhere.
#
# Year is referenced to 2021, the drought year, because the contrasts the text
# discusses are the drop into 2021 and the recovery out of it. Referenced to 2020
# the 2022 coefficient tests 2022 against 2020, which is not that comparison.

CH4_flux_annual <- readr::read_csv(p_raw_req(CH4_FLUX_CSV), show_col_types = FALSE,
                                   name_repair = "minimal") |>
  rlang::set_names(c("Year", "Plot", "CH4_flux")) |>
  dplyr::mutate(Year     = suppressWarnings(as.integer(Year)),
                CH4_flux = suppressWarnings(as.numeric(CH4_flux)),
                Plot     = suppressWarnings(as.integer(sub("^P0*", "", Plot)))) |>
  dplyr::filter(!is.na(Year), !is.na(CH4_flux), !is.na(Plot),
                Year %in% c(2020, 2021, 2022), Plot %in% STUDY_PLOTS) |>
  dplyr::left_join(
    # Plot and Year are factors in metadata_df, so they are taken back through
    # as.character(); as.integer() on a factor returns the level index.
    dplyr::select(metadata_df, Plot, Year, Temp_avg_top, WT_drawdown_cm) |>
      dplyr::mutate(Plot = as.integer(as.character(Plot)),
                    Year = as.integer(as.character(Year))) |>
      dplyr::distinct(),
    by = c("Plot", "Year")) |>
  dplyr::mutate(Year          = factor(Year, levels = c(2021, 2020, 2022)),
                Plot          = as.factor(Plot),
                log_CH4_flux  = log(CH4_flux),
                # Centred so that with the interaction in the model the Year
                # coefficients are year contrasts at the mean temperature of the
                # experiment rather than extrapolations to 0 degrees.
                Temp_c        = Temp_avg_top - mean(Temp_avg_top))

if (anyNA(CH4_flux_annual$Temp_avg_top))
  stop("Temperature is missing for ", sum(is.na(CH4_flux_annual$Temp_avg_top)),
       " enclosure-year(s) in the flux table.")
if (nrow(CH4_flux_annual) != dplyr::n_distinct(CH4_flux_annual$Plot,
                                               CH4_flux_annual$Year))
  stop("The metadata join duplicated enclosure-years in the flux table.")

cat(sprintf("Annual CH4 flux: %d enclosure-years across %d enclosures and %d years\n",
            nrow(CH4_flux_annual), dplyr::n_distinct(CH4_flux_annual$Plot),
            dplyr::n_distinct(CH4_flux_annual$Year)))
print(CH4_flux_annual |>
        dplyr::group_by(Year) |>
        dplyr::summarise(n = dplyr::n(),
                         median_flux = median(CH4_flux),
                         .groups = "drop") |>
        as.data.frame(), row.names = FALSE)

FLUX_LABEL <- "Annual CH4 flux ~ year x temperature (log scale)"
flux_lrt <- lrt_interaction(log_CH4_flux ~ Year * Temp_c + (1 | Plot),
                            log_CH4_flux ~ Year + Temp_c + (1 | Plot),
                            CH4_flux_annual, label = "Annual CH4 flux")
cat("Mean temperature used for centring: ",
    round(mean(CH4_flux_annual$Temp_avg_top), 2), " C\n", sep = "")
flux_fix <- drop_singular_re(flux_lrt$model, CH4_flux_annual, FLUX_LABEL)
register_model(tidy_glmmtmb(flux_fix$model, FLUX_LABEL))
print(tidy_glmmtmb(flux_fix$model, FLUX_LABEL))

# Year-only fit, kept as a reference point for how much of the between-enclosure
# variation the warming treatment accounts for.
flux_yearonly <- glmmTMB(log_CH4_flux ~ Year + (1 | Plot), family = gaussian(),
                         data = CH4_flux_annual)
cat("Enclosure SD, year only: ",
    signif(attr(VarCorr(flux_yearonly)$cond$Plot, "stddev")[1], 3),
    " | with temperature: ",
    signif(tryCatch(attr(VarCorr(flux_fix$model)$cond$Plot, "stddev")[1],
                    error = function(e) NA_real_), 3), "\n", sep = "")

# Back-transform each year contrast to a percentage change against 2021, the
# drought year, evaluated at the mean temperature of the experiment.
flux_coefs <- as.data.frame(summary(flux_fix$model)$coefficients$cond)
colnames(flux_coefs) <- c("Estimate", "Std_Error", "z_value", "p_value")
flux_coefs$term <- rownames(flux_coefs)
for (yr in grep("^Year", flux_coefs$term, value = TRUE))
  cat(sprintf("  %s: %+.1f%% relative to 2021 (P = %.3g)\n", yr,
              100 * (exp(flux_coefs$Estimate[flux_coefs$term == yr]) - 1),
              flux_coefs$p_value[flux_coefs$term == yr]))

# --- Unadjusted year-to-year contrasts, paired within enclosure ---------------
# The model above compares years at a common temperature. Ambient temperature
# also differed between years, so the two questions have different answers and
# the manuscript states both. These are the raw contrasts: each enclosure is its
# own control, which is the right pairing for a balanced 10 x 3 design and does
# not pool one residual variance across the drought year and the wet years.

flux_wide <- CH4_flux_annual |>
  dplyr::select(Plot, Year, log_CH4_flux) |>
  tidyr::pivot_wider(names_from = Year, values_from = log_CH4_flux,
                     names_prefix = "y")
temp_wide <- CH4_flux_annual |>
  dplyr::select(Plot, Year, Temp_avg_top) |>
  tidyr::pivot_wider(names_from = Year, values_from = Temp_avg_top,
                     names_prefix = "y")

flux_year_contrasts <- purrr::map_dfr(
  list(c("y2020", "y2021"), c("y2022", "y2021"), c("y2022", "y2020")),
  function(pr) {
    a <- flux_wide[[pr[1]]]; b <- flux_wide[[pr[2]]]
    tt <- stats::t.test(a, b, paired = TRUE)
    data.frame(
      contrast          = paste(sub("^y", "", pr[1]), "vs", sub("^y", "", pr[2])),
      n_enclosures      = length(a),
      mean_log_diff     = unname(tt$estimate),
      percent_change    = 100 * (exp(unname(tt$estimate)) - 1),
      t                 = unname(tt$statistic),
      p_value           = tt$p.value,
      n_higher          = sum(a > b),
      temp_diff_C       = mean(temp_wide[[pr[1]]]) - mean(temp_wide[[pr[2]]]))
  })
cat("\n=== Annual CH4 flux, unadjusted paired contrasts ===\n")
print(flux_year_contrasts, row.names = FALSE, digits = 3)

CH4_flux_annual$Year_display <- factor(as.character(CH4_flux_annual$Year),
                                       levels = c("2020", "2021", "2022"))
save_fig(
  ggplot(CH4_flux_annual, aes(x = Year_display, y = CH4_flux)) +
    geom_boxplot(outlier.shape = NA, fill = "grey92") +
    geom_line(aes(group = Plot), colour = "grey60", linewidth = 0.4) +
    geom_point(size = 5, pch = 21, fill = "grey40", colour = "black") +
    scale_y_continuous(trans = "log10") +
    labs(x = NULL, y = expression(paste("Annual C", H[4], " flux (", mu,
                                        "mol ", m^-2, " ", y^-1, ")"))) +
    theme_bw() +
    theme(axis.text = element_text(size = 12), axis.title = element_text(size = 16)),
  p_fig("Figure_S8C_methane_flux.pdf"), width = 5, height = 5)

# --- Figure 6A: mcrA per MAG --------------------------------------------------
# Full models, no term elimination. Zero-inflation term for MAGs with >20% zeros.

df_sensitive.mcrA <- psmelt(methanogen_gene_ps) |>
  mutate(WT_drawdown_cm = WT_drawdown * 100, Year = as.factor(Year), Plot = as.factor(Plot))

df_sensitive.mcrA |>
  group_by(SPRUCE_ID) |>
  summarise(n = n(), n_zero = sum(Abundance == 0),
            pct_zero = round(mean(Abundance == 0) * 100, 1), .groups = "drop") |>
  print()

mcrA_table <- purrr::map_dfr(unique(df_sensitive.mcrA$SPRUCE_ID), function(sid) {
  d  <- filter(df_sensitive.mcrA, SPRUCE_ID == sid)
  zi <- mean(d$Abundance == 0) >= 0.2
  m  <- glmmTMB(Abundance ~ Temp_avg_top * WT_drawdown_cm + (1 | Plot),
                ziformula = if (zi) ~ WT_drawdown_cm else ~0,
                family = gaussian(link = "identity"), data = d)
  out <- tidy_glmmtmb(m, sid) |> mutate(component = "conditional", model_type = if (zi) "ZIGA" else "Gaussian")
  if (zi) {
    z <- as.data.frame(summary(m)$coefficients$zi)
    colnames(z) <- c("Estimate","Std_Error","z_value","p_value")
    out <- bind_rows(out, data.frame(
      response = sid, formula = paste(deparse(formula(m)), collapse = " "), n = nobs(m),
      z, term = paste0("ZI:", rownames(z)), sig = "", R2_marginal = NA, R2_conditional = NA,
      singular_fit = any(is.nan(z$Std_Error)), component = "zero-inflation",
      model_type = "ZIGA", row.names = NULL))
  }
  out
})

# Flag coefficients whose standard error swamps the estimate.
mcrA_table <- mcrA_table |>
  mutate(unstable = is.na(Std_Error) |
                    (!is.na(Estimate) & Std_Error > 100 * abs(Estimate)))
if (any(mcrA_table$unstable))
  warning("Unstable mcrA coefficients (SE >> estimate): ",
          paste(unique(mcrA_table$response[mcrA_table$unstable]), collapse = ", "))
print(mcrA_table)
write_out(mcrA_table, p_sd("Suppl_Data_16_mcrA_models.csv"))

df_sensitive.mcrA$taxonomy <- paste0(df_sensitive.mcrA$SPRUCE_ID, ": ", df_sensitive.mcrA$Phylum)
save_fig(
  ggplot(df_sensitive.mcrA, aes(x = WT_drawdown_cm, y = Abundance)) +
    facet_wrap(~taxonomy, scales = "free_y", ncol = 2) +
    geom_point(aes(fill = Warming_Treatment.ord, shape = Year), colour = "black", size = 5) +
    scale_shape_manual(values = c(21, 22, 24)) + scale_fill_manual(values = allGroupsColors) +
    xlab("Water table drawdown (cm)") +
    ylab("Relative activity mcrA\n(TAD80/ribosomal expression)") +
    theme_bw() +
    theme(legend.position = "none", axis.text = element_text(size = 12),
          axis.title = element_text(size = 16), strip.text = element_text(size = 12)),
  p_fig("Figure_6A_mcrA_water_table.pdf"), height = 6, width = 7)

# --- Figure 6B: methanogen phenol-active genes decreasing ---------------------
# Four genomes. Effective n is 4, not the number of genes.

ps_meth <- subset_genes(METHANOGEN_IDS)
if (REFIT_MAASLIN) {
  fit_maaslin(ps_meth, "RE_Methano_campergenes",
              fixed = c("AvgTemp10cm", "WT_drawdown_cm"), random = "Plot")
}
df_Coef_Meth <- read_gene_maaslin(maaslin_path("RE_Methano_campergenes.rds"), ps_meth)
stopifnot(all(c(TEMP_GENE, WT_GENE) %in% df_Coef_Meth$metadata))
cat("Methanogen CAMPER genes: ", nrow(df_Coef_Meth), " tests, ",
    sum(df_Coef_Meth$coef < 0), " negative coefficients, across ",
    n_distinct(df_Coef_Meth$SPRUCE_ID), " genomes\n", sep = "")

genes_df_meth <- add_taxonomy(count_gene_responses(df_Coef_Meth, "down", METHANOGEN_IDS),
                              data.frame(tax_table(ps_meth)))
pink <- scico(max(5, n_distinct(genes_df_meth$taxonomy)), palette = "acton")

fig6b_A <- camper_scatter(filter(genes_df_meth, metadata == TEMP_GENE), "Temperature",
                          pink, "decreasing")
fig6b_B <- camper_scatter(filter(genes_df_meth, metadata == WT_GENE),   "Water table drawdown",
                          pink, "decreasing", "right")
save_legend(fig6b_B, legend_path(p_fig("Figure_6B_methanogen_phenolic_genes.pdf")),
            width = 3.5, height = 5)
save_fig(ggarrange(fig6b_A, fig6b_B + theme(legend.position = "none"),
    ncol = 2, labels = c("A","B"), font.label = list(size = 18)),
  p_fig("Figure_6B_methanogen_phenolic_genes.pdf"), width = 10, height = 4)

write_out(filter(df_Coef_Meth, pval < 0.05) |> arrange(metadata, pval),
          p_sd("Suppl_Data_17_methanogen_CAMPER_genes.csv"))
```

## 10. SUPPLEMENTARY DATA 1 AND 2 — MODEL COEFFICIENT TABLES

Every `glmmTMB` fit registers itself in `MODEL_TABLE` with the formula it fitted.

```{r stat-summary-tables}

all_models <- bind_rows(MODEL_TABLE)

env_responses <- c("Porewater CH4 (mM)", "Porewater CO2 (mM)", "Porewater CO2:CH4 ratio",
                   "Gallic-acid-equivalent reducing capacity",
                   "Porewater CH4 ~ total methanogen activity",
                   "Annual CH4 flux ~ year x temperature (log scale)")
write_out(filter(all_models, response %in% env_responses |
                             grepl("^Sample NOSC", response)),
          p_sd("Suppl_Data_01_environmental_models.csv"))
write_out(filter(all_models, grepl("glucosidase|Cellobiohydrolase|acetylglucosaminidase|phosphatase|Phenol oxidase",
                                   response)),
          p_sd("Suppl_Data_02_enzyme_models.csv"))

if (any(all_models$singular_fit, na.rm = TRUE))
  warning("Singular fits present: ",
          paste(unique(all_models$response[all_models$singular_fit]), collapse = ", "))
```

## 11. README AND OUTPUT MANIFEST

```{r readme-and-manifest}

expected <- c(
  "Suppl_Data_01_environmental_models.csv",    "Suppl_Data_02_enzyme_models.csv",
  "Suppl_Data_03_metabolites_WT.csv",          "Suppl_Data_04_metabolites_temp.csv",
  "Suppl_Data_05_classlevel_competitive.csv",  "Suppl_Data_06_MAG_taxonomy_quality.csv",
  "Suppl_Data_07_dbRDA_VIF.csv",               "Suppl_Data_08_MAG_responses.csv",
  "Suppl_Data_09_cytochrome_c_temperature.csv","Suppl_Data_10_linked_metabolites.csv",
  "Suppl_Data_11_hydrogenases_HydDB.csv",      "Suppl_Data_12_WSOX_CAMPER_genes.csv",
  "Suppl_Data_13_WSAN_CAMPER_genes.csv",       "Suppl_Data_14_networked_genes.csv",
  "Suppl_Data_15_methanogen_pathways.csv",     "Suppl_Data_16_mcrA_models.csv",
  "Suppl_Data_17_methanogen_CAMPER_genes.csv", "Suppl_Data_18_annotation_citations.csv",
  "Suppl_Data_03_04_column_definitions.csv")

manifest <- data.frame(
  file    = expected,
  present = file.exists(p_sd(expected)),
  rows    = vapply(p_sd(expected),
                   \(f) if (file.exists(f)) nrow(readr::read_csv(f, show_col_types = FALSE))
                        else NA_integer_, integer(1)))
print(manifest, row.names = FALSE)

if (any(!manifest$present))
  warning("Missing supplementary data: ",
          paste(manifest$file[!manifest$present], collapse = ", "))

writeLines(paste0(
"# SPRUCE Drought Manuscript Analysis Code

Manuscript: Combined effects of warming and drought accelerate microbially
mediated organic matter decomposition and suppress methanogenesis in peatlands
Authors: Katherine Duchesneau et al.
Journal: Nature Communications, NCOMMS-26-018889-T
Code DOI (Zenodo): 10.5281/zenodo.19865237

## How to run

1. Open `Drought_R_analyses.Rproj`.
2. Set `PROJ` at the top of `Drought_Manuscript_Analysis.Rmd` to your local
   path. It is the only absolute path in the document.
3. Knit `Drought_Manuscript_Analysis.Rmd`. It runs top to bottom and reads
   nothing from outside `2Revisions_out/00_inputs/`.

## Run switches, all at the top of the setup chunk

REFIT_MAASLIN   refit every Maaslin2 model from the phyloseq objects
REFIT_LARGE     also refit the two large gene models (hours to overnight)
MAASLIN_CACHE   skip a refit when the stored fit matches the same inputs
RUN_NETWORK     TRUE fits SpiecEasi, FALSE reloads 08_network/se_FULL.RDS
RESAMPLE_N      replicate count behind the Figure 3 edge-recovery input
RUN_RESAMPLE    TRUE recomputes edge recovery when the plotted edge set no
                longer matches 00_inputs/Check_Fig3_edge_stability.csv;
                FALSE stops the knit
WIPE_OUTPUTS    delete 01_ to 08_ before the run; 00_inputs is never touched

## Files

Drought_Manuscript_Analysis.Rmd   all analyses, figures and tables
Drought_R_analyses.Rproj          RStudio project

## Folders

2Revisions_out/   the whole analysis, inputs and outputs
  00_inputs/      every file the Rmd reads
  01_metadata/    merged metadata, NOSC, porewater gas, enzyme, CAMPER counts
  02_phyloseq/    phyloseq objects this script creates
  03_maaslin/     every Maaslin2 fit
  04_suppl_data/  Supplementary Data 1-18
  06_figures/     all main and supplementary figure PDFs, named as they are
                  cited: Figure_1 to Figure_6, Figure_S1 to Figure_S9
  08_network/     SpiecEasi fit, association matrix, linking tables, Figure 3

## Supplementary Data

01 environmental, porewater and annual CH4 flux mixed models
02 enzyme activity mixed models
03 metabolite feature responses to water table drawdown
04 metabolite feature responses to temperature
05 class-level competitive tests on the metabolite coefficients
06 MAG taxonomy and genome quality
07 VIF for all dbRDA models
08 clustered MAG responses to temperature, water table and NOSC
09 cytochrome c gene responses to temperature in WS-OX genomes
10 metabolites linking warming-responsive MAGs to methanogens (Figure 3)
11 hydrogenase gene responses in WS-AN genomes, HydDB annotated
12 phenol-active gene responses in WS-OX genomes (Figure 4A)
13 phenol-active gene responses in WS-AN genomes (Figure 4B)
14 gene responses to temperature in methanogen-connected genomes (Figure 5)
15 methanogen KEGG module presence (hand-curated, copied from 00_inputs)
16 per-MAG mcrA models (Figure 6A)
17 phenol-active gene responses in methanogens (Figure 6B)
18 custom KEGG module definitions (hand-curated, copied from 00_inputs)

## Figures

Figure_1_dbRDA.pdf                        assembled four-panel Figure 1
Figure_1A_MAG_activity.pdf                MAG activity dbRDA
Figure_1B_metabolites.pdf                 metabolite composition dbRDA
Figure_1C_16S.pdf                         16S community dbRDA
Figure_1D_ITS.pdf                         ITS community dbRDA
Figure_2A_genome_response_tree.pdf        MAG responses on the phylogeny
Figure_2B_CO2_genome_scatter.pdf          porewater CO2 against genome group
Figure_2C_phylogenetic_tree.pdf           294 MAGs, phenol-active gene counts
Figure_3_network.pdf                      SpiecEasi metabolite-informed network
Figure_4A_WSOX_phenolic_genes.pdf         phenol-active gene responses, WS-OX
Figure_4B_WSAN_phenolic_genes.pdf         phenol-active gene responses, WS-AN
Figure_5_KEGG_enrichment.pdf              KEGG pathway enrichment
Figure_6A_mcrA_water_table.pdf            mcrA transcripts against drawdown
Figure_6B_methanogen_phenolic_genes.pdf   phenol-active genes in methanogens
Figure_S1_environmental_drivers.pdf       water table, moisture, temperature
Figure_S2_porewater_gas.pdf               porewater CO2, CH4 and their ratio
Figure_S3_enzyme_activity.pdf             phenol oxidase and hydrolase assays
Figure_S4_MAG_abundance_dbRDA.pdf         MAG relative abundance dbRDA
Figure_S5_cytochrome_c.pdf                cytochrome c genes in WS-OX genomes
Figure_S6_central_metabolism_genes.pdf    K00382, K01915 and K00873
Figure_S7_peptide_amino_acid_carbohydrate.pdf  class mean intensity by temperature
Figure_S8_methanogen_activity.pdf         porewater CH4 against methanogen activity
Figure_S8C_methane_flux.pdf               annual CH4 emission by year
Figure_S9_variation_partitioning.pdf      varpart Venn diagrams

"), file.path(PROJ, "README.md"))
cat("README.md written.\n")

cat("\n=== Files written this run ===\n")
print(data.frame(file = FILE_LOG, exists = file.exists(FILE_LOG)), row.names = FALSE)
sessionInfo()
```
